---
title: "Intermittent-flow respirometry: Alternative approaches"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Intermittent-flow respirometry: Alternative approaches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 4))
```


The following are alternative methods of extracting rates from multiple replicates in intermittent-flow respirometry using `calc_rate` or `auto_rate`. They have been mostly superseded by `calc_rate.int` and `auto_rate.int` but we have left the details here in case they are of use in other cases. The main advantage to these are that they allow rates to be extracted from different data regions within each replicate or allow different methods to be used.

## `calc_rate` {#calc_rate}

`calc_rate.int` allows consistent region selection criteria to be applied to each replicate, and is usually the best way of extracting rates from intermittent-flow data. `calc_rate` however also allows you to extract multiple rates from a dataset in a single command, and this can allow for rates from different regions in each replicate. Obviously, this requires you to know the row locations or timings of replicates. 

`calc_rate` allows data regions to be chosen by `oxygen`, `time`, or `row` ranges. In the case of `row` or `time`, multiple subset regions can be specified, with the `from` and `to` inputs acting as vectors of paired values. Using this we can extract a rate from each replicate with one command. 

### Rate from multiple row regions

We can use the same replicate start and end locations as in the example [here](https://januarharianto.github.io/respR/articles/intermittent_short.html) to extract a rate from each complete replicate. 

```{r tidy = FALSE, results = "hide"}
# inspect data
urchin_int <- inspect(intermittent.rd)

# calc rates
urchin_int_rates <- calc_rate(urchin_int, 
                              from = c(1, 2101, 3901),
                              to = c(1900, 3550, 4831),
                              by = "row")
```

```{r echo = T}
summary(urchin_int_rates)
```

We can see this produces exactly the same result as in the example [here](https://januarharianto.github.io/respR/articles/intermittent_short.html). Note, because `calc_rate.int` has not been used there are no values in the `$rep` column. `calc_rate` does not consider multiple rates as coming from separate replicates. Instead it ranks them in order of inputs, as indicated by the `$rank` column. 

### Rate from multiple time regions

We can also extract by time values, and here we will also apply a different time window within each replicate.

```{r tidy = FALSE, results = "hide"}
urchin_int_rates <- calc_rate(urchin_int, 
                              from = c(200, 2300, 4100), 
                              to = c(1800, 3000, 4400), 
                              by = "time")
```

By default, the first is shown in `print` and `plot`, but the `pos` input can be used to view others. 

```{r results='hide'}
plot(urchin_int_rates, pos = 3)
```

Calling `summary()` will show the coefficients, locations and values of all rates:

```{r}
summary(urchin_int_rates)
```

Using `calc_rate` like this allows you to extract rates from *different* regions within each replicate, or even multiple rates from each. 

## `subset_data` {#subset}

Another option for extracting replicates from a larger dataset is the `subset_data` function. This allows you to easily subset both data frames and `inspect` objects by time, row, or oxygen ranges. You can then pipe (`|>` or `%>%`) the subset directly to other functions such as `calc_rate` or `auto_rate`, or alternatively save replicates as separate objects for further analysis. 

### Separate replicates

Here, we use the `inspect` object we saved earlier containing the whole dataset, to create new `inspect` objects for each replicate. These can be treated like any other `inspect` object, including being passed to `print` and `plot`.

```{r fig.keep="none", message=FALSE, warning=FALSE, results="hide"}
# Create separate replicate data frames
u_rep1 <- subset_data(urchin_int, from = 1, to = 1900, by = "time")
u_rep2 <- subset_data(urchin_int, from = 2100, to = 3500, by = "time")
u_rep3 <- subset_data(urchin_int, from = 3700, to = 4831, by = "time")
```

Now we can calculate a rate from each, showing the results from the third one here. Unlike `calc_rate.int`, using this method you are not restricted to extracting the rate from the exact same region within each replicate. In addition, this approach allows you to use the `by = "oxygen"` method.

```{r eval = F}
u_rate1 <- calc_rate(u_rep1, from = 7.1, to = 6.7, by = "oxygen")
u_rate2 <- calc_rate(u_rep2, from = 7.1, to = 6.8, by = "oxygen")
u_rate3 <- calc_rate(u_rep3, from = 7.0, to = 6.8, by = "oxygen")
```

```{r eval = T, echo = F, results = "hide"}
u_rate3 <- calc_rate(u_rep3, from = 7.0, to = 6.8, by = "oxygen")
```

### Piping

Alternatively, pipe the result, which has the advantage of not filling your local environment with redundant objects and overall makes for a tidier workflow.

```{r fig.keep="none", message=FALSE, warning=FALSE, results="hide"}
u_rate3 <- urchin_int |>
  subset_data(from = 3700, to = 4831, by = "time") |>
  calc_rate(from = 7.0, to = 6.8, by = "oxygen")
```

```{r}
summary(u_rate3)
```


## Iterating using loops

A number of approaches could be used to iterate `respR` functions to analyse these types of experiment; this is merely an example to illustrate how `respR` can be easily iterated over multiple replicates. For most purposes `calc_rate.int` or `auto_rate.int` are better options. 

Here we will show a simple `for` loop to subset each replicate from the `zeb_intermittent.rd` data and run `auto_rate(method = "linear")` on it. We use a 'wait' period of 2 minutes (120s), and a 'measure' period of 7 minutes (420s), leaving 2 minutes of flushing excluded.

For actual analyses it is highly recommended you examine the plot of each replicate, or at very least those used in determining a final rate. Here in the interests of speed we suppress it with `plot = FALSE`. Note also this code will create a `list` object containing an `auto_rate` object for every replicate, which will be quite large (several MB).

### Inspect data

```{r results="hide", message=FALSE, warning=FALSE} 
zeb <- inspect(zeb_intermittent.rd)
```

### Analysis loop

```{r fig.keep="none", message=FALSE, warning=FALSE, results="hide"}
# define wait and measure periods
wait <- 120   # 2 mins wait
measure <- 420  # 7 mins measure

## start rows for each rep using sequence function (from, to, by)
reps <- seq(5840, 74480, 660)
## data starts - apply wait
starts <- reps + wait
## data ends - apply wait and measure period
ends <- reps + wait + measure

## Empty list for saving results
zeb_rmr <- list()

## loop
for(i in 1:105){
  st <- starts[i] # start time
  et <- ends[i] # end time

  ## subset replicate and pipe the result into auto_rate
  zeb_rmr[[i]] <- subset_data(zeb, from = st, to = et, by = "time") |>
    auto_rate(method = "linear", plot = FALSE)
}
```

### View results

We can extract and view the top-ranked rate result from each replicate using the `sapply` function.

```{r}
## extract rates
rmr_rate <- sapply(zeb_rmr, function(z) z$rate[1])
plot(rmr_rate, ylim = rev(range(rmr_rate)))
```

Note, we plot on a reverse axis so higher rates are higher on the plot. We can see that rates are higher in the initial stages of the experiment, then are quite consistent after they stabilise. 

### Background

This saves pre- and post-experiment background rates for use in the next step.

```{r fig.keep="none", message=FALSE, warning=FALSE, results="hide"} 
bg_pre <- zeb |>
  subset_data(from = 1, to = 4999, by = "row") |>
  calc_rate.bg()

bg_post <- zeb |>
  subset_data(from = 75140, to = 79251, by = "row") |>
  calc_rate.bg()
```


### Adjust RMR

Here we use `lapply` to apply the adjustment to each element of the RMR results list of `auto_rate` objects and return a new list of `adjust_rate` objects.

```{r results = "hide", message=FALSE, tidy = FALSE}
zeb_rmr_adj <- lapply(zeb_rmr, function(z) adjust_rate(z,
                                                       by = bg_pre,
                                                       by2 = bg_post,
                                                       method = "linear"))
```

#### Convert RMR

Once again we will use `lapply` to loop through the SMR list of `adjust_rate` objects and convert the rates for each replicate.

```{r message = FALSE}
zeb_rmr_conv <- lapply(zeb_rmr_adj, function(z) convert_rate(z,
                                                             oxy.unit = "mg/L",
                                                             time.unit = "secs",
                                                             output.unit = "mg/h/g",
                                                             volume = 0.12,
                                                             mass = 0.0009))
```

We'll look at two examples. This is the top-ranked result from these two replicates, though the actual object will contain more.

```{r}
summary(zeb_rmr_conv[[10]], pos = 1)
summary(zeb_rmr_conv[[90]], pos = 1)
```

### Final RMR

This is similar to the other operations above, in that we use an `apply` function to extract the top ranked final converted rate from each replicate.

```{r message = FALSE}
zeb_rmr_all <- sapply(zeb_rmr_conv, function(z) z$rate.output[1])
```

Now we can plot them. Again, we reverse the y-axis.

```{r message = FALSE}
plot(zeb_rmr_all, ylim = rev(range(zeb_rmr_all)))
```

It depends on the experiment how we might want to define the final RMR. This is the routine metabolic rate, so we want a rate that represents routine behaviour. Here, rates are very consistent after number 20, apart from one obvious outlier in number 89.

```{r message = FALSE}
zeb_rmr_all[88:90]
```

Therefore, we will not use this one, but take the mean of all others from 20 onwards. This is just one approach of many we could apply. See [here](https://januarharianto.github.io/respR/articles/intermittent_long.html).

```{r message = FALSE}
zeb_rmr_final <- mean(zeb_rmr_all[c(20:88,90:105)])
zeb_rmr_final
```

This is our final RMR: -0.96 mg/h/g.

### Complete analysis

Thius is the same analysis as above but using exclusively the `apply` family of functions.

```{r eval = F, tidy = F}
# Import and inspect raw data ---------------------------------------------
# Importing would normally be the first step, e.g. import_file("path/to/file")
zeb <- inspect(zeb_intermittent.rd)

# Background --------------------------------------------------------------
bg_pre <- subset_data(zeb, from = 0, to = 4999, by = "time") |>
  calc_rate.bg()
bg_post <- subset_data(zeb, from = 75140, to = 79251, by = "time") |>
  calc_rate.bg()

# Replicate structure -----------------------------------------------------
wait <- 120   # 2 mins wait
measure <- 420  # 7 mins measure
reps <- seq(5840, 74480, 660) ## start rows
starts <- reps + wait ## data starts
ends <- reps + wait + measure ## data ends

# Subset each replicate ---------------------------------------------------
zeb_rmr_subsets <- apply(cbind(starts,ends), 1, function(z) subset_data(zeb,
                                                                        from = z[1],
                                                                        to = z[2],
                                                                        by = "time"))

# auto_rate on each replicate ---------------------------------------------
zeb_rmr <- lapply(zeb_rmr_subsets, function(z) auto_rate(z,
                                                         method = "linear",
                                                         plot = FALSE))

# Adjust ------------------------------------------------------------------
zeb_rmr_adj <- lapply(zeb_rmr, function(z) adjust_rate(z,
                                                       by = bg_pre,
                                                       by2 = bg_post,
                                                       method = "linear"))

# Convert -----------------------------------------------------------------
zeb_rmr_conv <- lapply(zeb_rmr_adj, function(z) convert_rate(z,
                                                             oxy.unit = "mg/L",
                                                             time.unit = "secs",
                                                             output.unit = "mg/h/g",
                                                             volume = 0.12,
                                                             mass = 0.0009))

# Extract rates -----------------------------------------------------------
zeb_rmr_all <- sapply(zeb_rmr_conv, function(z) z$rate.output[1])

# Calculate final rmr -----------------------------------------------------
zeb_rmr_final <- mean(zeb_rmr_all[c(20:88,90:105)]) 
```

```{r echo = F}
zeb_rmr_final
```

