---
title: "auto_rate.int: Automatic extraction of rates in intermittent-flow respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{auto_rate.int: Automatic extraction of rates in intermittent-flow respirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F, warning = F, message = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 3))

```


## Introduction

`auto_rate.int()` is a function for automatically extracting rates from each replicate in
intermittent-flow respirometry data using the `auto_rate()` function. You can use it to extract from each replicate the most linear (i.e. most consistently sustained) rates, the lowest or highest rates of a specific duration, or any of the other methods supported by `auto_rate`. 

This page contains descriptions and simple examples of the functionality. You should familiarise yourself with how `auto_rate` works before using `auto_rate.int`. See `auto_rate()` and [here]() for a thorough overview. See `vignette("intermittent_short")` for how `auto_rate.int` can be used to analyse a relatively brief intermittent-flow respirometry experiment, and `vignette("intermittent_long")` for an example of analysing a longer experiment.

## Overview

`auto_rate.int` uses the `starts` locations to subset each replicate from the data in `x`. It assumes the end location of each replicate is the row preceding the start of the next, or for the final replicate the last row of the dataset. It runs `auto_rate` on the specified `measure` phase of each replicate, extracting rates using the `method`, `width`, and `by` inputs which are passed to `auto_rate`. A `wait` phase can be specified to exclude data at the start of each replicate. The flushes can be excluded by using an appropriate `measure` phase. 

`auto_rate` returns multiple rate results depending on the inputs. The `n` input determines how many of these are returned to `auto_rate.int` for each replicate and saved in the summary tables. The default is `n = 1`, so for example in using the `"linear"` method this will be the most linear rate in each replicate, or if using `"lowest"` it will be the single lowest rate across the specified `width`. It is possible to return multiple rates per replicate by changing `n`, however consider carefully if this is necessary as the output will necessarily contain many more rate results which may make it difficult to explore and select results (although this is possible using `select_rate()`). 

An `auto_rate` object is saved for each replicate in the output in `$results`, and the output `$summary` table contains the results from all replicates in order with the `$rep` column indicating the replicate number. If `n` has been changed from the default to return more than one result per replicate, the `$rank` column indicates ranking of results *within* each replicate. This output object can be passed to subsequent functions such as `adjust_rate()` and `convert_rate()` for further processing. 

## Inputs

There are three required inputs: 

- `x`: The time~oxygen data from a single intermittent-flow experiment as either a `data.frame` or `inspect` object

- `starts`: The location of the start of each replicate. This can be either: 
- A single numeric value representing a regular row or time interval starting at row 1. This option should only be used when replicates cycle at regular intervals. If the first replicate does not start at row 1 of the data in `x`, it should be subset so that it does. See `subset_data()` and examples below for how to do this.  
- A numeric vector of row numbers or times indicating each replicate start location. Useful if replicates do *not* cycle at regular intervals.

- `width`: This is the width used in the `auto_rate` analysis in either rows or a time duration in the units of the raw data. It must be entered in the units of the `by` setting (`"row"` or `"time"`), which also sets the units of the `starts`, `wait` and `measure` inputs. See `auto_rate()` and [here]() for how widths are applied. Note that proportional `width` inputs accepted in `auto_rate` if `by = "row"` (e.g. `0.2` to represent 20% of the total data length) will apply to the length of the `measure` phase. 

### Other inputs 

- `wait`: A row length or time duration to be applied at the start of each replicate to *exclude* these data from any rate calculations. Can be a single value to apply the same wait phase to each replicate, or a vector of the same length as `starts` of different wait phases for each replicate. Default is `NULL` in which case no wait phase is applied.

- `measure`: This is the region within which the `auto_rate` analysis is conducted for each replicate. It is a row length or time duration to be applied at the end of the `wait` phase (if used, otherwise it applies from the start of the data), and should be the correct length to exclude the flush period. Can be a single value to apply the same measure phase to each replicate, or a vector of the same length as `starts` of different measure phases for each replicate. Default is `NULL` in which case the entire replicate is used (which is rarely what is wanted).

- `by`: Controls how the `starts`, `wait`, `measure`, and `width` inputs are applied. The default is `"row"` with `by = "time"` being the other option. We would strongly advise you use `"row"` whenever possible as, while the function has been tested, the code for parsing inputs by row is much more robust and less likely to throw up obscure bugs. (Please [provide feedback]() for **any** issues you encounter using either method). 

- `method`: Selects the rate extraction method used in `auto_rate`. The default method is `"linear"`. See `auto_rate()`.

- `n`: Sets how many rate results to extract from each replicate. Default is `1`. 

- `plot`: Controls if a plot is produced. See [Plot](#plot) section. 

## Data

The `urchin` dataset used for examples below is the included `intermittent.rd` example data with the time values changed to minutes to better demonstrate time region selection. Experimental data such as volume, weight, and row locations of replicates, flushes etc. can be found in the data help file: `?intermittent.rd`. 

```{r}
urchin <- intermittent.rd
urchin[[1]] <- urchin[[1]] / 60 # change time values to minutes
```

This is what the whole dataset look like. There are three replicates of different duration.

```{r message=FALSE, warning=FALSE, results='hide'}
urchin <- inspect(urchin)
```

Therefore in the following examples we specify three locations in `starts`. See the later [example](#regspaced) for how the `starts` of regularly spaced replicates are specified. 

## Defaults

By default, if no inputs other than `x`, `starts`, and `width` are entered, the `auto_rate` default of `method = "linear"` is applied to all the data in each replicate. This finds the most linear region of a dataset (see `vignette("auto_rate")`).

```{r results='hold'}
auto_rate.int(urchin,
              starts = c(1, 2101, 3901),
              width = 400) |>
  summary()
```

Here it performs well in identifying linear regions. However, these inputs mean the flush data is included. It is usually a good idea to use the `measure` input to ensure flushes are excluded and that only relevant data is included in the analysis. We also typically want to exclude some data at the start of a replicate right after the flush, which we can do using `wait`. All of these are applied in the default units of `by = "row"`. (We also pass `legend = TRUE` so that the phases are labelled on the plots). 

```{r results='hold'}
auto_rate.int(urchin,
              starts = c(1, 2101, 3901),
              wait = 300,
              measure = c(1500, 1100, 600),
              width = 400, 
              legend = TRUE) |>
  summary()
```
Note how the rates (yellow points) do not use the entire `measure` phase. This is because the `auto_rate` analysis occurs *within* the measure phase and the actual lengths of the rate results depends on the `method` and `width`. If you want to use the entire `measure` phase see `calc_rate.int()`. 

These results show the single top-ranked linear region in each replicate and we can see the rates are fairly consistent. Note there is one result per replicate, with the replicate number in the `rep` column. The `rank` column shows ranking of rates *within* a replicate for the particular `method` used, and are all `1` here because only one rate has been extracted. You can modify the `n` input to output more than one rate per replicate. This is what the analysis looks like if you do that. 

```{r results='hold'}
auto_rate.int(urchin,
              starts = c(1, 2101, 3901),
              wait = 300,
              measure = c(1500, 1100, 600),
              width = 400, 
              n = 3,
              plot = TRUE,
              legend = TRUE) |>
  summary()
```

Note the `rep` and `rank` columns which keep track of rates. 

These results can be passed to `adjust_rate` and `convert_rate` for further processing, however consider carefully if you really need to extract more than one rate from each replicate, or if you do be prepared to do a little extra processing of the results (see `select_rate()`).

## Lowest and highest rates

Often in intermittent-flow respirometry we want to get the lowest or highest rates from each replicate, to identify, for example, standard or active metabolic rates. We can do that using the `method`, `width`, and `by` inputs which are passed to `auto_rate`. 

In these examples we use `by = "time"` selection of regions. This is the **lowest** rate over a period of 8 minutes from each replicate:

```{r results='hold'}
auto_rate.int(urchin,
              starts = c(0, 35, 65),
              wait = 5,
              measure = c(25, 15, 15),
              method = "lowest",
              width = 8,
              by = "time") |>
  summary()
```

This is the **highest** rate over a period of 8 minutes from each replicate:

```{r results='hold'}
auto_rate.int(urchin,
              starts = c(0, 35, 65),
              wait = 5,
              measure = c(25, 15, 15),
              method = "highest",
              width = 8,
              by = "time") |>
  summary()
```
See `help("auto_rate")` and `vignette("auto_rate")` for other methods that can be applied. 

## Regularly spaced replicates {#regspaced}

For intermittent-flow experiments where the replicates are regularly spaced you do not have to specify the start location of every single one in `starts`, but can instead enter a single value indicating the interval in rows or a time duration between each one starting at the first row of the input data. 

### Specify regular replicate structure

The `zeb_intermittent.rd` example dataset uses a regular replicate structure of 9 minutes measure and 2 minutes flush (660 rows total) for most of its length. See `help("zeb_intermittent.rd")` for exact locations, but essentially after a background control and a single replicate of 14 minutes this structure is maintained until the ending background recording. Therefore to use this option we need to use `subset_data()` (or other methods in `R` to subset data) to extract only the regular replicates, and pipe the result to `auto_rate.int`. We use `starts` to specify that replicates cycle at 660 rows, impose a `wait` phase of 2 minutes, a `measure` phase of 7 minutes to exclude the flush, and use the other inputs to extract the lowest rate of 3 minutes duration.

```{r results='hide', fig.keep='last', message=FALSE}
zeb_all <- zeb_intermittent.rd |>
  
  # Inspect the data
  inspect() |>
  
  # Subset regular replicates from larger dataset
  subset_data(from = 5840,
              to = 75139,
              by = "row",
              quiet = TRUE) |>
  
  # Use auto_rate.int to get lowest 3 minute rate from each replicate, 
  # plotting first 3 and last 3 (using 'pos')
  auto_rate.int(starts = 660,
                wait = 120,
                measure = 420,
                width = 180,
                method = "lowest",
                plot = TRUE,
                pos = c(1:3, 103:105)) 
```

Here we use the `pos` input which is passed to `plot` to select the first and last three replicates for plotting to check everything looks okay. This is only for illustration here. In an actual experiment we **strongly** recommend every replicate rate result be examined visually. Up to 20 results at a time can be plotted using the `pos` input. See [Plot](#plot) section below for further details. 

### View summary

Each replicate result is saved in the `$summary` element of the output, or we can use `summary()`. 

```{r}
summary(zeb_all)
```

This object is now ready to be passed to `adjust_rate()` and `convert_rate()`. Final rate selection can be handled using `select_rate()` after conversion. See `vignette("select_rate")` and `vignette("intermittent_long")` for examples. 

## Plot {#plot}

If `plot = TRUE`, a plot is produced of each replicate rate result on a grid up to a total of 20. By default this is the first 20 (i.e. `pos = 1:20`). Others can be selected by modifying the `pos` input, either in the main function call or when calling `plot()` on output objects. 

```{r results='hide'}
plot(zeb_all, pos = 50:69)
```

There are three ways in which these `auto_rate.int` results can be plotted, selected using the `type` input in either the main function call or when calling `plot()` on output objects.
For all plots, the bottom blue time axis shows the time values of the original larger dataset, whilst the upper red row axis shows the rows of the replicate subset. 

### `type = "rep"`

The default is `type = "rep"` in which each replicate is plotted individually with the rate region highlighted. The rate data is the yellow points, the red shaded region is the `wait` phase, and the green shaded region the `measure` phase.  

```{r results='hide'}
urch <- auto_rate.int(urchin,
                      starts = c(1, 2101, 3901),
                      wait = 300,
                      measure = c(1500, 1100, 600),
                      width = 400, 
                      plot = TRUE,
                      type = "rep",
                      legend = TRUE)
```

### `type = "full"`

Entering `type = "full"` will show each replicate rate highlighted in the context of the entire dataset.

```{r results='hide'}
auto_rate.int(urchin,
              starts = c(1, 2101, 3901),
              wait = 300,
              measure = c(1500, 1100, 600),
              width = 400, 
              plot = TRUE,
              type = "full")
```

Note this may be of limited use when the dataset is large. 

```{r results='hide'}
plot(zeb_all, type = "full", pos = 50)
```

### `type = "ar"`

Lastly, the `auto_rate` result (as found in the `$results` element of the output) for each individual row of the summary table can be plotted using `type = "ar"` and the `pos` input. Note these plots show only the `measure` phase of the data and the rate result from within it. 

```{r results='hide'}
plot(zeb_all, type = "ar", pos = 50)
```

The `pos` input here can also be of multiple replicates but this will produce multiple individual plots. 

### Additional plotting options

`legend` can be used to show labels (default is `FALSE`). `quiet` suppresses console output. Additional plotting parameters can also be passed to adjust margins, axis label rotation, etc. See [here](). 

## S3 generic methods

As well as `plot`, saved `auto_rate.int` objects work with the generic S3 methods `print`, `summary`, and `mean`

### `print`

This simply prints a single summary row rate result to the console, by default the first one. The `pos` input can be used to print others. 

```{r}
print(zeb_all)
print(zeb_all, pos = 50)
```

### `summary`

This prints the summary table to the console which contains linear model coefficients and other metadata for each replicate rate. The `pos` input can be used to select which replicates (`$rep` column) to include. 

```{r}
summary(zeb_all)
summary(zeb_all, pos = 1:4)
```

`export = TRUE` can be used to export the summary table as a data frame, or those rows selected using `pos`

```{r results='hide'}
zeb_exp <- summary(zeb_all, 
                   pos = 1:4, 
                   export = TRUE)
```

```{r}
zeb_exp
```

### `mean`

This averages all replicate rates in the `$rate` column, or those selected using `pos`. The result can be saved as a value by using `export = TRUE`.

```{r}
zeb_mean <- mean(zeb_all, pos = 1:4, export = TRUE)
zeb_mean
```

This could be passed as a numeric value to later functions such as `convert_rate()`, although it is usually better to do this kind of result filtering after conversion. 


<!-- ## Notes and tips -->

<!-- - Note that . -->
