---
title: "auto_rate: Automatic detection of metabolic rates"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{auto_rate: Automatic detection of metabolic rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F, warning = F, message = F}
library(knitr) # load knitr to enable options
library(respR) # load respR
library(DiagrammeR) # flowcharts... and more

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                                scipen = 999, 
                                digits = 3))
```

## Introduction

In respirometry, we want to report oxygen uptake or production rates from experimentally important stages or that represent behavioural or physiological states. These include:

- the **most linear**, that is most consistent or monotonic rates observed, often most representative of *routine* metabolism
- the **lowest** rates observed, often most representative of *standard*, *resting*, *basal* or *maintenance* metabolism
- the **highest** rates observed, representative of *active* or *maximum* metabolic rates

Identifying and extracting these rates from large datasets is difficult, and if selected visually, subject to observer bias and lack of objectivity. 
Methods such as fitting multiple, fixed-width linear regressions over an entire dataset to identify regions of lowest or highest slopes (i.e. rates) is computationally intensive, and slopes found via this method highly sensitive to the width chosen, especially if the specimen's metabolic rate changes rapidly or the data is noisy. 

Here we detail `auto_rate()`, a function that uses machine learning techniques to *automatically* detect the **most linear** regions of a dataset. This allows an investigator to extract rates in a statistically robust, objective manner. It can also extract and order **highest** and **lowest** rates, or return an unordered **rolling** rate across the whole dataset. 

In this vignette we detail how `auto_rate` works, and how it can be used to extract rates from respirometry data. 
Importantly, `auto_rate` has been optimised to be *extremely fast*. Other methods on large datasets can take minutes, hours or even days to run. 
`auto_rate` can reduce this wait by orders of magnitude, fitting tens of thousands of regressions and detecting linear regions in seconds. 

## Overview

This illustrates the main processes involved in `auto_rate`:

```{r, echo = F, fig.height = 6, fig.align = "left"}
library(DiagrammeR)
grViz("
digraph a_nice_graph {

# node definitions with substituted label text

node [fontname = Helvetica, shape = circle]
rd [label = 'Raw Data']

node [fontname = Helvetica, shape = rectangle]
rr [label = 'Rolling linear regression']
kd [label = 'Kernel density\nestimates']
re [label = 'Bandwidth\nresampling']
lr [label = 'Localised linear\nregression']
ar [label = 'Rank']
fi [label = 'Filter']
un [label = 'Unorder']
or [label = 'Order']

node [fontname = Menlo, shape = none]
au [label = 'auto_rate()']
li [label = 'linear']
mm [label = ' highest\nlowest\nmaximum\nminimum']
in [label = 'interval']
ro [label = 'rolling']

# edge definitions with the node IDs

rr -> kd -> re -> lr -> ar -> li
rd -> au -> rr -> or -> mm
rr -> fi -> in
rr -> un -> ro

}")
```


`auto_rate` works by performing an optimised rolling regression on the dataset of a specified width.  For the `linear` method, a kernel density estimate is performed on the rolling regression output, and the kernel bandwidth used to re-sample linear regions of the data for re-analysis. For other methods, the results are filtered, ordered, or returned unordered. 

## Rolling linear regression

The function `auto_rate` uses a novel method of combining rolling regression and kernel density estimate algorithms to detect patterns in time series data. The rolling regression runs all possible ordinary least-squares (OLS) linear regressions $(y = \beta_0 + \beta_1 X + \epsilon)$ of a fixed sample width across the dataset, and is expressed as: $$y_t(n) = X_t(n) \beta (n) + \epsilon_t(n), \ t = n,\ ...,\ T$$ where $T$ is the total length of the dataset, $n$ is the window of width $n < T$,  $y_t(n)$ is the vector of observations (e.g. oxygen concentration), $X_t(n)$ is the matrix of explanatory variables, $\beta (n)$ is a vector of regression parameters and $\epsilon_t(n)$ is a vector of error terms.
Thus, a total of $(T - n) + 1$ number of overlapping regressions are fit. 

## Methods

`auto_rate` has several methods to process, order or filter the rolling regression results. 

### `method = "linear"`

<!-- Additional methods are automatically applied when we use `auto_rate` to detect linear sections of the data.  -->
<!-- First, a Gaussian kernel density estimate (KDE) is used to process the regression coefficient $\beta$, based on the key assumption that linear sections of the data are reflected by relatively stable parameters across the rolling estimates. -->
<!-- This assumption is used often in financial statistics to evaluate stability and make predictions on time-series data. -->
<!-- Here, KDE automatically aggregates stable (i.e. linear) sections as they naturally form one or more local maxima ("modes") in the probability density estimate, and if the distribution is multimodal, the modes are ranked by size. -->

This method uses kernel density estimation (KDE) to automatically identify linear regions of the dataset.

First, we take advantage of the key assumption that linear sections of a data series are reflected by stable parameters across the rolling estimates, a property that is often applied in financial statistics to evaluate model stability and make forward predictions on time-series data (see Zivot and Wang 2006).
We use kernel density estimation (KDE) techniques, often applied in various inference procedures such as machine learning, pattern recognition and computer vision, to automatically aggregate stable (i.e. linear) segments as they naturally form one or more local maxima ("modes") in the probability density estimate.

KDE requires no assumption that the data is from a parametric family, and learns the shape of the density automatically without supervision. 
KDE can be expressed as: $$\hat{f}(x) = \frac{1}{nh^d}\sum_{i = 1}^{n} K \left(\frac{x - X_i}{h} \right)$$ where $f$ is the density function from an unknown distribution $P$ for $X_1,...,X_n$, $K$ is the kernel function and $h$ is the optimal smoothing bandwidth. 
The smoothing bandwidth is computed using the solve-the-equation *plug-in* method (Sheather et al. 1996, Sheather and Jones 1991) which works well with multimodal or non-normal densities (Raykar and Duraiswami 2006).

<!-- The smoothing bandwidth $h$ is computed using an adjusted kernel bandwidth selector based on Silvermans "rule of thumb" (Silverman, 1986): $$h = \left(\frac{4\hat{\sigma}^5}{3n} \right)^{\frac{1}{5}} \approx 1.06\hat{\sigma}n^{-\frac{1}{5}}$$ where $\hat{\sigma}$ is the standard deviation of the samples and $n$ is the total number of samples.  -->

We then use $h$ to select all values in the rolling regression output that match the range of values around each mode ($\theta_n$) of the KDE (i.e. $\theta_n \pm h$).
These rolling estimates are grouped and ranked by size, and the upper and lower bounds of the data windows they represent are used to re-select the linear segment of the original data series. 
The rolling estimates are then discarded while the detected data segments are analysed using linear regression. 

The output will contain only the regressions identified as coming from linear regions of the data, ranked by order of the KDE density analysis. This is present in the `$summary` component of the output as `$density`. Under this method, the `width` input is used as a starting seed value, but the resulting regressions may be of any width.

#### When this method might be applied

This method could be applied to virtually any respirometry data when you are looking for linear regions where rates are *stable*, *consistent* and *representative* rates for the behavioural or physiological state of the specimen. This could be a routine metabolic rate, standard or basal metabolic rate, or in the case of an animal under constant exercise a consistent active metabolic rate. 

### `method = "lowest"`

Every regression of the specified `width` across the timeseries is calculated, then ordered using *absolute* rate values from lowest to highest. This option can only be used when rates all have the same sign, and it essentially ignores the sign. Rates will be ordered from lowest to highest in the `$summary` table by absolute value regardless of if they are positive or negative.

#### When this method might be applied

When you are looking for the lowest rates across a specific time or row window, representative of resting, basal or standard metabolic rates. 

### `method = "highest"`

Every regression of the specified `width` across the timeseries is calculated, then ordered using *absolute* rate values from highest to lowest. This option can only be used when rates all have the same sign, and it essentially ignores the sign. Rates will be ordered from highest to lowest in the `$summary` table by absolute value regardless of if they are positive or negative.

#### When this method might be applied

When you are looking for the highest rates across a specific time or row window, representative of maximum or active metabolic rates. 

### `method = "minimum"`, `method = "maximum"`

These methods are strictly numerical and take full account of the sign of the rate. In `respR` oxygen uptake rates are negative since they represent a negative slope of oxygen against time, and oxygen production rates are positive. 

Every regression of the specified `width` across the entire timeseries is calculated, then ordered using *numerical* rate values from minimum to maximum for the `minimum` method, or vice versa for `maximum`. Generally this method should only be used when rates are a mix of oxygen consumption and production rates, such as when positive rates may result from regressions fit over flush periods in intermittent-flow respirometry. 

#### When these methods might be applied

Generally, for most analyses where high or low rates are of interest the `highest` or `lowest` methods should be used instead. However, when rates are a mix of negative and positive and you want the highest or lowest these can be used, but note they order by *numerical* value; the highest oxygen *uptake* rates will be most `minimum`. 

### `method = "rolling"`

This method returns all regressions of the specified `width` in sequential order across the dataset. All results are returned in the summary table. 

#### When this method might be applied

This method can be applied when you want to extract a rolling rate of a specified width for further analyses. Alternatively, if you don't want to rely on the `"linear"` selection or other methods but want to filter and select the results manually according to various criteria using `subset_rate()`. See `vignette("subset_rate")`. 

### `method = "interval"`

Multiple, successive, **non**-overlapping regressions of the specified 'width' are extracted from the rolling regressions, ordered sequentially. 

#### When this method might be applied

This is chiefly for comparison with historical data, or if you have a specific reason to specify non-overlapping regressions. For example, in P~crit~ measurements, it was historically common for users to generate rate~oxygen data by interval-based regressions.

## Adjusting the width {#width}
 
By default, `auto_rate` rolling regression uses a rolling window width in rows of `0.2` multiplied by the total rows of the dataset, that is across a rolling window of 20% of the data. 
This can be changed using the `width` input to a different relative proportion (e.g. for 10% `width = 0.1`). Alternatively, if not between 0 and 1, the `width` by default equates to a fixed value in rows (e.g. `width = 2000, by = "row"`), or can be entered as a fixed value in the time metric (`width = 3000, by = "time"`).  

Note that `by = "row"` is computationally faster. Specifying a `"time"` window tells `auto_rate` that the time data may have gaps or not be evenly spaced, and so the function calculates each time width using the raw time *values*, rather than assuming a specific row width represents the same time window, a less computationally intensive process. If the data are without gaps and evenly spaced with regards to time, `by = "row"` and the correct row `width` to represent the time window you want will be much faster. 

The `width` determines the exact width of the data segments produced for `highest`, `lowest`, `rolling` etc. rates. This allows the user to consistently report results across experiments, such as reporting the highest or lowest rates sustained over a specific time period. 

Importantly however, for the `linear` detection method the `width` is a starting seed value, and does *not* restrict the width of the segments produced. The minimum width of the segments tends to be close to or slightly lower than the `width` input (though not always), however the upper widths are not restricted and can be of any width if the segments are found to be linear. 

While you should experiment with different values to see the effects on the analysis, users should use caution when changing the `width`. Choosing an inappropriate width tends to overfit or underfit the rolling rates. See [Prinzing et al. 2021](https://januarharianto.github.io/respR/articles/refs.html#references) for an excellent discussion of appropriate widths in rolling regressions to determine maximum metabolic rates, much of which is relevant to extracting rates of any kind. 

### Overfitting

Below, we show the differences in the shape of the rolling regressions when using the default `width = 0.2` versus a value of `0.6` with the dataset `sardine.rd`:

```{r rollreg_demo, fig.keep='none', warning=FALSE}
# Perform linear detection; default width when not specified is 0.2:
normx <- auto_rate(sardine.rd)

# Perform linear detection using manual width of 0.6:
overx <- auto_rate(sardine.rd, width = 0.6)
```

```{r echo = F, message=FALSE, results='hide'}
# rolling regression

r2 <- adjustcolor("goldenrod1", alpha.f = 1)  

rollreg.plot <- function(rolldf, ranked.b1, rate.rev = TRUE, 
                         xlim = NULL, ylim = NULL) {
  if(is.null(xlim)) xlim <- range(rolldf[[1]])
  if(is.null(ylim)) ylim <- range(rolldf[[2]])
  if(rate.rev) ylim <- rev(ylim) ## reverse y-axis
  plot(rolldf, xlab = "", ylab = "", bg = r2, col = r2,
       xlim = xlim,
       ylim = ylim,
       pch = 21,
       lwd = 1, 
       cex = 0.9, 
       panel.first = grid(lwd = .7),
       tck = -0.02, 
       mgp = c(0, 0.4, 0))
  # rect(par("usr")[1],par("usr")[3],par("usr")[2],par("usr")[4],col = r3)
  abline(h = ranked.b1, lty = 2)
  title(main = ("Rolling Rate vs Time"), line = 0.3, font = 2)
}

# Plot ONLY the rolling regression plots for comparison:
par(mfrow = c(1, 2), 
    mai = c(0.4, 0.4, 0.3, 0.3),
    oma = c(0.4, 0.4, 0.3, 0.3),
    cex.main = 1,
    cex = 0.9)

rolldt_norm <- data.table::data.table(x = (normx$roll$endtime + normx$roll$time)/2, 
                                      y = normx$roll$rate)
rate_norm <- normx$summary$rate_b1[1]
rollreg.plot(rolldt_norm, rate_norm, rate.rev = TRUE)

rolldt_over <- data.table::data.table(x = (overx$roll$endtime + overx$roll$time)/2, 
                                      y = overx$roll$rate)
rate_over <- overx$summary$rate_b1[1]
rollreg.plot(rolldt_over, rate_over, rate.rev = TRUE, 
             xlim = range(rolldt_norm[[1]]), 
             ylim = range(rolldt_norm[[2]]))

```


For the `linear` method, since KDE automatically aggregates stable values, a poor selection of the `width` may result in a badly-characterised rolling rate estimate output.
Under perfectly linear conditions, that is completely monotonic rates, we would expect a rolling regression output plot such as this to consist of a straight, horizontal line. 
In these data, while the default width allowed a pattern of relative stability in rate after around 2500 seconds to be identified, this information was lost when a `width` of `0.6` was used, with stable rates only being identified much later in the dataset.

Similarly, if we are interested in highest rate values, under the lower width input we could see values around -0.0010 occurring within the first 1000s of the experiment. This information was also completely lost under the higher width input. 

### Underfitting

By contrast, if the `width` is too low the rolling rate is unstable and heavily influenced by data noise and residual variation. This leads to poor results for the `linear` method, and also highly variable results under the other methods. 

Here we'll compare the default `width = 0.2` to a lower value of 5% of the data, `width = 0.05`. 

```{r rollreg_demo_2, fig.keep='none', warning=FALSE}
# Perform linear detection; default width when not specified is 0.2:
normx <- auto_rate(sardine.rd)

# Perform linear detection using manual width of 0.05:
underx <- auto_rate(sardine.rd, width = 0.05)
```

```{r echo = F, message=FALSE, results='hide'}
# rolling regression

r2 <- adjustcolor("goldenrod1", alpha.f = 1)  

rollreg.plot <- function(rolldf, ranked.b1, rate.rev = TRUE, 
                         xlim = NULL, ylim = NULL) {
  if(is.null(xlim)) xlim <- range(rolldf[[1]])
  if(is.null(ylim)) ylim <- range(rolldf[[2]])
  if(rate.rev) ylim <- rev(ylim) ## reverse y-axis
  plot(rolldf, xlab = "", ylab = "", bg = r2, col = r2,
       xlim = xlim,
       ylim = ylim,
       pch = 21,
       lwd = 1, 
       cex = 0.9, 
       panel.first = grid(lwd = .7),
       tck = -0.02, 
       mgp = c(0, 0.4, 0))
  # rect(par("usr")[1],par("usr")[3],par("usr")[2],par("usr")[4],col = r3)
  abline(h = ranked.b1, lty = 2)
  title(main = ("Rolling Rate vs Time"), line = 0.3, font = 2)
}

# Plot ONLY the rolling regression plots for comparison:
par(mfrow = c(1, 2), 
    mai = c(0.4, 0.4, 0.3, 0.3),
    oma = c(0.4, 0.4, 0.3, 0.3),
    cex.main = 1,
    cex = 0.9)

rolldt_norm <- data.table::data.table(x = (normx$roll$endtime + normx$roll$time)/2, 
                                      y = normx$roll$rate)
rate_norm <- normx$summary$rate_b1[1]

rolldt_over <- data.table::data.table(x = (underx$roll$endtime + underx$roll$time)/2, 
                                      y = underx$roll$rate)
rate_over <- underx$summary$rate_b1[1]

rollreg.plot(rolldt_norm, rate_norm, rate.rev = TRUE,
             xlim = range(rolldt_over[[1]]), 
             ylim = range(rolldt_over[[2]]))

rollreg.plot(rolldt_over, rate_over, rate.rev = TRUE, 
             xlim = range(rolldt_over[[1]]), 
             ylim = range(rolldt_over[[2]]))

```

A lower width leads to much more variable rolling rate estimates. Note how we have had to adjust the y-axis limits to fit the results (the left plot is the same data as in the previous section with different axis values). In this particular analysis (results not shown) the `linear` method performed poorly. If we were interested in highest or lowest rates, this would also prove problematic since the rates are so variable. 

### Appropriate widths

The `width` value should be carefully considered; too low and it fails to capture accurate rolling rates and is unduly influenced by data noise or variability, too high and the data is overfitted with important physiological or behavioural states smoothed out. Whatever value is used, this should be reported in the analytical methods alongside results. 

See [Prinzing et al. 2021](https://januarharianto.github.io/respR/articles/refs.html#references) for an excellent discussion of appropriate widths in rolling regressions to determine maximum metabolic rates, much of which is relevant to extracting rates of any kind. 

## Example Analysis

Here we'll run through examples of how to use `auto_rate` to extract rates from respirometry data. 

### Automatic detection of linear rates

By default, `auto_rate` identifies the *most linear* regions of the data (i.e. `method = "linear"`):

```{r results='hide', warning=FALSE}
sard_ar <- auto_rate(sardine.rd)
```

This method detects the *most consistently linear* regions of the data, that is the most consistent rates observed during the experiment. It does this in a rigorous, unsupervised manner, with the advantage being that this removes observer subjectivity in choosing which rate is most appropriate to report in their results. 
It is a statistically robust way of indentifying and reporting consistent rates in respirometry data, such as those representative of routine or standard metabolic rates. 

### Interpreting the plots {#plot1}

The `linear` method uses the input `width` as a starting seed value to calculate a rolling rate (panel 3). It then uses these rates to identify linear regions using kernel density estimation (KDE, panel 6). Peaks in this plot represent linear regions, that is areas of stable rates at that `width` representative of that region as a whole. It then re-samples these regions and runs additional linear analysis at different widths to arrive at a final rate. This is why the final, high ranked rates tend to be over widths greater than the input `width`, as can be seen here with the top ranked result. 

Generally, the higher and wider the peak in the KDE plot, the more linear the region. Here there are several (the current plotted one denoted by a dashed line) and the strongest results are towards the end of the data, including the highest ranked result. The rate value of the current plotted result can also be seen in panel 3 as a horizontal line. This can help assess if it is a representative rate, although bear in mind these rolling rates are at a different width. 

See [Plot](#plot2) section below for more information.

### Exploring the results

Typically, `auto_rate` will identify multiple linear regions. These are ranked using the kernel density analysis, with the results reflected in the ordering of the `$summary` table in the output, which is ordered by the `$density` column. The first row is the top ranked, most linear region, and subsequent rows progressively lower rank. By default, this highest ranked result is returned when `print` or `plot` are used, but other results can be output using the `pos` input with those functions.  

```{r}
print(sard_ar, pos = 2)
```

```{r results='hide'}
plot(sard_ar, pos = 2)
```

**Users should take special note** that as an automated, unsupervised machine learning method of identifying linear data `auto_rate` is fallible, and the results should *always* be inspected and explored. In this case the function has identified a total of 46 linear regions. They can be viewed by calling `summary()`

```{r}
summary(sard_ar)
```

In this case the first rate result looks good: it has a high r-squared, is sustained over a duration of 50 minutes, and the rate value is consistent with the other results. 
However in some cases, other ranked results may be more appropriate to report depending on the metabolic rate metric being investigated. 

While the first result is the highest in terms of the kernel density value, the user is free to select and report other linear results if they satisfy other desirable criteria, for instance are above a particular r-squared value or span a minimum time window. One peculiarity to take note of in the KDE analysis, is that sometimes the top ranked result does not necessarily have the highest r-squared. This is a counter intuitive result, but explained by the fact that the function learns the shape of the entire dataset, so a particular regression from a linear region might be most representative of the rates in that region, but just happen to be fit to values which return a lower r-squared. This does not mean it is not a valid result; arguably, it is more valid, since it is accurately describing the localised shape of the data. See Chabot et al. ([2021](https://januarharianto.github.io/respR/articles/refs.html#references)) for discussion of r-squared values in metabolic rate measurements.

The `pos` input can also be used in `summary` to view particular row ranges. We'll look at the first 10. 

```{r}
summary(sard_ar, pos = 1:10)
```

Here, the 6th ranked result, while being a valid linear region, is conspicuously higher in rate value and occurs close to the start of the experiment. If we were interested in routine or standard metabolic rates, we would want to exclude this one, as it suggests the specimen might not yet be acclimated to the chamber. 

In most cases the best option is to use the top ranked result unless there are specific reasons to exclude it. However, an investigator may opt to select several and take a mean of the resulting rates. For instance here, we might decide on a mean of the top 3 since they have the highest `$density` values (note that typically we would do this after rates have been adjusted and converted - see later sections - but it is possible here too). We could just average the top 3 values ourselves, but the `mean` function will also work with `auto_rate` objects and accepts the `pos` input.

```{r}
mean(sard_ar, pos = 1:3)
```

Any rate value determined after such selection can be saved as a variable, or entered manually as a value in later functions such as `adjust_rate` and `convert_rate`. It can also be exported as a value by using `export = TRUE` in the `mean` call. However, see next section.  

### `subset_rate`

Alternative selection criteria might also be applied. This might include excluding all results below a certain r-squared, use only the top n'th percentile of results, or exclude those from the certain stages of the experiment. 

New in `respR v2.0` is the `subset_rate()` function which can apply these criteria to `auto_rate` results, along with many others. 

See `vignette("subset_rate")` for how to do advanced filtering of `auto_rate` results. 

### Highest rates

`auto_rate` can also be used to detect the *highest* and *lowest* rates over a fixed `width`. This allows for consistent reporting of respirometry results, such as the highest or lowest rates sustained over a specified time period. 

Here we want to know the highest rates sustained over 15 minutes, or 900s, in the `sardine.rd` data. Since in these data, oxygen is recorded every second  and `inspect()` tells us the time data is gapless and evenly spaced, we can simply specify width in the same number of rows.

```{r fig.keep='none'}
sard_insp <- inspect(sardine.rd)
```

```{r results='hide'}
high_rate <- auto_rate(sard_insp, width = 900, by = "row", method = "highest")
```

```{r}
summary(high_rate)
```

In the `highest` and `lowest` methods the rates are ordered by the *absolute* rate value, regardless of the sign. The top results here have the same rate value as printed, but likely have small differences at higher precision (what you see printed depends on your own R `options()` settings). Again, a user may choose to report the top result or perform further selection and filtering using `subset_rate()`. This includes the option to remove rates which overlap, that is share the same rows of data; as we can see here the top results all come from the same region of the data.

### Lowest rates

We can similarly find the `lowest` rate over 15 minutes.

```{r results='hide'}
low_rate <- auto_rate(sard_insp, width = 900, method = "lowest")
```

```{r}
print(low_rate)
```

```{r}
summary(low_rate)
```

Note, the output objects of the `highest` and `lowest` methods are essentially identical, the only difference being the results are ordered descending or ascending by absolute rate value. 

### Rolling rates

The `rolling` method allows a rolling regression of the specified `width` to be returned in sequential order. 

```{r results='hide'}
roll_rate <- auto_rate(sard_insp, width = 900, method = "rolling")
```

```{r}
summary(roll_rate)
```

This outputs every regression of the `width` in order. The main utility of this method is for passing to `subset_rate` where various criteria can be applied to filter the results manually. See `vignette("subset_rate")`.

## Further processing

Saved `auto_rate` objects can be passed to subsequent `respR` functions for further processing, such as `adjust_rate()` to adjust for background respiration, or `convert_rate()` to convert to final oxygen uptake units. See other vignettes for examples of these operations. 

## Plot {#plot2}

When using `auto_rate`, a plot of the results is produced (unless `plot = FALSE`). If there are multiple results each can be plotted individually using `pos`. Each panel can be plotted on its own using `panel` with values 1 to 6. If labels or legends obscure parts of the plot they can be suppressed using `legend = FALSE`. Console output messages can be suppressed with `quiet = TRUE`. Lastly, the rolling rate plot can be plotted on an unreversed y-axis with `rate.rev = FALSE`, if for instance you are examining oxygen production rates. 

The first panel is the complete timeseries of oxygen against both time (bottom blue axis) and row index (top red axis) with the `pos` rate regression (defualt is `pos = 1`) highlighted in yellow. The next plot is a close-up of this rate region. The next is a rolling rate plot across the entire timeseries at the input `width` (see [here](#width)). The next two are diagnsotic plots of the fitted values vs. residuals for the current rate result. Lastly, the sixth panel (only for the `linear` method) is a plot of the kernel density analysis output, which is the density peaks of stable rate values. 

See [Interpreting the plots](#plot1) section above for more information.

## Notes

- `auto_rate` does not currently support analysis of flowthrough respirometry data. See `vignette("flowthrough")` for analysis of these experiments. 

- The `auto_rate` `linear` method works best with data that is fairly monotonic, that is shows an either downward or upward trend without strong fluctuations. With intermittent-flow respirometry data there is a strong possibility flush periods will confuse the `auto_rate` algorithms, so it should typically only be run on subsets of the data containing actual specimen measurements. The `subset_data` function is ideal for subsetting and passing data regions of interest to other functions. See `vignette("intermittent_long")`.
