---
title: "calc_rate.int: Manual extraction of rates in intermittent-flow respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{calc_rate.int: Manual extraction of rates in intermittent-flow respirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F, warning = F, message = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 3))

```


## Introduction

`calc_rate.int()` is a function in `respR` for manually determining rates across
user-defined ranges of time, row or oxygen value in each replicate in
intermittent-flow respirometry data. This page contains descriptions and simple
examples of the functionality. See also the help file
at `help("calc_rate.int")`, go [here]() for an example of how it can be used in
analysis of a relatively brief intermittent-flow respirometry experiment, and
[here]() for an example of analysing a much longer experiment.

## Overview

How `calc_rate.int` works is fairly straightforward. 
The function uses the `starts` and optional `ends` locations to subset each replicate from the data in `x`. It runs `calc_rate` on each subset, extracting a rate between the `from` and `to` inputs in the `by` metric. See `calc_rate()` and [here]() for full details of how these inputs are applied by `calc_rate`. 

A `calc_rate` object is saved for each replicate in the output in `$results`, and the output `$summary` table contains the results from all replicates in order with the `$rank` column indicating the replicate number. This output object can be passed to subsequent functions such as `adjust_rate()` and `convert_rate()` for further processing. 

## Minimum requirements 

There are two required inputs: 

- `x`: The time~oxygen data from a single
intermittent-flow experiment as either a `data.frame` or an `inspect` object

- `starts`: The location of the start of each replicate. This
can be either a numeric vector of the row number for each replicate start location, or in
the case of regularly spaced replicates a single numeric value representing a
regular interval from the first row. If the first replicate does not start at row 1 of the data in `x`, see `subset_data()` and examples below for how to subset it first. This option should only be used when replicates cycle at regular intervals of row numbers. 


## Other inputs 

- `ends`: The end row locations of each replicate. Optional, but useful for removing flush periods and subsetting only relevant data. If `starts` is a numeric vector of starting row locations, this is simply a vector of equal length indicating end row locations for each replicate, or where you want each replicate to end to remove the flush. If `starts` is a single value indicating regularly cycling replicates, `ends` specifies the duration in rows of each replicate. Essentially this is added internally to the `starts` locations to get the `ends` locations. For both options, if left as `ends = NULL` the function simply assumes the end location of each replicate is the row preceding the start of the next, or for the final replicate the last row of the dataset. 

- `from`, `to`, `by`: For selecting the region within each replicate from which to extract a rate. See `calc_rate()`.

- `plot`: Controls if a plot is produced. See [Plot](#plot) section. 

## Differences from `calc_rate`

There are two major differences from `calc_rate`: 

- Only one rate can be extracted per replicate, and this must be the same data region in each replicate as specified via `from`, `to`, and `by`.

- The `by = "proportion"` method is not supported.

If you need to get around these limitations there are multiple ways using regular `R` methods to iterate `calc_rate` over intermittent-flow data. See [here](). 

## Data

The `urchin` dataset used for examples below is the included `intermittent.rd` example data with the time values changed to minutes to better demonstrate time region selection. Experimental data such as volume, weight, and row locations of replicates, flushes etc. can be found in the data help file: `?intermittent.rd`. 

```{r}
urchin <- intermittent.rd
urchin[[1]] <- urchin[[1]] / 60
```

This is what the whole dataset look like. 

```{r message=FALSE, warning=FALSE, results='hide'}
urchin <- inspect(urchin)
```

## Defaults

By default, if no inputs other than `x` and `starts` are entered, the `calc_rate` default behaviour of  calculating a rate across an entire dataset as entered is applied to each replicate. 

```{r tidy=FALSE, results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751)) 
```

Here, this is obviously not going to produce an appropriate rate. To exclude flush periods or other undesirable regions, enter `ends` locations for each replicate.

```{r tidy=FALSE, results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              ends = c(1900, 3550, 4831)) 
```

However, we usually want to further exclude some parts of the data, for example the first few minutes to allow a period of settling or acclimation. In intermittent-flow experiments these are often described as as 'flush', 'wait', and 'measure' periods. We can use the region selection functionality of `calc_rate` to achieve this.  

## Rate from `time` range in each replicate

The default method by which a rate region can be specified in each replicate is `by = "time"`. Note that in `calc_rate.int` the `from` and `to` inputs are *relative* to the replicate as subset using `starts` and `ends`, not the original larger dataset. Essentially the rate is extracted between the `from` and `to` times *within* each replicate. 

```{r results='hide'}
# Get rate from 5th to 20th minute within each replicate
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              from = 5,
              to = 20,
              by = "time")
```

The function uses the closest matching values if the exact values do not occur in the time data. If the `to` value lies outside the actual time data range it will use the last value instead, as seen here for the third replicate which is shorter than the others. 

In the case of an experiment like this where the replicates are not the same length but we want to use all available data before the flush we can use the `ends` input to exclude the flush, and not specify a `to` input in which case the default behaviour is to calculate a rate to the end of the subset data. 

```{r results='hold'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              ends = c(1900, 3550, 4831),
              from = 5,
              by = "time") |>
  summary()
```


## Rate from `row` range in each replicate

Similarly the rate region can be specified `by = "row"`. Again, the `from` and `to` row inputs are *relative* to the subset replicate, not the original larger dataset. 

```{r results='hide'}
# Get rate from row 500 to 1000 within each replicate
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              from = 500,
              to = 1000,
              by = "row")
```

## Rate from `oxygen` range in each replicate

Lastly, the region can be specified `by = "oxygen"`. This option finds the first occurrence of the `from` value and the last occurrence of the `to` value (or the closest matching values) within each replicate. 

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              from = 7.0,
              to = 6.8,
              by = "oxygen")
```

This result is clearly not what we want. Oxygen range selection requires careful subsetting of the replicates because the `to` value will also likely occur as oxygen increases during flushes. We can instead use the `ends` input to exclude flushes. 

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              ends = c(1900, 3550, 4831),
              from = 7.0,
              to = 6.8,
              by = "oxygen")
```

## Regularly spaced replicates

For intermittent-flow experiments where the replicates are regularly spaced you do not have to specify the start location of every single one in `starts`, but can instead enter a single value indicating the interval in rows between each one from the first row of the input data. 

The `zeb_intermittent.rd` example dataset uses a regular replicate structure of 9 minutes measure, 2 minutes flush (660 rows) for most of its length. See `help("zeb_intermittent.rd")` for exact locations, but essentially after a background control and a single replicate of 14 minutes this structure is maintained until the ending background recording. Therefore to use this option we need to use `subset_data()` to extract only the regular replicates, and pipe the result to `calc_rate.int`. 

```{r results='hide', fig.keep='last', message=FALSE}
zeb_all <- zeb_intermittent.rd |>
  
  # inspect the data
  inspect() |>
  
  # subset regular replicates from larger dataset
  subset_data(from = 5840,
              to = 75139,
              by = "row",
              quiet = TRUE) |>
  
  # calc rate in each one from row 100 to 500, plotting first 3 and last 3
  calc_rate.int(
    starts = 660,
    from = 100,
    to = 500, 
    by = "row",
    plot = TRUE,
    pos = c(1:3, 103:105))

```

Here we use the `pos` input which is passed to `plot` to select the first and last three replicates for plotting to check everything looks okay. This is only for illustration here. In an actual experiment we **strongly** recommend every replicate rate result be examined visually. Up to 20 results can be plotted using the `pos` input. See [Plot](#plot) section below for further details. 

Each replicate result is saved in the `$summary` element of the output, or we can use `summary()`. 

```{r}
summary(zeb_all)
```

The replicate number is indicated by the `$rank` column. 

## Plotting {#plot}

If `plot = TRUE`, a plot is produced of each replicate rate result on a grid up to a total of 20. By default this is the first 20 (i.e. `pos = 1:20`). Others can be selected by modifying the `pos` input, either in the main function call or when calling `plot()` on output objects. 

```{r results='hide'}
plot(zeb_all, pos = 50:69)
```

For all plots, the bottom blue time axis shows the time values of the original larger dataset, whilst the upper red row axis shows the rows of the replicate subset. 

There are three ways in which these `calc_rate.int` results can be plotted, selected using the `type` input in either the main function call or when calling `plot()` on output objects.

### `type = "rep"`

The default is `type = "rep"` in which each replicate is plotted individually with the rate region highlighted. 

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              from = 5,
              to = 20,
              by = "time")
```

### `type = "full"`

Entering `type = "full"` will show each replicate rate highlighted in the context of the entire dataset.

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3751),
              from = 5,
              to = 20,
              by = "time",
              type = "full")
```

This is of limited use when the dataset is large. 

```{r results='hide'}
plot(zeb_all, type = "full", pos = 50)
```

### `type = "cr"`

Lastly, each individual replicate `calc_rate` result (as found in the `$results` element of the output) can be plotted using `type = "cr"` and the `pos` input. 

```{r results='hide'}
plot(zeb_all, type = "cr", pos = 50)
```

The `pos` input here can also be of multiple replicates but this will produce multiple individual plots that you will need to scroll through. 

### Additional plotting options

`legend` can be used to show axis labels (default is `FALSE`). `quiet` suppresses console output. Additional plotting parameters can also be passed to adjust margins, axis label rotation, etc. See [here](). 

## S3 generic methods

As well as `plot`, saved `calc_rate.int` objects work with the generic S3 methods `print`, `summary`, and `mean`

### `print`

This simply prints a single replicate rate result to the console, by default the first one. The `pos` input can be used to print others. 

```{r}
print(zeb_all)
print(zeb_all, pos = 50)
```

### `summary`

This prints the summary table to the console which contains linear model coefficients and other metadata for each replicate rate. The `pos` input can be used to select which replicates (`$rank` column) to include. 

```{r}
summary(zeb_all)
summary(zeb_all, pos = 1:4)
```

`export = TRUE` can be used to export the summary table as a data frame, or those rows selected using `pos`

```{r results='hide'}
zeb_exp <- summary(zeb_all, 
                   pos = 1:4, 
                   export = TRUE)
```

```{r}
zeb_exp
```

### `mean`

This averages all replicate rates in the `$rate` column, or those selected using `pos`. The result can be saved as a value by using `export = TRUE`.

```{r}
zeb_mean <- mean(zeb_all, pos = 1:4, export = TRUE)
zeb_mean
```

This could be passed as a numeric value to later functions such as `convert_rate()`, although it is probably better to do this kind of result filtering after conversion. 

  
## Notes and tips

- Take note that `starts` and `ends` inputs are always in **row widths or row locations**, while the `from` and `to` inputs can be time, row, or oxygen values as determined by the `by` input.
