---
title: "calc_rate.int: Manual extraction of rates in intermittent-flow respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{calc_rate.int: Manual extraction of rates in intermittent-flow respirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F, warning = F, message = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 3))

```


## Introduction

`calc_rate.int()` is a function for manually determining rates across
user-defined ranges of row or time in each replicate in
intermittent-flow respirometry data. This page contains descriptions and simple
examples of the functionality. See also the help file
at `help("calc_rate.int")`, see `vignette("intermittent_short")` for how it can be used to analyse a relatively brief intermittent-flow respirometry experiment, and
`vignette("intermittent_long")` for an example of analysing a longer experiment.

## Overview

How `calc_rate.int` works is fairly straightforward. 
The function uses the `starts` locations to subset each replicate from the data in `x`. It extracts a rate using the `wait` and `measure` inputs in the `by` metric. 

A `calc_rate` object is saved for each replicate in the output in `$results`, and the output `$summary` table contains the results from all replicates in order with the `$rep` column indicating the replicate number. This output object can be passed to subsequent functions such as `adjust_rate()` and `convert_rate()` for further processing. 

## Inputs

There are two required inputs: 

- `x`: The time~oxygen data from a single
intermittent-flow experiment as either a `data.frame` or an `inspect` object

- `starts`: The location of the start of each replicate. This
can be either: 
  - A numeric vector of row numbers or times indicating each replicate start location.
  - A single numeric value representing a regular row or time interval starting from the first row or time value in the data. This option should only be used when replicates cycle at regular intervals. If the first replicate does not start at row 1 of the data in `x` it should be subset so that it does. See `subset_data()` and examples below for how to do this.  


### Other inputs 

- `wait`: Rows or time period to exclude at the start of each replicate.

- `measure`: Rows or time period over which to calculate rate in each replicate. Applied directly after `wait` phase. 

- `by`: Method by which `starts`, `wait` and `measure` are applied. Either `"row"` or `"time"`. 

- `plot`: Controls if a plot is produced. See [Plot](#plot) section. 

## Irregularly spaced replicates

### Data

The `urchin` dataset used for examples below is the included `intermittent.rd` example data with the time values changed to minutes to better demonstrate time region selection. Experimental data such as volume, weight, and row locations of replicates, flushes etc. can be found in the data help file: `?intermittent.rd`. 

```{r}
urchin <- intermittent.rd
urchin[[1]] <- round(urchin[[1]] / 60, 2) # change time values to minutes and round them
```

This is what the whole dataset look like. There are three replicates, and note they are of different duration.

```{r message=FALSE, warning=FALSE, results='hide'}
urchin <- inspect(urchin)
```

### Rate from `row` range in each replicate

If no inputs other than `x` and `starts` (in the default units of `by = "row"`) are entered, the default behaviour is to calculate a rate across all the data in each replicate. 

```{r tidy=FALSE, results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3901)) 
```

This is obviously not going to produce an appropriate rate. To exclude flush periods use the `measure` input. Here, using a vector of the same length as `starts` we can specify a different `measure` phase in each replicate. The default is to specify this in row widths, that is `by = "row"`. 

```{r tidy=FALSE, results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3901),
              measure = c(1800, 1200, 800)) 
```

However, we usually want to use the same region within each replicate to get a rate, and also exclude the first few minutes to allow a period of settling or acclimation after the flush. We can enter the `wait` and `measure` inputs as single values to acheive this.

```{r tidy=FALSE, results='hide'}
calc_rate.int(urchin,
              starts = c(1, 2101, 3901),
              wait = 180,
              measure = 600) 
```

### Rate from `time` range in each replicate

`starts`, `wait`, and `measure` can also be specified in the time units of the original data by using `by = "time"`.

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(0,35,64),
              wait = 3,
              measure = 10,
              by = "time")
```

The function uses the closest matching values if the exact values do not occur in the time data. 


## Regularly spaced replicates

For intermittent-flow experiments where the replicates are regularly spaced you do not have to specify the start location of every single one in `starts`. You can instead enter a value specifying the interval between each replicate starting at the first row of the input data in either rows or time units, as specified via the `by` input.  

### Specify regular replicate structure

The `zeb_intermittent.rd` data is from an intermittent-flow experiment on a zebrafish. It has a regular replicate cycle of 11 minutes (660 rows) for most of its length, comprising 9 minutes of measuring and 2 minutes flush. See `help("zeb_intermittent.rd")` for exact locations, but essentially after a background control and a single replicate of 14 minutes this cycle is maintained until the background recording at the end. 

Here we use `subset_data()` to extract only the regularly spaced replicates, and pipe the result to `calc_rate.int`. We specify the 660 row cycle using `starts`, exclude the first 2 minutes of each replicate using `wait`, and extract a rate from the following 6 minutes using `measure`, leaving the flush excluded. 

```{r results='hide', fig.keep='last', message=FALSE}
zeb_all <- zeb_intermittent.rd |>
  
  # inspect the data
  inspect() |>
  
  # subset regular replicates from larger dataset
  subset_data(from = 5840,
              to = 75139,
              by = "row",
              quiet = TRUE) |>
  
  # calc rate in each one from row 100 to 500, plotting first 3 and last 3
  calc_rate.int(
    starts = 660,
    wait = 120,
    measure = 360, 
    by = "row",
    plot = TRUE,
    pos = c(1:3, 103:105))

```

We use the `pos` input which is passed to `plot` to select the first and last three replicates for plotting to check everything looks okay. In a full analysis however we **strongly** recommend every replicate rate result be examined visually. Up to 20 results at a time can be plotted using the `pos` input. See [Plot](#plot) section below for further details. 

### View summary

Each replicate result is saved in the `$summary` element of the output, or we can use `summary()`. 

```{r}
summary(zeb_all)
```

The replicate number is indicated by the `$rep` column. The `$rank` column indicates ranking or ordering of rates *within* each individual replicate. Since `calc_rate.int` only returns one rate per replicate it is not relevant here, but is useful for `auto_rate.int()` results where multiple rates per replicate may be returned. 


## Plot {#plot}

If `plot = TRUE`, a plot is produced of each replicate rate result on a grid up to a total of 20. By default this is the first 20 (i.e. `pos = 1:20`). Others can be selected by modifying the `pos` input, either in the main function call or when calling `plot()` on output objects. 

```{r results='hide'}
plot(zeb_all, pos = 50:69)
```

For all plots, the bottom blue time axis shows the time values of the original larger dataset, whilst the upper red row axis shows the rows of the replicate subset. 

There are three ways in which these `calc_rate.int` results can be plotted, selected using the `type` input in either the main function call or when calling `plot()` on output objects.

### `type = "rep"`

The default is `type = "rep"` in which each replicate is plotted individually with the rate region highlighted. 

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(0,35,64),
              wait = 5,
              measure = 15,
              by = "time")
```

### `type = "full"`

Entering `type = "full"` will show each replicate rate highlighted in the context of the entire dataset.

```{r results='hide'}
calc_rate.int(urchin,
              starts = c(0,35,64),
              wait = 5,
              measure = 15,
              by = "time",
              type = "full")
```

Note this may be of limited use when the dataset is large. 

```{r results='hide'}
plot(zeb_all, type = "full", pos = 50)
```

### `type = "cr"`

Lastly, the `calc_rate` result for each individual replicate (as found in the `$results` element of the output) can be plotted using `type = "cr"` and the `pos` input. Note these are results from subsets of the initial data. 

```{r results='hide'}
plot(zeb_all, type = "cr", pos = 50)
```

The `pos` input here can also be of multiple replicates but this will produce multiple individual plots that you will need to scroll through. 

### Additional plotting options

`legend` can be used to show axis labels (default is `FALSE`). `quiet` suppresses console output. Additional plotting parameters can also be passed to adjust margins, axis label rotation, etc. See [here](https://januarharianto.github.io/respR/articles/inspecting.html#additional-plotting-options). 

## S3 generic methods

As well as `plot`, saved `calc_rate.int` objects work with the generic S3 methods `print`, `summary`, and `mean`

### `print`

This simply prints a single replicate rate result to the console, by default the first one. The `pos` input can be used to print others. 

```{r}
print(zeb_all)
print(zeb_all, pos = 50)
```

### `summary`

This prints the summary table to the console which contains linear model coefficients and other metadata for each replicate rate. The `pos` input can be used to select which replicates (`$rep` column) to include. 

```{r}
summary(zeb_all)
summary(zeb_all, pos = 1:4)
```

`export = TRUE` can be used to export the summary table as a data frame, or those rows selected using `pos`

```{r results='hide'}
zeb_exp <- summary(zeb_all, 
                   pos = 1:4, 
                   export = TRUE)
```

```{r}
zeb_exp
```

### `mean`

This averages all replicate rates in the `$rate` column, or those selected using `pos`. The result can be saved as a value by using `export = TRUE`.

```{r}
zeb_mean <- mean(zeb_all, pos = 1:4, export = TRUE)
zeb_mean
```

This could be passed as a numeric value to later functions such as `convert_rate()`, although it is usually better to do this kind of result filtering after conversion. 

  
## Notes and tips

- Internally `calc_rate.int` uses `calc_rate()` to extract rates, which explains why each replicate result in `$results` is a `calc_rate` object. The `wait` and `measure` inputs are simply converted to the `from` and `to` inputs of `calc_rate`. Essentially `from` is equivalent to `wait`, and `to` equivalent to `wait + measure`. If you want to use the `by = "oxygen"` method of `calc_rate` you will have to look into looping the function. See [here](https://januarharianto.github.io/respR/articles/archive/intermittent_old.html).

- Note the default behaviour of `calc_rate` if values are outside the range present within each replicate is used. For example if the entered `measure` phase is longer than the actual replicate, the last row or time value in the replicate will be used instead. Similarly, if `wait` or `measure` inputs do not match exactly to a value within the data the closest matching value is used instead. See `calc_rate()` for full details.

- The examples above assume flushes occur at the end of each replicate. If your data is subset or processed so that flushes are at the start instead this is easy to deal with. Simply use `wait` to exclude the flush at the start plus whatever subsequent wait phase you want to use before the `measure` phase. See example [here](https://januarharianto.github.io/respR/articles/adjust_rate.html#example-2---intermittent-flow-respirometry).


