---
title: "calc_rate: Manually select and calculate rates"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{calc_rate: Manually select and calculate rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F, warning = F, message = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                                scipen = 999, 
                                digits = 3))

sard <- sardine.rd[,1:2]
sard[[1]] <- sard[[1]]/60
```


## Introduction

`calc_rate()` is the general function in `respR` for manually determining rates across user-defined ranges of time, row or oxygen value. This contrasts with `auto_rate()` which uses a degree of automation to determine rates in the `"linear"` method, or allows rolling rates of a fixed width to be calculated across an entire dataset and then ordered in various ways. 

## Defaults

By default, if no other inputs are entered `calc_rate` will calculate a rate across an entire dataset as entered. The data used here is the `sardine.rd` dataset, but the time data has been converted to minutes. 

```{r}
cr <- calc_rate(sard)
```


## Time range

The default method by which a rate region can be specified is `by = "time"`. The function finds the closest matching values if these exact values do not occur in the time data. If either of the values lie outside the actual time data range it will use the first or last value instead. 

```{r}
cr <- calc_rate(sard,
                  from = 20, 
                  to = 80,
                  by = "time")
```

## Row range

Similarly the rate region can be specified `by = "row"`.

```{r}
cr <- calc_rate(sard,
                  from = 2000, 
                  to = 6000,
                  by = "row")
```

## Oxygen range

Lastly, it can be specified `by = "oxygen"`. This option finds the first occurrence of the `from` value (or the closest matching value), and the last occurrence of the `to` value. 

```{r}
cr <- calc_rate(sard,
                  from = 94, 
                  to = 92,
                  by = "oxygen")
```


## Multiple rates

`calc_rate` can be used to extract multiple rates by entering vectors of paired values as `from` and `to` in any of these metrics.

```{r}
cr <- calc_rate(sard,
                  from = c(10, 20, 30, 40, 50, 60, 70), 
                  to = c(20, 30, 40, 50, 60, 70, 80),
                  by = "time")

summary(cr)
```


## S3 generic methods

Saved `calc_rate` objects work with the generic S3 methods `print`, `summary`, `plot`, and `mean`

### `print`

This simply prints the result to the console. If there are multiple rates it will print the first one. The `pos` input can be used to print others. 

```{r}
print(cr)
print(cr, pos = 2)
```

### `summary`

This prints the summary table to the console which contains linear model coefficients and other metadata for each rate. If there are multiple rates the `pos` input can be used to select which to print. The `export` input can be used to export the `pos` selected rows as a dataframe, or the entire table if this is left `NULL`.

```{r}
summary(cr)
summary(cr, pos = 1:4)
```

```{r results='hide'}
cr_exp <- summary(cr, pos = 1:4, export = TRUE)
```

```{r}
cr_exp
```

### `mean`

This averages all the values in the `$rate` column, or those selected using `pos`. The result can be saved as a value by using `export = TRUE`.

```{r}
cr_mean <- mean(cr, pos = 1:4, export = TRUE)
cr_mean
```

### `plot`

`calc_rate` objects can be plotted. `pos` can be used to select which result to plot, and `panel` to plot individual panels. `legend` can be used to suppress labels and legend, and `quiet`to suppress console output. Additional plotting parameters can also be passed to adjust margins, axis label rotation, etc. 

Here we plot result 2, panel 2, hide the equation legend, rotate the axis labels, give the left axis more space, and increase the space from axis labels to axis ticks.

```{r}
plot(cr, pos = 2, panel = 2, legend = FALSE, quiet = TRUE, 
     las = 1, mai = c(0.3, 0.4, 0.35, 0.15), mgp = c(0, 0.5, 0))
```


## Two-point rate

The output also includes a `$rate.2pt`. This is the rate determined by simple two-point calculation of difference in oxygen divided by difference in time. See `vignette("twopoint")` for an explanation of this output and when it might be useful.

