---
title: respR in use by the scientific community
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{respR in use by the scientific community}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references_citations.bib
csl: resprcsl.csl
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = F, warning=FALSE, message=FALSE}
# This is ridiculously over the top to get this number, but good to learn :D
# Quite slow tho...
# Dump GScholar page
dump <- suppressWarnings(readLines("https://scholar.google.co.uk/scholar?cites=15799904892345725732&as_sdt=2005&sciodt=0,5&hl=en"))
# Find location of n citations line
loc <- grep("About [0-9]+ results \\(", dump)
# Extract it
strng <- stringr::str_extract(dump[loc], "About [0-9]+ results \\(")
# Extract the number
ncites <- as.numeric(stringr::str_extract_all(strng, "\\d+"))
```

In the `r xfun::numbers_to_words(lubridate::year(Sys.Date())-2019)` years since publication, `respR` has picked up [**`r ncites` citations**](https://scholar.google.co.uk/scholar?cites=15799904892345725732&as_sdt=2005&sciodt=0,5&hl=en).

We are very happy to see the package being used to conduct quality, reproducible science, and excited to see what experiments and data users will put it towards in the future. It's been especially encouraging to see it used for calculating the respiration rates of a wide range of species, as well as oxygen production rates of algae and corals.

If you use the package please [**cite**](https://januarharianto.github.io/respR/authors.html#citation) the publication in [**Methods in Ecology and Evolution**](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13162). If you use it for any other purpose (e.g. teaching, industry use, environmental reports and monitoring, etc.) [**we'd love to hear about it**](https://januarharianto.github.io/respR/articles/contact.html).

## Studies using `respR`

The following publications (most recent first) have used and cited `respR`. If we have missed any [**let us know**](https://januarharianto.github.io/respR/articles/contact.html).

#### @gomesImpactsAcuteHypoxia2023

Preprint

#### @zhangLossBehavioralStress2023

Cavefish - uses plot

#### @quigleyPhysiologicalCharacterizationCoral2023

mine

#### @holmes-hackerdNaupliarExposureAcute2023a

tmp

#### @luceyClimateWarmingErodes2023

tmp

#### @hawadleEffectsMultipleEnvironmental2023

Student research

#### @lowesCopperExposureDoes2023

tmp

#### @castrillon-cifuentesSpatiotemporalVariabilityOxygen2023

This study looked at the effects of hypoxic conditions on corals from the eastern Pacific and used `respR` to get MO2 and PCrit. What's really interesting about this study is that it also determined the same metrics using the [`respirometry`](https://cran.r-project.org/web/packages/respirometry/index.html) package, and we were happy to see that results between the two packages are largely the same. It is also good to see the authors varying the width used in the PCrit analysis to see how it affects the results. The width over which a rolling rate is determined from the oxygen timeseries is a really important parameter that is frequently not reported for PCrit results, so good to see the authors taking account of it. Of course `respR` makes this easy!

#### @hawkePetroleumderivedBiopolymerMicroplastic2022

tmp

#### @somoIntegrativeAnalysisRespiratory2022

Derek Somo's PhD project looked at a wide range of physiological functioning in an intertidal fish under warming and hypoxia. He used the `oxycrit` function to get the PCrit of his intertidal sculpins. Congrats on the PhD! 🥳

#### @thambithuraiFishVulnerabilityCapture2022

Excellent work from Davide Thambithurai and collaborators. They found parasite density affected capture probability in zebrafish, with implications for vulnerability to capture in commercial fisheries. As for metabolic rate, they showed parasites increased SMR, but not MMR or AAS. Davide was one of our early beta testers. In fact the `zeb_intermittent.rd` example data comes from this very study!

#### @kuttiFishFarmEffluents2022

This is an interesting study looking at the effects of effluents from fish farms on cold water corals in Norway. It used the `auto_rate` function to get the respiration rates of coral fragments to establish the negative effects of effluent on metabolic rates, and also showed it has negative effects on energy stores and growth. 

#### @luceyCompoundExtremeEvents2022

Another awesome sea urchin respirometry study! This looked at the combined effects of heatwaves and deoxygenation events on a tropical sea urchin, and used `respR` to determine both metabolic rates and $P_{crit}$ under these stressors. The authors show that the compound effects of marine heatwaves and hypoxic events, which usually occur together, are devastating to the populations of this species. 

#### @guitardIncreasedParasiteLoad2022a

Joëlle Guitard and colleagues examined how parasite load affected metabolic rates and escape responses in a freshwater fish. They discovered that parasites reduced metabolic rates and also adversely affected escape responsiveness. They used `respR` to calculate MO2 and the `auto_rate` function to do 60 second rolling regressions from which MMR were calculated. Really nice write-up of the methods. Lots of detail, just what we like to see! 👌

#### @grigorSwimmingActivityIndicator2022

Jordan Grigor and collaborators used `respR` to compare respiration and swimming activity from epi- and mesopelagic copepods and how they may be indicators of the onset of diapause. They found differences in size, activity and metabolism between epi- and mesopelagic groups. Nice work, and another example of how the package is being used to analyse micro-respirometry data.

#### @klementievDevelopmentVersatileLowCost2022

We've always noted in our documentation that except for the final conversion step, `respR` treats data as unitless, and it could be used to identify and extract rates from other types of experimental data, and here is the proof. It was used in this study, specifically `auto_rate`,  to quantify $H_2O_2$ consumption rates as part of development of a system to study biofilm activity. Really great to see the package being used like this. 

#### @paulaAccessCleaningServices2022

José Ricardo Paula *et al*. looked at the interactions between  ocean acidification, access to cleaner species, and parasite infection and how they affect physiology (MMR, RMR and AS) in damselfish on beautiful Lizard Island on the GBR. Looks like increased parasite tolerance comes at the cost of responsiveness to other stressors such as OA. Nice work. Thanks for the citation!

#### @joyceRegulationHeartRate2022

This study by William Joyce *et al*. looked at heart rates in larval zebrafish and conducted respirometry trials to determine if altered heart rates were associated with changes in oxygen consumption or $P_{crit}$ (spoiler: they weren't). Great to see `respR` put to use on microrespirometry data on such small specimens - 0.16 mg 😱. 

#### @schusterDistinctRealizedPhysiologies2022

More sea urchin respirometry in this study looking at differences in physiology between green sea urchin populations from urchin barrens and kelp forests. `respR` was used to determine the metabolic rates of the urchins, and the authors found those from barrens to have lower oxygen consumption rates and also that these populations were more sensitive to temperature changes. 

#### @morganReducedPhysiologicalPlasticity2022

A really cool and extensive study by Rachael Morgan *et al*. Zebrafish were acclimated to 15 (🤯) temperatures and then a host of traits measured, including physiology, gene expression, and behaviour. They found that physiological plasticity was lost from fish adapted to stable temperatures. Incredible amount of work, and really clear results with important implications for anyone doing experimental work on animals in different temperatures. `respR` was used to get SMR, MMR, and aerobic scope. 

#### @rochePathsGreaterConsensus2022

Dom Roche *et al*. review the status of open science and reproducibility in experimental biology and their role in consensus building and cross-discipline collaborations. Very interesting paper and essential reading for anyone doing collaborative experimental biology and sharing their results. They gave `respR` a quick shout-out as an example of open-source software aiding cross-disciplinary collaboration. Thanks for the citation! 

#### @garnerCanSeagrassModify2022

This study on Sydney rock oysters and the buffering effects of seagrasses on seawater pH used `respR` to examine the metabolic rates of oysters under different pH. The authors showed seagrasses boosted oyster growth at ambient pH, but did not do so at elevated pH. So seagrasses may not be able to buffer out the negative effects of ocean acidification. Cool study.

#### @bergerRegionalVulnerabilityAssessment2022

Halle Berger used `respR` in this MSc thesis, looking at the response of the Dungeness crab to climate stressors in the context of a regional vulnerability assessment. They found that respiration rates of this species increased exponentially under higher temperatures. We hope the package proved useful in establishing this from your respirometry data. Congrats on the Masters! 🥳


#### @wuEndocrineDisruptionPlastic2022

In this study Nicholas Wu examined how an endocrine disruptor in the environment exacerbated the effects of high temperatures on growth and metabolism in zebrafish. He used `respR` to calculate the resting MO2 of zebrafish under warming and exposure to plastic pollution. Nice work Nic!


#### @prokkolaGeneticCouplingLifehistory2022

In this study Jenni Prokkola *et al*. examine the genetic basis of how energy metabolism and life history interact in Atlantic Salmon. This used the `respR` function `auto_rate()` to calculate rates across one and two-minute time windows to identify maximum metabolic rates. After seeing it as a preprint we're delighted to see it published in a top journal. Nice work Jenni!  

#### @burfordRapidRangeExpansion2022a

`respR` (and one of its developers) had a small role in this huge and extensive study by Ben Burford as part of his PhD. Ben used metabolic rate measurements to model the physiological sensitivity in terms of temperature and dissolved oxygen of a squid which occurs across a wide latitude range in the eastern Pacific, and how these may allow for rapid range expansions. Incredibly impressive, wide-ranging work.

#### @mallonLightDrivenDynamics2022

Jennifer Mallon is another researcher who took advantage of our open offer to help get users started with `respR` by sending us a data file. Great to see the work published! They looked at both oxygen uptake and production in corals and coralline algae to calculate net photosynthesis. Really great to again see oxygen production being calculated using the package, not just consumption. 

#### @burnsMetabolicRateCritical2021
 
Alexandra Burns used `respR` in this MSc project, looking at oxygen supply capacity under different temperatures in a species of shrimp. Congrats on the Masters! 🥳

#### @killenGuidelinesReportingMethods2021a

Excellent summary and guide to best practices in conducting, analysing and reporting intermittent-flow respirometry studies. Essential reading!

#### @prinzingAnalyticalMethodsMatter2021

An excellent summary and guide to the best practices and tools available to estimate maximum metabolic rates. Goes into a lot of detail about rolling regressions, and the importance of the sampling window used, something that was instrumental to the design of `respR`, and is relevant to all metabolic rate calculations not just MMR. Very much required reading if you are interested in respirometry analyses in general.

#### @rosEnvironmentalStressResponses2021

This study looked at how physiological plasticity in invasive species can potentially predict invasion success. `respR` was used to calculate the Pcrit of native and invasive shrimp. Really nice results showing the invader having a lower Pcrit at higher temperatures than the native, potentially allowing it to tolerate hypoxic events better.

#### @rodgersDoubleWhammyNitrate2021

This study examines the effects of nitrate pollution on thermal and hypoxia sensitivity in European grayling. `respR` was used to calculate the Pcrit. They show (quite definitively, using three different PCrit methods) that Pcrit significantly increases under nitrate exposure, indicating a lower hypoxia tolerance. Nice work!

#### @pilletPhysiologicalPerformanceCommon2021

This study looked at the performance of the common carp when exposed to heavy metals, and found metabolic rates increased after 3 days of exposure. `respR` was used to process the respirometry data and calculate both SMR and MMR. Marion Pillet was a researcher who took advantage of our open offer to help get users started with `respR` by sending us a data file, so we are happy to see the work published!

#### @petreikyteEfficiencyNileTilapia2021

This MSc project examined the effects of salinity on Nile tilapia growth and physiology in aquaculture recirculation tanks, and used `respR` to calculate the metabolic rates. It's mostly in Lithuanian, so we haven't been able to figure out the details, but congrats on the MSc!

#### @legrandReducedPhysiologicalPerformance2021

Another study we were happy to see examining both oxygen production rates as well as consumption rates. It used `respR` to determine gross primary production by determining both oxygen production and respiration rate in coralline algae. Take home message: lots of salmon poop is bad for algae.

#### @ladakisOverviewBiosignalAnalysis2021

This review on methods in stress assessment studies mentions `respR` as one of the analytical options for examining respiration rates.

#### @hariantoImpactsAcclimationWarmlow2021

Here, the package was used to calculate mass-specific metabolic rates of sea urchins from intermittent-flow respirometry data, to examine the carryover effects of acclimation to different temperatures and ocean acidification conditions. Can't say too much about this one, except it's good to see a package author still using his own tools. 😉

#### @evensenRemarkablyHighConsistent2021

This study is a great example of how respirometry is not all about respiration, and exactly what we were thinking of when we designed `respR` to handle oxygen production rates as well as consumption. Here, it was used to calculate oxygen flux (net photosynthesis and respiration) of Red Sea corals under simulated heatwaves in large mesocosms.

#### @durtscheThermalConditionsEmbryogenesis2021

This study used `respR` to calculate the aerobic scope of juvenile brown trout to investigate the effects of egg incubation temperature. We were very happy to see the `auto_rate()` function being used to extract both most linear and maximum metabolic rates to calculate the aerobic scope, one of the primary use cases we designed it to handle.

#### @bouyoucosInvestigatingLinksThermal2021

This study looked at thermal tolerance and oxygen supply capacity in newborn sharks. This shows a great use for `respR` outside of the lab, in that it was used to determine oxygen decline from field recordings of oxygen off the coast of Moorea. We were especially pleased to see the `auto_rate()` function get a mention in the manuscript for calculating maximum metabolic rates.

#### @zhangMeasuringMaximumOxygen2020

This study compared the results of different methods of determining maximum metabolic rates. The authors point out minimum sampling window is a very important factor in determining MMR. We always thought this was pretty obvious, which is why `respR` has always allowed rolling regressions of different window sizes to be performed, with visualisations to see the results of how this affects rate calculations. 🤔

#### @zavorkaReducedExplorationCapacity2020

This is a really cool study looking at how warming may affect brain size and cognition in minnows, as well as metabolic rate. Warm fish had bigger brains, but weren't as good at finding their way through a maze! Awesome stuff. It used `respR` to determine SMR, MMR and aerobic scope.

#### @spindelZombiesNearshoreMetabolic2020

We have a special fondness for studies doing respirometry on sea urchins (we feel your pain, especially the spines under the fingernails). This study compared the physiology of urchins from kelp barrens and kelp forests, and found those in barrens have dramatically reduced resting metabolic rates, by up to 40%. Cool stuff!

#### @rodriguezRoleTemperatureOxygen2020

This is another study that used `respR` to determine the Pcrit, this time of an invasive bivalve, and the authors used it to show that the lack of metabolic regulation may limit the spread of this invader.

#### @mullerEffectsExperimentalOcean2020

This study looked at the effects of ocean acidification on a larval fish. The authors used `respR` to calculate SMR and MMR rates, then filtered the rates according to various criteria (above 80% oxygen, r2 above 0.85, etc.). This is a great idea, and exactly why we introduce the function `select_rate()` in v2.0, which allows for exactly this kind of application of easily reported rate selection criteria.

#### @kurchabaEffectsMPPolyethylene2020

Amazingly, this is the only study on zebrafish to cite `respR` so far, which is surprising given how many groups are working and doing respirometry on them. This study looked at the effects of microplastics on oxygen consumption and gene expression in larval zebrafish. They did not find any effects on metabolic rate, but did on the fish microbiome, which could lead to higher susceptibility to disease.

#### @internicolaMicroplasticsDelawareBay2020

Anna M. Internicola used `respR` in this MSc project, looking at the effects of microplastics on zooplankton in Delaware Bay. She found, amongst other effects, chronic exposure increased metabolic rates. Nice work, and congrats on the Masters! 🥳

#### @gomezisazaAnthropogenicDisturbancesFreshwater2020; @gomezisazaThermalAcclimationOffsets2020

Daniel F. Gomez Isaza used the `calc_rate()` function to calculate mass-specific metabolic rates of juvenile silver perch to examine the effects of nitrates on thermal sensitivity in Chapter 5 of his PhD thesis and accompanying JEB paper. Thanks for the citations, and congrats on the PhD! 🥳

#### @gilbertThermalLimitsCardiorespiratory2020

Matthew Gilbert cited `respR` in his PhD thesis as an example of how analytical tools in physiology are improving and evolving. Thanks for the citation, and congrats on the PhD! 🥳

#### @negreteMethodologicalEvaluationDetermination2019a

Another study that used `respR` to determine the Pcrit, this time of an estuarine fish. This was an examination of methods of determining Pcrit, both the analytical method of processing the metabolic rate data, and the respirometry method used (closed vs. intermittent-flow). There has been concern that closed respirometry mis-estimates Pcrit, because of the build-up of toxic metabolic end products, and intermittent-flow methods are preferable. The authors showed here this is not the case, and results were the same with either method. Really interesting stuff, and essential reading if you are wanting to determine Pcrit via respirometry.

#### @harveyDiatomsDominateAlter2019

This study examined how diatoms dominate food webs under ocean acidification conditions. Another study that uses `respR` to examine both oxygen production as well as consumption rates. The authors used the package to determine net oxygen production by determining both production and respiration.

#### @burfordGroupingReducesMetabolic2019a

This paper looked for the first time at the effects of grouping on metabolic rate in a social squid. The package was used here to calculate metabolic rates of both individual and groups of squid in a swim tunnel, and also their Pcrit. The very first paper to cite `respR` - a co-author may have had some influence on that... 🤔

## Full references
