---
title: "Closed-chamber respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4

vignette: >
  %\VignetteIndexEntry{Closed-chamber respirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(scipen = 999, 
                                digits = 3))
```


## Introduction

Here we describe a typical workflow for a closed-chamber respirometry experiment. Whether or not this is the type of experiment you are conducting, this is a good starting point to understand the general `respR` workflow, functions and how they work together.

The example data used here is `urchins.rd`, where the first column of the data frame is time data in *minutes*, while the remaining 18 columns are dissolved oxygen data in *mg/L*. Columns 18 and 19 (`b1` and `b2`) contain background recordings (i.e. from empty or "blank" control chambers). 

```{r}
head(urchins.rd)
```

## A typical respR workflow

The typical workflow in `respR` is to process a data frame containing paired values of numeric time and oxygen through several functions, saving the output object each time and inputting it into the next function. Much of this is optional; most `respR` functions also accept numeric values, vectors and `data.frame` objects, depending on the input or function. However, the *[object-oriented](https://en.wikipedia.org/wiki/Object-oriented_programming)* approach allows for several benefits, such as data integrity checks and reducing the need for additional inputs, and we would strongly recommend using it whenever possible. 

A typical workflow involves some or all of the following functions:

|  |  |
| -- | -- |
| `import_file()` & `format_time()` | Import data and prepare it for use in `respR` |
| `inspect()` | Visualise the data and check it for common issues|
| `calc_rate()` & `auto_rate()` | Extract rates from the entire dataset or regions of it, manually or automatically |
| `adjust_rate()` | Adjust the rate values for background oxygen consumption or production |
| `convert_rate()` | Convert the final adjusted rate to any common units of oxygen consumption or production |


### Importing and preparing data

`respR` has a very simple structural data requirement: data must be in the form of paired values of numeric time and oxygen amount. Most if not all oxygen sensing equipment or probe systems output to this format, or at least to a format that is easy to coerce to this structure. 

See `vignette("importing")` for an guide to importing and preparing data to this form.

### `inspect` - visualise and check for errors

Once data is in the form of a `data.frame` of paired time and oxygen values, we use `inspect()` to visualise the data and to check for common issues:

- Time and oxygen columns are *numeric*

- Time and oxygen data contain *infinite* (`Inf`) values

- Time and oxygen have *missing* (`NA`/`NaN`) values

- Time data are *sequential*

- Time data contains *duplicates*

- Time data are numerically *evenly-spaced* 

See `vignette("inspecting")` for a full description and examples of what these checks entail, the implications of them failing or producing warnings, and other inputs in the function. 

#### Inspect entire dataset

By default, `inspect` assumes the first column of the data frame is `time`, while the second column is `oxygen`. However, we can use the `time` and `oxygen` inputs to select different columns, including multiple columns. Here, we inspect all columns without `<-` assigning (i.e. saving) the result.

```{r results='hold', fig.height = 6}
inspect(urchins.rd, time = 1, oxygen = 2:19)
```

This is chiefly exploratory functionality to allow for a quick overview of a dataset. Here the plot lets us see which are the specimen columns, which might have data anomalies, and which are controls. 

The data checks tell us all the columns pass the various checks, except one. There is a warning that the time data are not evenly spaced. This is a common warning, and in this case can be safely ignored. It results from using decimalised minutes as the time metric, which happen to be numerically unevenly spaced, but are perfectly usable as the time metric in `respR`.

Rather than make assumptions that rows represent evenly spaced datapoints, the functions in `respR` use actual time values for analyses and rate calculations, and so even irregularly spaced data are analysed correctly. Such warnings are for informative purposes: to make the user aware of unusual data gaps, and also to remind users that if they use row numbers for manual operations such as subsetting, the same row width in different parts of the data may not necessarily represent the same time period. 

#### Inspect individual columns

To extract rates, it is best to assign each time-oxygen column pair individually as a separate `inspect` object. Using the `time` and `oxygen` inputs we can select particular columns.

```{r results='hold', warning=FALSE, message=FALSE}
urchin <- inspect(urchins.rd, time =  1, oxygen = 15)
```

Note how the data is plotted against both time (bottom blue axis) and row index (top red axis). From this plot, we can see irregularities in these data near the end of the timeseries (in this case the specimen had interfered with the oxygen sensor). A linear regression across the entire data series would therefore not be a representative estimate of the true routine respiration rate. However, the bottom plot, a rolling rate across a moving window of 10% of the data, shows that over the initial stages of the experiment oxygen uptake rate is consistent at around -0.02. In this experiment this region would be therefore be suitable for extracting rates, and we would not want to use any data after approximately the 27 minutes timepoint. 

For now, the data is saved as an object, `urchin` which contains the original data columns we selected coerced into a new data frame, and various other metadata. 

Again, note that using `inspect()` is optional. Functions in `respR` will generally accept regular `R` data structures (e.g. data frames, tibbles, vectors, etc.). `inspect()` is a quality control and exploratory step that helps highlight potential issues about the data before analysis. We use this particular example with an obvious anomaly to illustrate the point that you should *always visualise and explore your data before analysis*. `respR` has been designed to make this straightforward. 

### `calc_rate` - calculate oxygen uptake rate

#### Defaults

Using the `inspect` object `urchin` that we just created in `calc_rate()` with no additional inputs, will prompt the function to perform a linear regression on the entire data series. 

```{r fig.show='hold'}
calc_rate(urchin) 
```

Note how the function recognises the `inspect()` object, with no other inputs necessary. Alternatively, you can input a `data.frame` object containing raw data, in which case the function will automatically consider the first column as time data, and the second column as oxygen data (if they are not in the first two columns, they should be processed via `inspect()` or otherwise put into this structure).

#### Other options

In many cases, we want to select the region over which the rate is determined. For example, we may want to exclude initial stages of instability at the start of an experiment, determine the rate over an exact time duration, or within a threshold of oxygen concentrations. Equipment interference or other factors may cause irregularities in the data we want to exclude. `calc_rate` allows us to specify exact data regions and allow us to work around such issues.

Using the `from`, `to`, and `by` inputs, a user may use `calc_rate()` to specify data ranges in several ways:

- **Time:** 
    - *"What is the rate over a specific 25 minute period?"*
    - `from = 4, to = 29, by = "time"`
- **Oxygen:** 
    - *"What is the rate between oxygen concentrations of 7.5 to 7.0 mg/L?"*
    - `from = 7.5, to = 7.0, by = "oxygen"`
- **Row:** 
    - *"What is the rate between  rows 11 and 273."*
    - `from = 11, to = 273, by = "row"`

These inputs do not need to be overly precise; for `oxygen` and `time` if input values do not match exactly to a value in the data, the function will identify the closest matching values and use these for calculations. Similarly, if `from` or `to` values are beyond the maximum or minimum values in the data, the function will use the maximum or minimum value instead. For example, the above experiment is 45 minutes long. If we used `to = 60` as the upper time range, the function would simply apply a `to` value of 45 instead.

#### Calculate urchin rate

Here, to calculate our rate we'll select a 25 minute period before the interference occurred.

```{r results='hide', fig.keep='none'}
urch_rate <- calc_rate(urchin, from = 4, to = 29, by = "time")
```

Plotting the output provides a series of diagnostic plots of the data subset that was analysed. 

```{r fig.show='hold'}
plot(urch_rate)
```

The saved object can also be explored using generic `S3` R methods.
```{r}
print(urch_rate)

summary(urch_rate)
```

The rate, which at this stage is unitless, can be seen as the final column, and other summary data and model coefficients are saved in the object. In this case the `rsq` is 0.99, so this appears to be a very good estimate of this urchin's routine respiration rate. 

#### Two-point rate

The output also includes a `rate.2pt`. This is the rate determined by simple two-point calculation of difference in oxygen divided by difference in time. For almost all analyses, the `$rate` should be used. See `vignette("twopoint")` for an explanation of this output and when it might be useful.

### `calc_rate.bg` - calculate background oxygen

The presence of micro-organisms and their oxygen use may be a potential source of experimental bias, and we usually want to account for background respiration rates during experiments by conducting empty or "blank" control experiments with no specimens to quantify it. 

These control experiments are routinely conducted alongside, or before and after specimen experiments, or at an entirely different time, althoguh it is important they are done under the same conditions. Since the oxygen signal from these can be noisy and trends easily inflenced by outliers, often the background rates from several controls are averaged to obtain a more accurate estimate of the background rate. Once a background rate is determined, specimen rates are then adjusted by it. See `vignette("adjust_rate")` for detailed examples of the many types of adjustments that can be performed using `respR`.

The function `calc_rate.bg()` is used to extract background rates from control recordings. These must share the **same units of time and oxygen** as the experimental rates they will be used to adjust. It can also process multiple background oxygen recordings, so that an average background rate will be applied in the `adjust_rate` function, or other adjustments requiring multiple recordings can be performed. In the `urchins.rd` data, background respiration was recorded and saved in columns 18 and 19. 

We will determine background rates using `calc_rate.bg()`, which can calculate rates from multiple columns (unlike `calc_rate`), and save the output as an object. 

```{r results='hide', message=FALSE, warning=FALSE, fig.keep='last'}
bg_insp <- inspect(urchins.rd, time = 1, oxygen = 18:19)
bg_rate <- calc_rate.bg(bg_insp)
```

```{r echo = FALSE}
bg_rate
```

The `bg_rate` object contains both individual background rates for each data column entered (`$rate.bg`), and an averaged rate (`$rate.bg.mean`). We will determine how these are applied as an adjustment to the specimen rate in  `adjust_rate()`.

### `adjust_rate` - adjust for background respiration

See `vignette("adjust_rate")` for detailed examples of the types of adjustments that can be performed using `adjust_rate`. Here the adjustment is relatively straightforward. We will adjust the `urch_rate` using the `bg_rate` object we saved in the previous section. 

The rate input to be adjusted can be an object of class `calc_rate` or `auto_rate`, or any numeric value. The `by` adjustment value can be a `calc_rate.bg` object or numeric value or vector. `adjust_rate` has several methods determining how the `by` is applied, but the default one is `"mean"`, which is the one we want here, since we want to apply the average of the two background rates we just calculated. 

```{r message=FALSE, warning=FALSE, results='hide'}
urch_rate_adj <- adjust_rate(urch_rate, by = bg_rate, method = "mean")
```

```{r echo = FALSE}
urch_rate_adj
```

The urchin rate has been adjusted to a slightly lower uptake rate because, as we found by looking at the controls, some of this uptake was due to respiration by micro-organisms. 

Background adjustments can also be entered manually. Care should be taken to include the correct *sign*. In `respR` oxygen uptake rates are negative since they represent a negative slope of oxygen against time. Background rates are usually (though not always) also negative. In this case, the default `"mean"` method will not alter the `by` value. 

```{r message=FALSE, warning=FALSE, results='hide'}
urch_rate_adj <- adjust_rate(-0.0218, by = -0.000833)
```

```{r echo = FALSE}
urch_rate_adj
```

We can see this is essentially the same result as above. The small mismatch in values is simply due to the lower precision of entered values compared to internal ones. 

### `convert_rate` - convert the results

Note, that until this point `respR` has *not required units of time or oxygen to be specified*. Now we convert calculated, unitless rates to specified output units. 

`convert_rate()` can be used to convert the unitless rate values we have dealt with up to now to these reportable metrics:

- **Absolute metabolic rate** 
  - Oxygen consumption or production per unit time. This is the whole specimen, whole chamber or whole group metabolic rate
- **Mass-specific metabolic rate** 
  - Oxygen consumption or production per unit time per unit *mass* of the specimen
- **Area-specific metabolic rate** 
  - Oxygen consumption or production per unit time per unit *area* of the specimen

Conversion requires the units of the original raw data (`time.unit`, `oxy.unit`), and the `volume` of fluid in the respirometer in Litres (***L***). Mass-specific rates require the `mass` of the specimen in kilograms (***kg***), and area-specific rates require the `area` of the specimen in metres squared (***m^2***). Lastly, an `output.unit` appropriate to the inputs should be specified. 

*Note*: the `volume` is volume of fluid in the respirometer or respirometer loop, **not the volume of the respirometer**. That is, it represents the *effective volume*. A specimen occupies space in the respirometer, and so displaces some proportion of the water volume, which depending on its size might be significant. Therefore the volume of water entered here should equal the total volume of the respirometer *minus the volume of the specimen*. There are several approaches to determine the effective volume; calculating the specimen volume geometrically or via water displacement in a separate vessel, or calculated from the mass and density (e.g. for fish it is often assumed they have an equal density as water, that is ~1000 kg/m^3). Water volume could also be determined directly by pouring out the water at the end of the experiment, or by weighing the respirometer after the specimen has been removed. See the **[`respfun`](https://github.com/nicholascarey/respfun)** respirometry utilities package for several functions to assist with determining the effective volume. 

### Convert adjusted urchin rate

For an example of absolute oxygen uptake rate, we can convert the output of `adjust_rate()` to oxygen consumed in mg per hour:

```{r tidy = FALSE, eval = F}
convert_rate(urch_rate_adj,         # urchin rate adjusted for background
             oxy.unit = "mg/L",     # oxygen units of the original raw data
             time.unit = "min",     # time units of the original raw data
             output.unit = "mg/h",  # output unit
             volume = 1.09)         # effective volume of the respirometer
```

```{r tidy = FALSE, echo = F}
convert_rate(urch_rate_adj,         # urchin rate adjusted for background
             oxy.unit = "mg/L",     # oxygen units of the original raw data
             time.unit = "min",     # time units of the original raw data
             output.unit = "mg/h",  # output unit
             volume = 1.09)         # effective volume of the respirometer
```

We can also convert to a mass-specific rate by adding a specimen `mass` and specifying a mass-specific `output.unit`:

```{r tidy = FALSE, eval = F}
convert_rate(urch_rate_adj, 
             oxy.unit = "mg l-1", 
             time.unit = "m", 
             output.unit = "mg/s/kg",
             volume = 1.09, 
             mass = 0.19)
```

```{r tidy = FALSE, echo = F}
convert_rate(urch_rate_adj, 
             oxy.unit = "mgl-1", 
             time.unit = "m", 
             output.unit = "mg/s/kg",
             volume = 1.09, 
             mass = 0.19)
```

Note how we have changed the format of the time and oxygen units. A "fuzzy" string matching algorithm automatically recognises such variations, allowing natural, intuitive input. For example, `"ml/l"`, `"mL/L"`, "`ml L-1`", `"milliliter/liter"`, and `"millilitre/litre"` are all recognised as `ml/L`. Unit delimiters can be any combination of a space, dot (`.`), forward-slash (`/`), or the "per" unit (`-1`). Thus, `"ml/kg"`, `"mL / kg"`, `"mL /kilogram"`, `"ml kg-1"` or `"ml.kg-1"` are equally recognised as `mL/kg`. To see what units are available to use in various functions, call `unit_args()` or `?unit_args()`. 

```{r}
unit_args()
```

Note that some units of oxygen require temperature, salinity and atmospheric pressure to perform the conversion. This time we will save the result as an object.

```{r tidy = FALSE, eval = T}
urch_rate_final <- convert_rate(urch_rate_adj, 
                                oxy.unit = "mg/L", 
                                time.unit = "mins", 
                                output.unit = "ml/h/kg",
                                volume = 1.09, 
                                mass = 0.19,
                                t = 20,
                                S = 30,
                                P = 1.01)
```

```{r tidy = FALSE}
print(urch_rate_final)

summary(urch_rate_final)
```

### Final rate

The final rate can be seen above in these console outputs. It can be extracted for further use from the saved object where it is `$rate.output`.

```{r}
urch_rate_final$rate.output
```

### Check the result

A final and important step in respirometry analyses, particularly after the first rate from a new set of experiments or specimens has been calculated, is to *check the result* against rates from the same or similar species of the same size in as close as possible to the same conditions. This could be from previous analyses by the same user or group, or rates from one or more published studies. 

Dealing with so many different units and conversions it is incredibly easy to introduce a small conversion error that will have a massive effect on the final calculated rate. Some of these errors even make it into published papers. This is why we would also recommend you check the result against more than one independent publication. 

In the present example, at the time the experiments were done there were no published respirometry studies on this particular species of sea urchin. However, a quick Google Scholar search led to other papers from which we can get rates of other species of sea urchin at around the same size and temperature. In Moulin et al. ([2015](vignette("refs"))). *E. mathaei* has a respiration rate of around *0.2 to 0.5 umol/h/g* at 25°C, and in Kurihara et al. ([2013](vignette("References"))) *H. pulcherrimus* it is around *0.6 umol/h/g* at 22°C. Let's check what our urchins rate is in these same units:

```{r tidy = FALSE, eval = F}
convert_rate(urch_rate_adj, 
             oxy.unit = "mg/L", 
             time.unit = "mins", 
             output.unit = "umol/h/g",
             volume = 1.09, 
             mass = 0.19)
```

```{r tidy = FALSE, echo = F}
convert_rate(urch_rate_adj, 
             oxy.unit = "mg/L", 
             time.unit = "mins", 
             output.unit = "umol/h/g",
             volume = 1.09, 
             mass = 0.19)
```

We can see our rate is consistent with these others from the literature, and any difference is easily within what we might expect given variation between species and a slightly different temperature. Generally speaking, if there is a order of magnitude difference or more it is a good indication that there may be a problem. Therefore, here we can be fairly confident we have not made a major conversion error, and we have found that our urchin rate is in the right ballpark.

## Summary

This is an example of a fairly straightforward analysis of a closed-chamber respirometry experiment. This entire analysis can be documented and shared in only a few lines of code, making it easily reproducible if the original data file is included:

```{r results="hide", fig.keep="none", eval = FALSE, tidy = FALSE}
# import and inspect
urchin <- inspect(urchins.rd, time = 1, oxygen = 15)

# Background
bg_insp <- inspect(urchins.rd, time = 1, oxygen = 18:19)
bg_rate <- calc_rate.bg(bg_insp)

# Specimen rate
urch_rate <- calc_rate(urchin, from = 4, to = 29, by = "time")

# Adjust rate
urch_rate_adj <- adjust_rate(urch_rate, bg_rate)

# Convert to final rate units
urch_rate_final <- convert_rate(urch_rate_adj,
                                oxy.unit = "mgl-1", 
                                time.unit = "m", 
                                output.unit = "mg/s/kg", 
                                volume = 1.09, 
                                mass = 0.19)
```

Using pipes, either the new native `|>` pipes introduced in [R v4.1](https://www.r-bloggers.com/2021/05/new-features-in-r-4-1-0/) or `%>%`  [`dplyr`](https://dplyr.tidyverse.org) pipes, can condense this even further:

```{r results="hide", fig.keep="none", eval = FALSE, tidy = FALSE}
urchins.rd |>                                            # Using the urchins data,
  inspect(1, 15) |>                                      # inspect, then
  calc_rate(from = 4, to = 29, by = "time") |>           # calculate rate, then
  print() |>                                             # print for quick look,
  adjust_rate(
    calc_rate.bg(urchins.rd, time = 1,                   # calculate the background and
                 oxygen = 18:19)) |>                     # adjust rate, then
  print() |>                                             # print for another look,
  convert_rate(oxy.unit = "mgl-1", time.unit = "m",     
    output.unit = "mg/s/kg", volume = 1.09, mass = 0.19) # and finally convert
```

