---
title: "convert_DO: Convert between units of dissolved oxygen"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{convert_DO: Convert between units of dissolved oxygen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 6, 
               fig.height = 4,
               fig.align = "center")
```

# Introduction

Different oxygen probe systems output oxygen values in many different units of concentration or partial pressure, or even in a proportional unit such as relative percentage of dissolved air or oxygen. For example, it is very common to use water that has been 100% saturated with air through vigorous bubbling, mixing or stirring as the starting value of 100%, with percent saturation recorded as it declines, the assumption being this is the oxygen being removed by the specimen. 

The `convert_DO` function allows you to convert between any commonly used unit of oxygen concentration, partial pressure, or percent air or oxygen saturation.

Some unit conversions require temperature, salinity, and atmospheric pressure inputs in order to do conversions. The first two of these must be entered if they are needed. For the atmospheric pressure input (`P`), a default value of 1.013 bar (standard pressure at sea level) is applied if not entered. In most locations which have a normal range (outside of extreme weather
events) of around 20 millibars, any variability in pressure will have a
relatively minor effect on dissolved oxygen, and even less on calculated
rates. However, we would encourage users to enter the true value if they know
it, or use historical weather data to find out what it was on the day of the experiment.

Concentration units should be formatted as oxygen amount (volume or mass, in almost any unit) per unit volume or mass in SI units (L or kg) for the denominator. For example, `ml/kg` or `mg/L`. 

See `unit_args()` for the various units that are accepted.

## Example 1 - Saturation conversions

### Percent air saturation

The `sardine.rd` dataset has oxygen values in percent air saturation, and the accompanying help file, `help("sardine.rd")`, contains the temperature, salinity and atmospheric pressure inputs to allow it to be converted. 

```{r echo = T}
sardine.rd
```

We will convert it to `mmol/kg`. In `convert_DO`, the `t`, `S` and `P` inputs must be in °C, ppt (‰), and bar.  

```{r echo = T, tidy = F}
conv <- convert_DO(sardine.rd$Oxygen, # data to convert
                   from = "%Air",     # oxygen unit to convert from
                   to = "mmol/kg",    # oxygen unit to convert to
                   t = 15,            # in C
                   S = 35,            # in ppt
                   P = 1.013)         # in bar

print(conv)
```

Note, the output object is a `list` object of class `convert_DO` containing the input and output data and units. To use the converted values we can extract them from the `$output` element. For example, we can create a new data frame adding the column of the actual concentrations. 

```{r}
sardine_new <- cbind(sardine.rd,
                     mmol_kg = conv$output)

sardine_new
```

### Percent oxygen saturation

Note the difference between percent air saturation (`%Air`), where air saturated water is ~100%, and percent oxygen saturation (`%O2`), where air saturated water is ~20.946% *oxygen* saturated. In other words, ` %O2 = %Air * 0.20946 `. Some oxygen probe systems output in both units, or tend to use one over the other, so take care not to confuse them. 

```{r}
convert_DO(100, "%Air", "%O2",
           t = 15,
           S = 35,            
           P = 1.013)         
```

## Example 2 - Oxygen pressure units

`convert_DO` can also convert between units of oxygen partial pressure. We will add a column of the same oxygen values in hectopascals, converting from the `mmol/kg` column we just added, using `$output` to extract the values directly. 

```{r tidy = F}
sardine_new$hPa <- convert_DO(sardine_new$mmol_kg,  
                              from = "mmol per kg",    
                              to = "hPa",  
                              t = 15,            
                              S = 35,            
                              P = 1.013)$output     

sardine_new
```

Note, how the `from` unit is still recognised. Unit conversions in `respR` use a very forgiving, fuzzy algorithm to recognise different variations of units. 

## Example 3 - Converting single values

For quick conversions, `convert_DO` also accepts single values. These particular units here do not require temperature, salinity and pressure. 

```{r}
convert_DO(8, 
           "mg/L",
           "mmol/L")
```

## Example 4 - Enter additional inputs in correct units

`respR` has a convenient helper function to assist with `convert_DO` (and `convert_rate()`) where certain inputs such as temperature or atmospheric pressure have to be in specific units. 

`convert_val()` will convert between units of temperature, volume, mass, area, and atmospheric pressure. The utility of using it here, is that it will by default convert to the required units for the relevant `convert_DO` input. 

Here, we have been reading an old paper and want to convert an oxygen value to `mg/L`, and the temperature and atmospheric pressure values are in Fahrenheit and Torr. For the `t` and `P` inputs these must be in `°C` and `bar`. We could easily go to an online converter to do this, but `convert_val` can allow it within the `convert_DO` function call. 

```{r tidy = F}
convert_DO(4.6, 
           from = "ml/L",  
           to = "mg/L",
           t = convert_val(65, from = "F"), # must be in C
           S = 30, 
           P = convert_val(775, from = "Torr")) # must be in bar
```

Notice two convenient aspects:

 - We don't need to tell `convert_val` what the value is a measurement *of* - it is recognised automatically using the `from` unit. 
 
 - We don't need to specify a `to` unit (although you can). The defaults for these are set to the required input units of `t` and `P`, that is `°C` and `bar`. 

We can compare these to double check:

```{r}
convert_val(775, from = "Torr")
convert_val(775, from = "Torr", to = "bar")
```

See also `convert_rate()` where `convert_val` can be used in a similar way with the `volume`, `mass`, and `area` inputs.
