---
title: "Flowthrough respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{respR - Measuring intermittent-flow data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=7, fig.height=7)
```

### Introduction

Flowthrough respirometry differs from other methods in that oxygen depletion is measured from a known continuous flow of water through a respirometer, rather than a known volume, typically after the system has achieved equilibrium. In this method, two oxygen concentrations are needed; 'upstream' and 'downstream' of the experimental chamber, or inflow and outflow concentrations. The inflow concentration is typically a known fixed value and often not continuously monitored, such as fully air-saturated water, or water from a header tank of controlled oxygen concentration. The difference between inflow and outflow concentrations, the $O_2$ delta, and the flow rate allow calculation of the uptake by the specimen.

### Example data

The example data, `flowthrough`, contains measurements of oxygen consumption in a species of chiton, (*Mopalia lignosa*, Carey, *unpublished*).  Detailed information about the data, including its source and methods, can be obtained with the command `?flowthrough`. 

```{r}
library(respR)
data("flowthrough.rd")
head(flowthrough.rd)
```

We can see this dataset contains time values, both outflow and inflow $O_2$ concentrations, and an $O_2$ delta, which is simply the inflow substracted from the outflow. The $O_2$ delta column is not required; we include it here to show an exploratory step in the next `inspect_data` section. While any flowthrough data can be analysed in the `calc_rate.ft` function, inspecting the data requires a 3 column data frame of time, outflow $O_2$, and inflow $O_2$. Therefore the user should include all three, even if inflow $O_2$ is a column of a single fixed value. Here it varies, but not significantly from a mean value of 8.88. 

### Inspecting data
Continuous flowthrough data can be inspected for common errors using the `inspect_data` function, but because these data will be slightly different we need to use different input operators. By default, `inspect_data` assumes Time data in column 1 and $O_2$ data in column 2. For flowthrough data we typically want to look at the outflow O₂ data, which we have in column 2 here, so we can use the defaults:

```{r}
inspect_data(flowthrough.rd)
```

Unlike in  closed respirometry, in a flowthrough experiment oxygen should generally not decline over time, as long as equilibrium has been achieved, and so regions where rates are consistent are more easily identified. Here, using the top plot, we can see the outflow $O_2$ concentration is most consistent in the early stages of the experiment. Since we know in this experiment inflow $O_2$ is more-or-less unvarying, this means the specimen's rate is also most conistent here. In closed respirometry data, the bottom plot shows how the rate of the specimen varies over the dataset, but here it shows the rate of change in outflow $O_2$ concentration. Therefore, values close to zero indicate regions of highly consistent rates by the _specimen_. Here, we can see this illustrated well between rows 200 and 400.

If, for experimental reasons, the inflow $O_2$ concentration is variable, examining consistency in outflow $O_2$ would not necessarily equate to consistent rates by the specimen. Instead, we can examine $O_2$ delta data, by using other `inspect_data` operators to specify the column we are interested in. Here delta $O_2$ is in column 4:

```{r message=FALSE, warning=FALSE, results = "hide"}
inspect_data(flowthrough.rd, time = 1, oxygen = 4)

```

Here, we can see that after taking into account variable inflow $O_2$, the uptake rates of the specimen are most consistent between rows 200 and 300. Note however, this is an exploratory step: these $O_2$ delta data should not be saved via `adjust_rate`. Instead the original inflow and outflow $O_2$ data should be saved for passing to the next stage, through specifying the columns:

```{r results = "hide", fig.keep="none", results = "hide"} 
chiton.ft <- inspect_data(flowthrough.rd, time = 1, outflow.o2 = 2, inflow.o2 = 3)
```

```{r message=FALSE, warning=FALSE, fig.keep="none"}
chiton.ft$df
```

We can see only the specifed columns have been included, and the $O_2$ delta column removed. 

### Flowthrough rate function - `calc_rate.ft`
Rates can be detemined from flowthrough data using the `calc_rate.ft` function. This function accepts several forms of data input; 

1. Single values of inflow and outflow concentrations:

    ```calc_rate.ft(inflow.o2 = 8.88, outflow.o2 = 8,17, flowrate = 0.000039)```

2. Combinations of single or vector values. Vector inputs must be of equal length:

    ```calc_rate.ft(inflow.o2 = 8.88, outflow.o2 = c(8,17, 8.16, 8.15), flowrate = 0.000039)```  
    ```calc_rate.ft(inflow.o2 = c(8.88, 8.87, 8.89), outflow.o2 = c(8,17, 8.16, 8.15), flowrate = 0.000039)```   
    ```calc_rate.ft(inflow.o2 = 8.88, outflow.o2 = flowthrough.rd$o2.in, flowrate = 0.000039)```
    ```calc_rate.ft(inflow.o2 = flowthrough.rd$o2.in, outflow.o2 = flowthrough.rd$o2.out, flowrate = 0.000039)```

4. A data frame containing columns of paired outflow and inflow $O_2$ values, with column identifiers:

    ```calc_rate.ft(flowthrough.rd, time = 1, outflow.o2 = 2, inflow.o2 = 3, flowrate = 0.00039)```

5. An object of class `inspect_data`:

    ```calc_rate.ft(chiton.ft, flowrate = 0.00039)```  
    

The `flowrate` operator is required by the function, and it is assumed it will be an unvarying parameter within each separate experiment. Recordings at different flow rates should be analysed separately. Units for flow rate will be specified when rates are converted later (see below), but they should be in units of volume in L, per unit time (s,m,h), for example `L/s`.

We will calculate rates from the entire dataset using the `inspect_data` object we saved earlier. 

```{r}
chiton.ft.rate <- calc_rate.ft(chiton.ft,  flowrate = 0.000039)
print(chiton.ft.rate)
```

The output object includes all calculated rates over the dataset and a mean rate. 



### Subsetting data

The `calc_rate.ft` function currently does not support subsetting. The dedicated `subset_data` function can be used to prepare data (see Reference above or `?subset_data`), or we can use straightforward R subsetting syntax as follows. We saw earlier how rates between rows 200 and 300 appeared to be most consistent, so we will calculate rates from this data region using the object we saved earlier. 

```{r}
chiton.ft.rate.sub <- calc_rate.ft(chiton.ft$df[200:300,], time = 1, outflow.o2 = 2, inflow.o2 = 3, flowrate = 0.000039)
print(chiton.ft.rate.sub)
```
Because the function interprets this as a regular data frame, not `inspect_data` object, we need to specify the columns. Again, the output objects includes all calculated rates within the subset and a mean rate. Note the plot row index on the X-axis shows a row index of the subset, not the original row numbers. 

### Adjusting rates

Similar to the other respirometry methods, a background rate value (typically negative) can be used to adjust flowthrough rates. 
``` {r}
adjust_rate(chiton.ft.rate.sub$mean, by = -0.000001)
```
`respR` does not currently have a specialised background function for flowthrough data. However the main `calc_rate.ft` function can be used to examine data from 'blank' experiments for background rates, and these (or an average of several) can be entered as above. The user should however be aware that only background rates determined at the *same flow rate* should be used.

### Converting rates

The `convert_rate` function is used to convert rates to particular units. As with converting rates from closed and intermittent-flow respirometry (i.e. outputs of `calc_rate`), conversion of flowthrough rates (from `calc_rate.ft`) requires the `o2.unit` of the data, as well as a `volume.unit` and `time.unit`. 

However, inputs for `volume` or `time.unit` are not required since these are incorporated into the flow rate unit (e.g. L/s). The `flow.unit` must be in SI units of litres per unit time; L/s, L/m, or L/h. If an object of class `calc_rate.ft` is used, the function will prompt the user for the correct inputs. 

The function returns a rate for the whole specimen, and if a `mass` is entered (kg) it returns a mass-specific rate. 

```{r}
# obviously not yet working
#convert_rate(chiton.ft.rate$mean, o2.unit = 'mg/l', flow.unit = "l/s", output.unit = "mg/h")
#convert_rate(chiton.ft.rate$mean, o2.unit = 'mg/l', flow.unit = "l/s", mass = 0.000070, output.unit = "mg/h/g")
```

## so this funtion should then know from either a `flow.unit` being entered, or the object is class `calc_rate.ft` to do the correct conversion... I don't think it needs the actual flow rate since that's already been used to get the rate.....?????