---
title: "Flowthrough respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Flowthrough respirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                                scipen = 999, 
                                digits = 3))
```

## Introduction {#ftintro}

Flowthrough respirometry differs from other methods in that rates of oxygen change are measured from a controlled, continuous flow of water through a respirometer, rather than a known volume, typically after the system has achieved equilibrium. 
In this method, two oxygen concentrations are needed; 'downstream' and 'upstream' of the experimental chamber, or the outflow and inflow concentrations. 
These are used to calculate an oxygen delta (outflow minus inflow values), and this is used with the flow rate to calculate oxygen consumption or production rates.

An outflow recording or value, or already determined delta oxygen values are required to calculate the rate. 
The inflow concentration is typically a simultaneous recording, but an alternative is to use a known value from a source of constant oxygen concentration such as fully air-saturated water, or water from a header tank of controlled oxygen concentration, and so often is not continuously monitored. `respR` will accept any combination of available data to allow rates to be calculated.

`respR v2.0` contains an all new workflow to process flowthrough respirometry data, comprising four functions: 

- `inspect.ft`                 - check and visualise the data and calculate the oxygen delta (if you haven't already)
- `calc_rate.ft`            - calculate a rate from a region (or several) of the data 
- `adjust_rate.ft`      - adjust the rate for background activity (optional)
- `convert_rate.ft`   - convert the rate to specific units, including mass- or area-specific



## Exploring flowthrough respirometry data

Typically the first step in data analysis is inspecting and visualising the data. 

The `inspect.ft` function can be used to quickly examine and visualise outflow and inflow data from a single chamber, but also a large multi-column dataset. Using it with no arguments will inspect and plot every column, assuming `time` is in column 1 and all others are delta oxygen data (`delta.o2`). Even if your data is not delta oxygen values, you can use this without saving the result for a quick look at the data. Note, if you do not tell it otherwise, the function assumes data are delta oxygen and plots them on a reverse y-axis. However, we can use the `rate.rev` argument to override this

```{r}
inspect.ft(flowthrough_mult.rd, rate.rev = FALSE)
```

The result does not have to be saved. This functionality can be used simply to get a quick overview of a dataset, check there are no common issues, and make sure the columns contain the data we expect. Here, we can see columns 1 to 3 contain declining oxygen, and so as expected are the outflow oxygen data. We can see there are no anomalies in the inflow oxygen data (columns 5 to 8) that might affect calculated rates. We can quickly see the existing delta oxygen columns (9,10,11) have, when the experiment has reached equilibrium, consistent values (around -8), and that the background delta (column 12) is also consistent. 

Note, if multiple columns are inspected like this, and the object saved, subsequent functions such as `calc_rate.ft` will use only the first column of delta oxygen data for analysis. Best practice is to inspect each experimental dataset individually.

## Case Studies

`respR` contains several example flowthrough datasets:
`flowthrough.rd`, `flowthrough_mult.rd`, `flowthrough_sim.rd`. 

We will use these here to run through some specific use cases and example analyses of flowthrough respirometry that should cover most user needs. If there are any not covered please [**get in touch**](mailto:nicholascarey@gmail.com) and we will attempt to accommodate them. 


## Case 1: Outflow oxygen only {#case1}
*"We only have an outflow oxygen recording, but we know the concentration of the inflowing water does not vary much, because it was from an aerated, constant temperature and salinity header tank that we tested to be 100% air saturated frequently. We also know from pilot trials background is negligible, so we don't need to perform an adjustment."*

Here, we will inspect the data, calculate a rate from a stable, consistent region, and convert it to units

### Example data
The data `flowthrough.rd` contains measurements of oxygen consumption in a species of chiton, (*Mopalia lignosa*).  Detailed information about the data, can be obtained with the command `?flowthrough.rd`. 

```{r}
head(flowthrough.rd, n = 4)
```

We can see this dataset contains time values (`secs`), both outflow and inflow oxygen concentrations (`mg/L`), and an oxygen delta, which is simply the difference between outflow and inflow. We will ignore these last two columns here.

### Determine inflow oxygen {#case1oxy}
Knowing the temperature and salinity of the header tank allows us to calculate the saturated concentration (strictly speaking atmospheric pressure is also required, but we will use the default value). Obviously, we want the same oxygen units as the outflow recording.

```{r}
convert_DO(100, from = "%Air", to = "mg/L",
           t = 12, S = 30)
```

### Inspect data
Now we know the concentration of the header tank we can enter this as the `in.o2.value` in `inspect.ft`, which will allow it to calculate an oxygen delta. We use `time` and `out.o2` to specify the respective columns. 

```{r eval = F}
# inspect
insp1 <- inspect.ft(flowthrough.rd, time = 1, out.o2 = 2, in.o2.value = 8.919)
```

```{r eval = T, echo = F}
# inspect
insp1 <- (suppressMessages(inspect.ft(flowthrough.rd, time = 1, out.o2 = 2, in.o2.value = 8.919, plot = F)))
plot(insp1, quiet = TRUE)
```

The data checks show us there are no structural issues with the data, such as missing values or gaps. The top plot shows outflow and the entered, constant inflow oxygen against both time (bottom axis) and row index (top axis), which are the same here because data are recorded every second. The bottom plot shows the calculated delta oxygen values, which is a proxy for the rate. In these plots, consistent oxygen uptake or production rates will be represented by level regions. Here there is some fluctuation towards the end of the data, but the first 400 rows look to have stable rates. 

### Calculate rate {#case1plot}
Calculating a rate requires the `flowrate` of water through the respirometer, in a volume (`ul`, `ml`, or `L`) per unit time (`sec`, `min`, `hr`, or `day`). Here, only the value is required; the units will be entered in a later function. For this dataset experimental data are in `?flowthrough.rd`, and in this example the flowrate is `2.34 mL/min`.
```{r eval = F}
# calculate rate
rate1 <- calc_rate.ft(insp1, from = 1, to = 400, by = "row", flowrate = 2.34)
```

```{r eval = T, echo = F}
# calculate rate
rate1 <- suppressMessages(calc_rate.ft(insp1, from = 1, to = 400, by = "row", flowrate = 2.34, plot = FALSE))
print(rate1)
plot(rate1, quiet = TRUE)
```

`calc_rate.ft` calculates the rate by averaging all delta oxygen values within the chosen data region, and multiplying this by the flowrate. The bottom plot shows this region in close-up. Note, the dashed line is only there to illustrate the trend. 

### Convert rate {#case1rate}

New we take the unitless rate from `calc_rate.ft` and convert it to output units. This can be an absolute rate, that is of the whole animal or chamber, or a mass- or area-specific rate if either of these are entered. Here we will calculate both the absolute and mass-specific rate of this chiton.

Here we need to tell the function the units of oxygen of the original data and of the flowrate, as well as the mass of the specimen in ` kg `. Lastly, we specify the output units.

```{r eval = F, tidy = FALSE}
# absolute rate
rate1_abs <- convert_rate.ft(rate1,  
                             o2.unit = "mg/L",  
                             flowrate.unit = "ml/min",  
                             output.unit = "mg/h")  
# mass-specific rate  
rate1_ms <- convert_rate.ft(rate1,  
                            o2.unit = "mg/L",  
                            flowrate.unit = "ml/min",  
                            mass = 0.000070,  # mass must always be in kg
                            output.unit = "mg/h/g")  
```

```{r eval = T, echo = F}
# absolute rate
rate1_abs <- suppressMessages(convert_rate.ft(rate1, 
                                              o2.unit = "mg/L",
                                              flowrate.unit = "ml/min",
                                              output.unit = "mg/h"))
# mass-specific rate
rate1_ms <- suppressMessages(convert_rate.ft(rate1, 
                                             o2.unit = "mg/L",
                                             flowrate.unit = "ml/min",
                                             mass = 0.000070,
                                             output.unit = "mg/h/g"))
cat("########### Absolute rate ##########")
print(rate1_abs)

cat("\n######## Mass-specific rate ########")
print(rate1_ms)

```

Rates can easily be extracted from the output using `$rate.output`.

```{r}
rate1_final <- rate1_ms$rate.output
print(rate1_final)
```


## Case 2: Outflow and inflow oxygen
*"We have both outflow and inflow oxygen recordings. We also know from pilot trials background is negligible, so we don't need to perform an adjustment."*

This is a very similar analysis to [Case 1](#case1) above, but we will use the inflow oxygen recording as the `in.o2` input. The process is otherwise the same: `inspect.ft`, `calc_rate.ft`, and `convert_rate.ft`.


### Inspect data
```{r eval = F}
# inspect
insp2 <- inspect.ft(flowthrough.rd, time = 1, out.o2 = 2, in.o2 = 3)
```

```{r eval = T, echo = F}
# inspect
insp2 <- (suppressMessages(inspect.ft(flowthrough.rd, time = 1, out.o2 = 2, in.o2 = 3, plot = F)))
plot(insp2, quiet = TRUE)
```

Compared to the `inspect.ft` plot in [Case 1](#case1plot), we can see that at least some of the variation in rate (the delta oxygen values) is explained by variation in inflowing oxygen. This means in this case delta oxygen are more consistent, so we will use the entire dataset to calculate the rate. 

### Calculate rate

Since we are happy the rate is consistent across the whole datatset, running `calc_rate.ft` using the default values will calculate rate as the average of every delta oxygen value multiplied by the flowrate.
```{r eval = F}
# calculate rate
rate2 <- calc_rate.ft(insp2, flowrate = 2.34)
```

```{r eval = T, echo = F}
# calculate rate
rate2 <- suppressMessages(calc_rate.ft(insp2, flowrate = 2.34, plot = FALSE))
print(rate2)
plot(rate2, quiet = TRUE)
```

The slightly lower rate value we get here than in Case 1 (1.65 vs 1.74) suggests the header tank supply is not quite at the theoretical maximum air saturation that we calculated in [Case 1](#case1oxy), and demonstrates that it is always a good idea to have a recording of the inflow water or shared water source if it is practical to do so. 

### Convert rate

We will convert to only the mass-specific rate this time. 

```{r eval = F, tidy = FALSE}
# mass-specific rate  
rate2_ms <- convert_rate.ft(rate2,  
                            o2.unit = "mg/L",  
                            flowrate.unit = "ml/min",  
                            mass = 0.000070,  
                            output.unit = "mg/h/g")  
```

```{r eval = T, echo = F}
# mass-specific rate
rate2_ms <- suppressMessages(convert_rate.ft(rate2, 
                                             o2.unit = "mg/L",
                                             flowrate.unit = "ml/min",
                                             mass = 0.000070,
                                             output.unit = "mg/h/g"))
print(rate2_ms)

```


## Case 3: Outflow and inflow oxygen plus control {#case3}
*"We have both outflow and inflow oxygen recordings. We have also have a concurrent recording from a blank control chamber that we want to use to adjust the specimen rates."*

The `flowthrough_mult.rd` dataset contains several columns of data from a flowthrough experiment using three specimen chambers and one control chamber: *paired* columns of outflow (cols 2,3,4,5) and inflow (cols 6,7,8,9) oxygen recordings, delta oxygen columns (cols 10,11,12,13) calculated from these paired columns, and a recording from a shared header tank (col 14). Other experimental parameters can be seen with `?flowthrough_mult.rd`. The units of time are minutes, and of oxygen percent air saturation (`%Air`). We will convert to actual units of oxygen concentration when we convert the rate. 

### Inspect data

We will use `inspect.ft` to examine the first specimen column pair of outflow and inflow oxygen. 

```{r eval = F}
# inspect
insp3 <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 2, in.o2 = 6)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
insp3 <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 2, in.o2 = 6, plot = F)
plot(insp3, quiet = TRUE)
```

Here we see a warning about not evenly-spaced time values. This comes from this dataset using decimalised minutes as the time values and can be safely ignored (the full `print()` command provides more details). 

This data is typical of flowthrough experiments, in that it takes some time before the experiment stabilises, that is the specimen's respiration rate reaches equilibrium with the supply of oxygen. Here this occurs after around 25 minutes, so we don't want to use any data before this. Rates after this are very stable however. 

### Calculate rate

No inputs for `to` and `by` means the function will calculate the rate from the `from` input to the end of the dataset in the default units of `"time"`. 

```{r eval = F}
# calculate rate
rate3 <- calc_rate.ft(insp3, from = 30, flowrate = 0.1)
```

```{r eval = T, echo = F}
# calculate rate
rate3 <- suppressMessages(calc_rate.ft(insp3, from = 30, flowrate = 0.1, plot = FALSE))
print(rate3)
plot(rate3, quiet = TRUE)
```

### Calculate background rate {#case3bg}

The fourth column pair contains outflow and inflow oxygen from a blank control chamber that we will use to determine the contribution of microbial respiration to the oxygen used, with the assumption that this is the same in the specimen chambers. Note, it is important that background experiments are conducted with the same equipment, under the same conditions, and using the *same flowrate* as specimen experiments, and data are in the same units. They do not necessarily have to be run concurrently; `respR` allows you to save a background rate to adjust multiple different experiments. 

```{r eval = F}
# inspect
bg <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 5, in.o2 = 9)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
bg <- suppressWarnings(suppressMessages(inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 5, in.o2 = 9)))
plot(bg, quiet = TRUE)
```

Here we see there is a slight but noticeable difference in outflow and inflow oxygen due to microbial action. Moreover, it appears to be constant, so we can use the entire dataset to calculate a background rate. 

We use the same `calc_rate.ft` function to calculate and save background rates. 

```{r eval = F}
# calculate rate
bgrate <- calc_rate.ft(bg, flowrate = 0.1)
```

```{r eval = T, echo = F}
# calculate rate
bgrate <- suppressMessages(bgrate <- calc_rate.ft(bg, flowrate = 0.1, plot = FALSE))
print(bgrate)
plot(bgrate, quiet = TRUE)
```

### Adjust rate

Now we use the saved background rate object to adjust the specimen rate we determined earlier.

```{r eval = F}
# adjust rate
rate3adj <- adjust_rate.ft(rate3, by = bgrate)
```

```{r eval = T, echo = F}
# calculate rate
rate3adj <- suppressMessages(adjust_rate.ft(rate3, by = bgrate))
print(rate3adj)
```

We can see this performs a small but significant adjustment to the specimen rate.

Note how `adjust_rate.ft` accepts `calc_rate.bg` objects for both inputs. This same background rate object can similarly be applied to the other specimen experiments in this dataset, or indeed other datasets if they are collected under the same conditions. However, it also accepts numeric values. This will give the exact same result. 

```{r eval = F}
# adjust rate
rate3adj <- adjust_rate.ft(rate3, by = -0.03174)
```

```{r eval = T, echo = F}
# adjust rate
rate3adj <- suppressMessages(adjust_rate.ft(rate3, by = -0.03174))
print(rate3adj)
```

This means you can quantify a value for a background rate and apply it to many experiments as you see fit, for example the mean rate of several background experiments. In fact, if you enter multiple background rates this is the default behaviour. 

```{r eval = F}
# adjust rate
adjust_rate.ft(rate3, by = c(-0.030, -0.025, -0.040))
```

```{r eval = T, echo = F}
# adjust rate
tmp <- suppressMessages(adjust_rate.ft(rate3, by = c(-0.030, -0.025, -0.040)))
print(tmp)
```

***Note:*** be careful when entering rates manually like this. In `respR` oxygen uptake rates are negative since they represent a negative slope of oxygen against time. Background rates are typically similarly *negative* (though not always). If you are examining oxygen production, rates are positive, and background rates may be either sign. To summarise: oxygen removal or use is always negative, oxygen input or production is always positive. 

### Convert adjusted rate

Now we convert the adjusted rate. We will convert to an absolute rate this time, and a different output oxygen unit. 

Because the original oxygen data were in units of percent air saturation, we need to enter the temperature, salinity and atmospheric pressure to convert these to actual concentration units of oxygen. These can be found the data help file: `?flowthrough_mult.rd`. 

```{r eval = F, tidy = FALSE}
# absolute rate  
rate3_abs <- convert_rate.ft(rate3adj,  
                             o2.unit = "%Air",  
                             flowrate.unit = "L/min",  
                             output.unit = "ml/h",
                             t = 18, S = 0, P = 1.013)  
```

```{r eval = T, echo = F}
# absolute rate  
rate3_abs <- suppressMessages(convert_rate.ft(rate3adj,  
                                              o2.unit = "%Air",  
                                              flowrate.unit = "L/min",  
                                              output.unit = "ml/h",
                                              t = 18, S = 0, P = 1.013))
print(rate3_abs)

```


## Case 4: Delta oxygen values
*"We have already calculated a delta oxygen between outflow and inflow oxygen concentrations."*

For cases where delta oxygen between outflow and inflow concentrations has already been calculated, these data can be inspected using the `delta.o2` input and used to calculate rates. `flowthrough_mult.rd` contains these in columns 10 to 13. We will use the second column of these. 

### Inspect data

```{r eval = F}
# inspect
insp4 <- inspect.ft(flowthrough_mult.rd, time = 1, delta.o2 = 11)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
insp4 <- suppressWarnings(suppressMessages(inspect.ft(flowthrough_mult.rd, time = 1, delta.o2 = 11)))
plot(insp4, quiet = TRUE)
```

When delta oxygen data are inspected, only one plot is produced of the entered delta oxygen values. Again we see the experiment has not reached equilibrium until after around 20 minutes, so we will not use any data before this. 

### Calculate rate

This time we will calculate rates using subsetting by `"row"` which can be see in the top red x-axis.  

```{r eval = F}
# calculate rate
rate4 <- calc_rate.ft(insp4, from = 2000, to = 3000, by = "row", flowrate = 0.1)
```

```{r eval = T, echo = F}
# calculate rate
rate4 <- suppressMessages(calc_rate.ft(insp4, from = 2000, to = 3000, by = "row", flowrate = 0.1, plot = F))
print(rate4)
plot(rate4, quiet = TRUE)
```

### Adjust rate

We already determined background rate in [Case 3](#case3bg), so we can use the same object here.

```{r eval = F}
# adjust rate
rate4adj <- adjust_rate.ft(rate4, by = bgrate)
```

```{r eval = T, echo = F}
# calculate rate
rate4adj <- suppressMessages(adjust_rate.ft(rate4, by = bgrate))
print(rate4adj)
```

### Convert adjusted rate

Now we convert the adjusted rate. 

```{r eval = F, tidy = FALSE}
# mass-specific rate  
rate4_abs <- convert_rate.ft(rate4adj,  
                             o2.unit = "%Air",  
                             flowrate.unit = "L/min",  
                             output.unit = "ml/h",
                             t = 18, S = 0, P = 1.013)  
```

```{r eval = T, echo = F}
# mass-specific rate
rate4_abs <- suppressMessages(convert_rate.ft(rate4adj,  
                                              o2.unit = "%Air",  
                                              flowrate.unit = "L/min",  
                                              output.unit = "ml/h",
                                              t = 18, S = 0, P = 1.013))
print(rate4_abs)

```


## Case 5: Active rates
*"We imposed a treatment at a specific time to increase the specimen's activity and want to determine an active metabolic rate from this region of the data."*

### Inspect data
Inspecting the data from the third chamber, we can see a prominent area where the specimen's oxygen consumption rate increased, before slowly recovering to routine levels. 

```{r eval = F}
# inspect 
insp5 <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 4, in.o2 = 8)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
insp5 <- suppressWarnings(suppressMessages(inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 4, in.o2 = 8, plot = FALSE)))
plot(insp5, quiet = TRUE)
```

Note how delta oxygen are plotted on a reverse y-axis, so that higher uptake rates are plotted higher, even though they are lower numerically (more negative). If you are examining oxygen *production* rates, you can pass the `rate.rev = FALSE` argument in the main call or when plotting the output to see higher production rates plotted higher. 

Since we are only interested in this active region, we can use the `subset_data()` function and pipe (`%>%`) the result to `inspect.ft` for a closer look. 

```{r eval = F}
# subset and inspect
insp5 <- subset_data(flowthrough_mult.rd, from = 30, to = 50, by = "time") %>%
  inspect.ft(time = 1, out.o2 = 4, in.o2 = 8)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
insp5 <- suppressWarnings(suppressMessages(subset_data(flowthrough_mult.rd, from = 30, to = 50, by = "time") %>%
                                             inspect.ft(time = 1, out.o2 = 4, in.o2 = 8, plot = FALSE)))
plot(insp5, quiet = TRUE)
```

Note, the row index x-axis now refers to the subset not the original data. We can see the highest active rate occurs between around 38 and 42 minutes, so we will use this region to calculate the active rate.

### Calculate rate

```{r eval = F}
# calculate rate
rate5 <- calc_rate.ft(insp5, from = 38, to = 42, by = "time", flowrate = 0.1)
```

```{r eval = T, echo = F}
# calculate rate
rate5 <- suppressMessages(calc_rate.ft(insp5, from = 38, to = 42, by = "time", flowrate = 0.1, plot = FALSE))
plot(rate5, quiet = TRUE)
```

### Adjust rate

We can also adjust this rate by the background rate we saved earlier. 

```{r eval = F}
# adjust rate
rate5adj <- adjust_rate.ft(rate5, by = bgrate)
```

```{r eval = T, echo = F}
# calculate rate
rate5adj <- suppressMessages(adjust_rate.ft(rate5, by = bgrate))
print(rate5adj)
```

### Convert adjusted rate

Lastly we convert the adjusted rate, this time to a mass-specific rate, and try another different oxygen amount metric in the output units.

```{r eval = F, tidy = FALSE}
# mass-specific rate  
rate5_ms <- convert_rate.ft(rate5adj,  
                            o2.unit = "%Air",  
                            flowrate.unit = "L/min",  
                            output.unit = "umol/h/g",
                            mass = 0.020,
                            t = 18, S = 0, P = 1.013)  
```

```{r eval = T, echo = F}
# mass-specific rate
rate5_ms <- suppressMessages(convert_rate.ft(rate5adj,  
                                             o2.unit = "%Air",  
                                             flowrate.unit = "L/min",  
                                             output.unit = "umol/h/g",
                                             mass = 0.020,
                                             t = 18, S = 0, P = 1.013))
print(rate5_ms)

```


## Case 6: Non-constant background rates
*"We want to correct for both non-constant background activity and any possible fluctuation in inflow oxygen concentrations using a concurrent empty control or \"blank\" chamber."*

In some experiments, as well as possible fluctuations in the oxygen content of the inflowing water, the background rate may change over the course of the experiment. While the former will be accounted for by having an inflow oxygen recording, they can both be accounted for simultaneously by using the outflow recording from a blank control chamber to adjust a specimen chamber. This can be done by using it as the `in.o2` input in `inspect.ft`. 

The `flowthrough_sim.rd` dataset contains data from an experiment where the background rate increases as the experiment progresses. There are four columns; time, outflow oxygen from the specimen and blank chambers, and the inflow oxygen in a header tank supplying both.

### Inspect data

We will inspect the specimen chamber as we normally would, using the header tank as the inflow recording. 

```{r eval = F}
# inspect 
insp6 <- inspect.ft(flowthrough_sim.rd, time = 1, out.o2 = 2, in.o2 = 4)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
insp6 <- suppressMessages(inspect.ft(flowthrough_sim.rd, time = 1, out.o2 = 2, in.o2 = 4, plot = FALSE))
plot(insp6, quiet = TRUE)
```

Here we see, after the initial period before the experiment has reached equilibrium, the specimen rate apparently continues to increase. This suggests that the background rate may not be constant. 

### Inspect background

If we examine the background data in the same way:

```{r eval = F}
# inspect 
bg <- inspect.ft(flowthrough_sim.rd, time = 1, out.o2 = 3, in.o2 = 4)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
bg <- suppressMessages(inspect.ft(flowthrough_sim.rd, time = 1, out.o2 = 3, in.o2 = 4, plot = FALSE))
plot(bg, quiet = TRUE)
```

We can see the background rate increases as the experiment progresses, in what looks like a constant manner. With the presumption that this is also occurring in the specimen chamber, we only need to use the outflow recording from the control chamber as the `in.o2` input to account for this. 

### Account for background

```{r eval = F}
# inspect 
insp6 <- inspect.ft(flowthrough_sim.rd, time = 1, out.o2 = 2, in.o2 = 3)
```

```{r eval = T, echo = F, results = "hide"}
# inspect
insp6 <- suppressMessages(inspect.ft(flowthrough_sim.rd, time = 1, out.o2 = 2, in.o2 = 3, plot = FALSE))
plot(insp6, quiet = TRUE)
```

Now we see that when the background is accounted for, the specimen rates are consistent, and we can go ahead and calculate a rate.

### Calculate rate

```{r eval = F}
# calculate rate
rate6 <- calc_rate.ft(insp6, from = 2000, to = 3000, by = "row", flowrate = 0.1)
```

```{r eval = T, echo = F}
# calculate rate
rate6 <- suppressMessages(calc_rate.ft(insp6, from = 2000, to = 3000, by = "row", flowrate = 0.1, plot = FALSE))
plot(rate6, quiet = TRUE)
```


The rate can now be converted as in the above examples. 

Note, that in this case the background rate increased in a constant, linear manner. Under other conditions, the background may vary in other ways, such as increase exponentially, or decrease due to lower light levels at night. Inflowing oxygen may also vary due to various reasons. It's important to note that the above method of accounting for background and other fluctuations in oxygen will work for *all* these cases, under the presumption that the same thing is occurring in specimen chambers. 

## Case 7: Single values only

*"We do not have continuous recordings. We took spot readings of inflow and outflow recordings, after they had stabilised (i.e. the experiment had reached equilibrium), and we want to use these to calculate the rate."*

Flowthrough respirometry has advantages over other respirometry methods, in that because there is a constant supply of water (and therefore oxygen) experiments can be run for much longer, allowing specimens to acclimate better to the experimental conditions without concerns about effects of hypoxia or waste build up. They can also be more practical and a more efficient use of equipment, in that continuous recordings of oxygen (while ideal practice) are not always necessary as long as experiments are monitored to ensure they have reached equilibrium. In these cases, a single oxygen probe could be used to sample the outflow and inflow oxygen of any number of separate chambers, and so in cases of limited equipment availability help with increasing the amount of data collected. 

All the functions (where possible) in `respR` accept numeric inputs, as well as other `R` objects such as data frames and data tables. As long as care is taken to keep units and inputs consistent, these can be used to calculate rates. In this example, after monitoring to ensure equilibrium has been reached, spot readings can be used to calculate the routine respiration rate of specimens. 

### Create dataframe

In this example, we have a spot sample of outflow and inflow oxygen. We just need to put these into a data frame with any `time` value to be able to process them in `inspect.ft`.

```{r}
## Single spot checks of outflow and inflow
## Create dataframe
df <- data.frame(time = 1,
                 outflow = 7.32,
                 inflow = 8.04)
```

### Inspect, calculate rate, convert

Now all we need to do is the usual workflow: `inspect.ft` > `calc_rate.ft` > `convert_rate.ft`.

```{r eval = F, tidy = FALSE}
insp7 <- inspect.ft(df, time = 1, out.o2 = 2, in.o2 = 3)
rate7 <- calc_rate.ft(insp7, flowrate = 0.25)
rate7_abs <- convert_rate.ft(rate7,
                             o2.unit = "mg/l",  
                             flowrate.unit = "L/min",  
                             output.unit = "mg/h")
```

```{r eval = T, echo = F, message=FALSE, warning=FALSE, results = "hide"}
insp7 <- inspect.ft(df, time = 1, out.o2 = 2, in.o2 = 3, plot = F)
```

```{r eval = T, echo = F, message=FALSE, warning=FALSE}
rate7 <- calc_rate.ft(insp7, flowrate = 0.25, plot = F)
rate7_abs <- convert_rate.ft(rate7,
                             o2.unit = "mg/l",  
                             flowrate.unit = "L/min",  
                             output.unit = "mg/h")

print(rate7)
print(rate7_abs)

```

### Delta oxygen values

In fact, things are even simpler if spot readings are converted to delta oxygen values, as `calc_rate.ft` can convert these directly. We can also adjust the rate in `adjust_rate.ft` using a rate calculated from similar spot readings from a control chamber. 

```{r eval = F, tidy = FALSE}
## specimen delta oxygen = outflow minus inflow
del7 <- 7.32 - 8.04
## control delta oxygen
delbg <- 7.97 - 8.04

rate7 <- calc_rate.ft(del7, flowrate = 0.25)
ratebg <- calc_rate.ft(delbg, flowrate = 0.25)

rate7_adj <- adjust_rate.ft(rate7, by = ratebg)

rate7_abs <- convert_rate.ft(rate7_adj,
                             o2.unit = "mg/l",  
                             flowrate.unit = "L/min",  
                             output.unit = "mg/h")
```

```{r eval = T, echo = F, message=FALSE, warning=FALSE}
del7 <- 7.32 - 8.04
print("# del7")
del7
delbg <- 7.97 - 8.04
print("# bg7")
delbg

rate7 <- calc_rate.ft(del7, flowrate = 0.25, plot = FALSE)
ratebg <- calc_rate.ft(delbg, flowrate = 0.25, plot = FALSE)

rate7_adj <- adjust_rate.ft(rate7, by = ratebg)

rate7_abs <- convert_rate.ft(rate7_adj,
                             o2.unit = "mg/l",  
                             flowrate.unit = "L/min",  
                             output.unit = "mg/h")
print(rate7)
print(rate7_adj)
print(rate7_abs)

```

### Multiple delta oxygen values

You can even use vectorised operations to convert multiple rates.

```{r eval = F, tidy = FALSE}
## specimen delta oxygen = outflow minus inflow
outflows <- c(7.32, 7.45, 7.19, 7.27)
## control delta oxygen
inflows <- 8.04

deltas <- outflows - inflows
rates <- calc_rate.ft(deltas, flowrate = 0.25)
rates_abs <- convert_rate.ft(rates,
                             o2.unit = "mg/l",  
                             flowrate.unit = "L/min",  
                             output.unit = "mg/h")
```


```{r eval = T, echo = F, message=FALSE, warning=FALSE}
## specimen delta oxygen = outflow minus inflow
outflows <- c(7.32, 7.45, 7.19, 7.27)
## control delta oxygen
inflows <- 8.04

deltas <- outflows - inflows

rates <- calc_rate.ft(deltas, flowrate = 0.25, plot = FALSE)
rates_abs <- convert_rate.ft(rates,
                             o2.unit = "mg/l",  
                             flowrate.unit = "L/min",  
                             output.unit = "mg/h")
print(rates_abs)
summary(rates_abs)

```


By default the first output is printed, but you can print others by using the `pos` input, e.g. `print(rates_abs, pos = 2)` or by using `summary()` to see all results. 

## Case 8: Rolling rate

*"We want to calculate every rate of a fixed width across the entire dataset and then filter the results manually. We are interested in the routine metabolic rate from a particular region of the data."*

`calc_rate.ft` can calculate a rolling rate of fixed width (in rows) across the dataset. In this case there will be multiple rates in `output$rate` and the same number of rows in `output$summary`. The parameters in the summary table can be used to filter out only the results of interest using regular R syntax.

Note, the `calc_rate.ft` summary table contains linear regression coefficients alongside other metadata. These *should not be confused* with those in other functions such as `calc_rate` where slopes represent rates and coefficients such as a high r-squared are important. Here, slope represents the stability of the data region, in that the closer the slope is to zero, the less the delta oxygen values in that region vary, which is an indication of a region of stable rates. They are included to enable possible future functionality where stable regions may be automatically identified, and should generally be ignored. However, advanced users can use regular R syntax to explore and subset the results using these if they wish. 

In this example, we want to extract the most consistent rate sustained across a five minute window from a particular region of the data. 

### Inspect data
```{r eval = F}
## inspect
insp8 <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 4, in.o2 = 7)
```

```{r eval = T, echo =F, results = "hide", message=FALSE, warning=FALSE}
## inspect
insp8 <- insp8 <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 4, in.o2 = 7, plot = F)
plot(insp8)
```

Here, we see a region before the experiment has reached equilibrium, and a region of higher rates. To get a routine rate we therefore want to extract it from approximately the 10 to 30 minute time range.

### Calculate rate across rolling window

In these data (per second, but in decimalised minutes) a five minute window would be 300 rows. 

```{r eval = F}
# calculate rate
roll_rate <- calc_rate.ft(insp8, width = 300, by = "row", flowrate = 0.1)
```

```{r eval = T, echo = F}
# calculate rate
roll_rate <- calc_rate.ft(insp8, width = 300, by = "row", flowrate = 0.1, plot = F)
summary(roll_rate)
```

### Filter results 

We can see there are 3441 results, the slope is in the third column, and the start and end times of each regression are in columns 7 and 8. Here, we will subset regressions with start times after 10 minutes, end times before 30 minutes, and then identify the one with the slope closest to zero. Note, slopes will normally be a mix of positive and negative, so we want the one with the lowest *difference* from zero.

```{r}
## summary rows with start time > 10
summ <- roll_rate$summary[which(roll_rate$summary$time > 10),]

## summary rows with end time < 30
summ <- summ[which(summ$endtime < 30),]

## which has slope with lowest absolute difference from zero?
summ <- summ[which(abs(summ$slope_b1 - 0) == min(abs(summ$slope_b1 - 0))),]

print(summ)
```

We can see from the `$rank` column, the 1044th result is in the time range of 10 to 30 and has the slope closest to zero. Note also how the `intercept_b0` is equal to the `delta_mean` (mean of all oxygen delta values within the data region), which is another indication of the stability of rates within the region. We can plot this particular result using `pos`. 

```{r eval = F}
plot(roll_rate, pos = 1044)
```

```{r eval = T, echo = F, results = "hide"}
plot(roll_rate, pos = 1044)
```

### Convert rate

To get the final rate we just need to adjust and convert all rates, and extract the one we are interested in. The `summary` command along with `pos` and `export` options can extract an exact row (or range of rows) to a separate data frame. 

```{r eval = F, tidy = FALSE}
roll_rate_adj <- adjust_rate.ft(roll_rate, by = bgrate)

roll_rate_ms <- convert_rate.ft(roll_rate_adj,  
                                o2.unit = "%Air",  
                                flowrate.unit = "L/min",  
                                output.unit = "umol/h/g",
                                mass = 0.020,
                                t = 18, S = 0, P = 1.013)

result <- summary(roll_rate_ms, pos = 1044, export = TRUE)
```

```{r eval = T, echo = F}
# mass-specific rate
roll_rate_adj <- suppressMessages(adjust_rate.ft(roll_rate, by = bgrate))

roll_rate_ms <- suppressMessages(convert_rate.ft(roll_rate_adj,  
                                                 o2.unit = "%Air",  
                                                 flowrate.unit = "L/min",  
                                                 output.unit = "umol/h/g",
                                                 mass = 0.020,
                                                 t = 18, S = 0, P = 1.013))

result <- suppressMessages(summary(roll_rate_ms, pos = 1044, export = TRUE))
```


This is just an example. Other results and columns in `$summary` can be used in a similar way to filter results, for instance to extract maximum or minimum rate values, only those within a specific value range, or all those within a specific percentile of rates.

## Case 9: Multiple rates from the same dataset

*"We want to extract rates from several different time periods during an experiment."*

As can be seen in the above example, where we used a rolling window, multiple rates can be extracted from different regions of the same dataset, even overlapping ones. This can be done by entering vectors of paired start and end values as the `from` and `to` inputs in the appropriate `by` units. 

### Inspect and calculate rates
```{r eval = F, tidy = FALSE}
# inspect 
insp9 <- inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 3, in.o2 = 7)
# calculate rates
rate9 <- calc_rate.ft(insp9, 
                      from = c(25, 35, 45), 
                      to = c(30, 40, 50), 
                      by = "time", 
                      flowrate = 0.1)
```

```{r eval = T, echo = F, results = "hide"}

insp9 <- suppressWarnings(suppressMessages(inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 3, in.o2 = 7, plot = FALSE)))

rate9 <- suppressMessages(calc_rate.ft(insp9, 
                                       from = c(25, 35, 45), 
                                       to = c(30, 40, 50), 
                                       by = "time", 
                                       flowrate = 0.1, plot = F))
```

### Examine output

The different rates can be plotted using the `pos` input. 

```{r eval = T, tidy = FALSE}
plot(rate9, pos = 2)

```

And they can all be viewed using `summary`. 

```{r eval = T, tidy = FALSE}
summary(rate9)
```

### Convert results

The results can be adjusted and converted as in the above examples. Both `adjust_rate` and `convert_rate` accept objects containing multiple rates and can be plotted and explored in the same way. 



## Case 10: Concise, reportable analysis using piping

*"We want to report our analyses in a reproducible way using as little space as possible."*

The overall aim of `respR` is to enable easy reporting of reproducible data analyses. This example is simply to show how a complete analysis of a flowthrough respirometry experiment can be reported concisely. 

We will repeat the analysis from [Case 3](#case3), but use piping to feed the results of one function into the next. We will use `%>%` or `dplyr` pipes here, but the new native `|>` pipes introduced in [R v4.1](https://www.r-bloggers.com/2021/05/new-features-in-r-4-1-0/) work just as well. 

```{r eval = FALSE, tidy = FALSE}
## calc background rate (might only need to be done once)
bgrate <- 
  inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 5, in.o2 = 9) %>%
  calc_rate.ft(flowrate = 0.1)

# inspect > calc rate > adjust > convert
rate3_abs <- 
  inspect.ft(flowthrough_mult.rd, time = 1, out.o2 = 2, in.o2 = 6) %>%
  calc_rate.ft(from = 30, flowrate = 0.1) %>%
  adjust_rate.ft(by = bgrate) %>%
  convert_rate.ft(o2.unit = "%Air",  
                  flowrate.unit = "L/min",  
                  output.unit = "ml/h",
                  t = 18, S = 0, P = 1.013)

print(rate3_abs)

```

```{r, eval = T, echo = F}
print(rate3_abs)
```

## Notes

- Note, that while it is required (for `inspect.ft` at least), `time` data are never used in calculations, only for plotting and choosing regions. Higher resolution data will give more delta oxygen values to average, but in general the resolution or recording frequency of the data is less important than other factors such as ensuring equilibrium has been reached. 

- To repeat, always keep in mind the sign of the rate. Oxygen consumption (respiration) rates are represented by negative values, production rates by positive values. 

- The functions in the flowthrough workflow, where appropriate, support the generic S3 functions `plot`, `print`, `summary`, and `mean`. These have various additional arguments such as `pos`, `quiet`, and `export`. See help files for additional information. In the case of `mean`, if there are multiple rates they return the mean value of the primary output (`rate`, `rate.adjusted`, or `rate.output`). This can be output as a value by passing `export = TRUE`. 
