---
title: "Importing your data"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Importing your data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 6, 
               fig.height = 4,
               fig.align = "center")
```

## Introduction

`respR` was designed to be a universal, end-to-end solution for analysing data and reporting analyses from any and all aquatic respirometry experiments, regardless of the equipment used. 
Therefore, it is system-agnostic; the data need only be put into a very simple structure (numeric time against oxygen in any common units) to allow a full analysis to be conducted. 
In fact, the entire package, with the exception of the final conversion step in `convert_rate`, treats data as unitless, so non-aquatic respirometry data, or any time-series data can be explored and analysed. 

Each function has been designed to output objects which can be entered into subsequent functions reducing the need for additional inputs. 
However, generic `R` data objects, including values, vectors and objects of class `data.frame`, `data.table` and `tibble`, are also recognised by most functions.
The only structural data requirement is that the initial time\~oxygen data be in a specific structure: paired values of numeric time-elapsed (in secs, mins, hours or days) and oxygen amount (in any commonly used unit). 
Every respirometry system, to our knowledge, allows data to be exported in such a format, or at least in a structure from which it is easy to parse it to this format. 

Two functions are provided to assist with bringing in and formatting your data correctly. 

## `import_file()`

Most systems allow data to be exported in easily readable formats (e.g. csv, txt) which contain the numeric time and O2 data `respR` requires. These files are usually easily imported into R using generic functions such as `read.csv()` and the relevant columns specified when used in `respR` functions, or extracted into separate data frames. 

Many systems however have raw output files with redundant information, or a structure that confuses importing functions. For example Loligo Systems AutoResp and Witrox files have several rows of metadata above the columns of raw data. While `read.csv()` has options to not import certain rows and columns, `respR` can import many raw data files from various systems without modification. 

The `import_file()` function uses pattern recognition to identify the originating system of the file, imports it, formats columns and column names, and outputs a data frame, that can be passed to `respR` functions.

This is an example import of a Witrox raw data file (situated in the current working directory, otherwise any external file can be specified with a filepath):
```{r eval = T}
data <- import_file("Witrox_eg.txt")

## There are six columns, but we will print only 
##the relevant date, numeric timestamp, and oxygen columns
data[,c(1,2,6)]
```

As we can see, the function automatically recognises that this is a Witrox file, removes redundant rows, and renames the columns, removing spaces while keeping unique identifiers. 

This function requires only a single input, the path to the file. Everything else is handled automatically. The function will even automatically recognise data formatted in the European style using commas for decimals.

`import_file` supports several systems at present, including Firesting, Pyro, PreSens, MiniDOT, Loligo Witrox, Vernier and more. See `help(import_file)` for full list. 
However, some files may fail to import as expected because of structural or version differences we have not encountered. We would encourage users to [send us sample files for testing](mailto:nicholascarey@gmail.com), especially any they have problems with, or from systems we do not yet support. 


## `format_time()`

For files types that are not yet supported, or if you have already imported your data by other means, the `format_time()` function can parse date-time columns to numeric time-elapsed data, if the imported file does not contain it. 

Here's an example of a 2-column data frame with date-time data and oxygen. 

```{r  echo = F, message = F, results="hide"}
data <- import_file("Witrox_eg.txt")
data <- data[,c(1,6)]
data_v <- data[[1]]
names(data) <- c("Date_Time", "O2_mg/L")
```

```{r}
head(data, n = 5)
```

`format_time` can parse these data to numeric (internally, it uses functionality in the package `lubridate()`). 
The date-times can either be passed as a `vector` (for example, so it can be appended to the original data), or as a `data frame`. If the input is a vector, the output is a vector of the same length. If a data frame, the `time` input indicates the column number (or numbers, if it is split across several) containing the date-times, the default being `time = 1`. The resulting data frame will be identical, with a new column named `time_num` with the converted numeric time added as the *last* column. 

We also need to specify the `format` of the date-times in the correct order. See `help(format_time)` for further info.

```{r}
## Pass as vector
data_2 <- format_time(data[[1]], format = "dmyHMSp")
head(data_2)

## Pass as data frame
data_3 <- format_time(data, format = "dmyHMSp")
head(data_3)
```

By default, the numeric time data will start at 1, but we can override this. This could be useful if data are split into separate files, and you want to append the start of one onto the end of another, or you simply want to link a specific numeric time value to the start of the experiment. 

```{r}
## as data frame
data_4 <- format_time(data, format = "dmyHMSp", start = 1000)
head(data_4)
```

Note, numeric time data will always output in *seconds* regardless of the input format.

## Dealing with timed events or notes

What if there are important notes or events associated with specific times in your experiment? For example, flushing of chambers, imposing a new swimming speed, changing the temperature, noting a response, etc. Resetting the times via formatting the time data may make these difficult to associate to certain stages of the analysis. This is easily dealt with by formatting the times of the events in the same way you formatted the data. You only need to make sure at least one event is associated with the same start time you used for experimental data. 

Here's an example of some experimental notes (in some systems such notes can be entered in the software, and so may be included in output files, or they could be copied from a lab book into a csv file and imported). 

``` {r echo = F, message = F, results="hide"}
## Creating example data frame of Times and 'Events' or experimental notes
times <- c("8/17/2016 9:42:02", "8/17/2016 9:52:02", "8/17/2016 9:54:34", "8/17/2016 10:19:02", "8/17/2016 12:04:54", "8/17/2016 14:31:22")
events <- c("Experiment start", "Flush period start", "Flush period end", "Specimen acting normally", "Went to lunch", "Swim speed set to 20 cm/s")

exp_notes <- data.frame(times = times, events = events)
```

```{r}
exp_notes

format_time(exp_notes, format = "mdyHMS")
```

These do not have to be in the same date-time format, or even at the same precision, depending on how accurately you need to know when events occurred. The important factors are associating at least one event with the *same start time* used to format the experimental time data, and using the correct `format` setting. 

``` {r echo = F, message = F, results="hide"}
## Creating example notes
times <- c("9:42", "9:52", "9:54", "10:19", "12:04", "14:31")
events <- c("Experiment start", "Flush period start", "Flush period end", "Specimen acting normally", "Went to lunch", "Swim speed set to 20 cm/s")

exp_notes <- data.frame(times = times, events = events)
```

```{r}
format_time(exp_notes, format = "HM")

```

## Next steps

After your data is in this paired, *numeric time~O2* structure, it can be passed to `inspect()` or other functions for analysis. See `vignette(inspecting)`.


