---
title: "Intermittent-flow respirometry: Short experiment"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Intermittent-flow respirometry: Short experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(width = 999))
```

Intermittent-flow respirometry involves obtaining multiple replicates in an experiment by periodically flushing the respirometer with new water or air after a set period of time or O~2~ decrease. 
Typically, the oxygen sensing system is left recording during flushes, and the system returning to normoxia can be seen in the data trace. 
If the dataset is not split into the separate replicates, this results in an dataset with oscillating levels of oxygen. 
This vignette shows an example of how an intermittent-flow respirometry experiment can be analysed in `respR` to extract oxygen uptake or production rates. See `vignette("intermittent_long")` for an example of how a much longer, more regular intermittent-flow experiment with an increasing background rate can be analysed.

## Example data

The example data, `intermittent.rd`, contains intermittent-flow measurements of oxygen consumption in the sea urchin *Heliocidaris erythrogramma*. 
Detailed information about the data can be viewed via `?intermittent.rd`. 

## Inspecting the data {#inspect}

The first step in any analysis should always be to visualise and inspect the data. For this we use the `inspect()` function. This scans for common issues in the data frame, plots the data, and saves it to an object to pass to the next stage of analysis.

```{r eval = F}
urchin_int <- inspect(intermittent.rd)
```

```{r echo = F, warning=FALSE}
inspect(intermittent.rd)
```

```{r echo = F, fig.keep = "none", message = FALSE, results = "hide"}
urchin_int <- inspect(intermittent.rd)
```

We can see from the top plot that this dataset has three replicates, separated by two flushes where new water was added. The flush periods obviously skew the bottom plot, which shows rate of change (positive or negative) in oxygen over a rolling window of 10% of the data. Note how rates are plotted on a reversed y-axis so that uptake rates are higher than production rates, or oxygen input rate in the case of the flushes (this can be changed by passing `rate.rev = FALSE`). 

This plot provides a quick visual inspection of how the rate varies over the course of the experiment.
Regions of stable and consistent rates can be identified on this plot as flat
or level areas. This plot is for exploratory purposes only; later functions
allow rate to be calculated over specific regions. Here, despite the flushes skewing the rate plot, specimen uptake rates (flat regions) appear to be relatively consistent within each experimental replicate with a value of around -0.005. 

For a quick closer look, we can always use `inspect` without saving the result to check regions of the data.

```{r results='hide', message=FALSE}
inspect(intermittent.rd[1:1800,])
```


## Analysis with `calc_rate` {#calc_rate}
Rates can be extracted from intermittent data via the `calc_rate()` function by using it to choose the regions of the data which make up the replicates. 

### Rate from time region
The `calc_rate` function allows data regions to be chosen by `oxygen`, `time`, or `row` ranges. In the case of`row` or `time`, multiple subset regions can be specified. The `from` and `to` operators can be vectors with paired values. Using this we can extract a rate from each replicate with one command. 

```{r tidy = FALSE, results = "hide"}
urchin_int_rates <- calc_rate(urchin_int, 
                              from = c(200, 2300, 4100), 
                              to = c(1800, 3200, 4600), 
                              by = "time")

```

```{r echo = F}
print(urchin_int_rates)
```

By default, the first is shown in `print` and `plot`, but the `pos` input can be used to view others. 

```{r results='hide'}
plot(urchin_int_rates, pos = 2)
```

Calling `summary()` will show the coefficients, locations and values of all rates:
```{r}
summary(urchin_int_rates)
```

See the [Converting rates](#convert) section for how to convert these to rates in specific units.

### Rate from row region

The rate region(s) can also be selected using row numbers. In this particular dataset, where oxygen has been recorded every second, these are equivalent to the time values, but this won't be the case for every dataset.  

```{r results = "hide"}
calc_rate(urchin_int, from = 400, to = 1600, by = "row")
```


### Rate from oxygen region {#oxygen}

If we are interested in looking at rates over particular oxygen ranges within each replicate, we need to take a slightly different approach, because time and row values are unique and increase regularly, while oxygen does not. 

Here, we want to calculate rates between the oxygen values of 7 and 6.6 mg/L within each replicate. There are several approaches to this. We could split up our original data frame into separate ones for each replicate. Or we could `inspect` each replicate separately, using regular R syntax before passing it to `calc_rate`. 

```{r eval = F}
u_rep1 <- inspect(intermittent.rd[1:1900,])
u_rate1 <- calc_rate(rep1, from = 7, to = 6.6, by = "o2")
```

However, we can also use the dedicated `subset_data()` function, which allows for easy subsetting by time, oxygen or row number from both data frames and `inspect` objects. Here, we use the `inspect` object we [saved earlier](#inspect) using the whole dataset, to create new `inspect` objects for each replicate. These can be treated like any other `inspect` object, including being passed to `print` and `plot`.

```{r fig.keep="none", message=FALSE, warning=FALSE, results="hide"}
# Create separate replicate data frames
u_rep1 <- subset_data(urchin_int, from = 1, to = 1900, by = "time")
u_rep2 <- subset_data(urchin_int, from = 2100, to = 3500, by = "time")
u_rep3 <- subset_data(urchin_int, from = 3700, to = 4831, by = "time")
```

Now we can calculate the rate from each, showing the results from the third one here.
```{r eval = F}
u_rate1 <- calc_rate(u_rep1, from = 7, to = 6.6, by = "o2")
u_rate2 <- calc_rate(u_rep2, from = 7, to = 6.6, by = "o2")
u_rate3 <- calc_rate(u_rep3, from = 7, to = 6.6, by = "o2")
```

```{r eval = T, echo = F, results = "hide"}
u_rate3 <- calc_rate(u_rep3, from = 7, to = 6.6, by = "o2")
```

See later section [Converting rates](#convert) for how to convert these to units.


## Analysis with `auto_rate`
To more rigorously identify *lowest*, *highest* or *most linear* rates within each replicate we can use the `auto_rate()` function. See other [vignettes]() for more detail on using `auto_rate`. 

`auto_rate` is designed to either run every regression of a fixed width across a dataset (`"rolling"`, `"lowest"` and `"highest"` methods), or use [kernel density analysis]() to identify the most`"linear"` regions, that is regions of most consistent and stable rates. Therefore it does not have region selection functionality by time, row or oxygen as `calc_rate` [does]({#calc_rate). 

It is also likely to work poorly with intermittent-flow data like this which shows sharp, oscillating oxygen patterns. Therefore, similar to the [above example](#oxygen) with `calc_rate` oxygen regions, we need to subset each replicate before running `auto_rate` on it. 

We will use only the first subset here, and calculate two rates:

- the most `"linear"` rate of any width, which might be representative of *routine* metabolic rate

- the `"lowest"` rate across a 5 minute (300s) time window, which might be more representative of *basal* or *standard* metabolic rate

First, we `inspect` the first replicate. 

```{r results="hide"}
## subset first replicate
u_rep1 <- subset_data(urchin_int, from = 0, to = 1900, by = "time")
```

### Most linear rate {#linear}

```{r results="hide"}
# default - most linear rate
ar_lin_1 <- auto_rate(u_rep1) 
```

Depending on the data, `auto_rate` usually returns more than one result, all of which could be considered linear regions. In this example it has fit 1522 regressions and found 8 to be linear. The summary table shows these ordered from most to least linear by `density`, which is the kernel density analysis metric determining linearity. 

```{r echo = T}
summary(ar_lin_1)
```

The first result is generally the one you want. However, we would strongly encourage users to explore `auto_rate` results, especially to plot the result. As a machine-learning based process, the `auto_rate` algorithms are fallible, and on occasion return less than ideal results. Different ranked results can be plotted using the `pos` input (e.g. `plot(ar_lin_1, pos = 2)`). 

See also `subset_rate()` and `vignette("subset_rate")`. This is a function specifically designed to filter `auto_rate` results according to varous criteria, including r-squared range, duration, highest and lowest percentile of rates, and many more. 

### Lowest rate

```{r results="hide"}
# lowest rate of 300 width in the time metric
ar_low_1 <- auto_rate(u_rep1, method = "lowest", width = 300, by = "time") 
```

This method fits every regression of the specified width across the entire dataset, and orders them from lowest to highest. (Note, this ordering is by *absolute* value, that is it ignores the sign of the rate; the `"minimum"` and `"maximum"` methods order strictly numerically).

```{r}
summary(ar_low_1)
```

Again, the results can be explored using `plot`, `summary`, and `print` and the `pos` input, and the resulting object can be further filtered using `subset_rate()` - see `vignette("subset_rate")`.

## Background adjustments

For background adjustments, the `adjust_rate()` function can be used on `calc_rate` and `auto_rate` objects containing single and multiple rates. `adjust_rate` can be used to adjust rates in a variety of ways, including by a single value or dynamic background rates which change over the course of an experiment. See example in `vignette("intermittent_long")`, or `vignette("adjust_rate")` for a comprehensive guide to this function.

## Converting rates {#convert}

Once the rates have been determined, they can be converted to absolute (i.e. whole animal or whole chamber), mass-specific, or area-specific rates using `convert_rate()`. 

We will convert the [most linear](#linear) rates we determined above, but this will be the same for any of the rate objects we have saved. 

```{r tidy = F}
# absolute rate
rep1_abs_rate <- convert_rate(ar_lin_1,               # auto_rate, calc_rate, or adjust_rate object
                              o2.unit = "mg/L",       # units of the original oxygen data
                              time.unit = "s",        # units of the original time data
                              output.unit = "mg/h",   # see unit_args() for supported units
                              volume = 2.379)         # effective volume of the respirometry chamber in L

# mass-specific rate
rep1_ms_rate <- convert_rate(ar_lin_1,               
                             o2.unit = "mg/L",       
                             time.unit = "s",        
                             output.unit = "mg/h/g", # note - changed to a mass-specific unit
                             volume = 2.379,         
                             mass = 0.006955)        # mass in kg - for mass-specific rates only

rep1_abs_rate
rep1_ms_rate
```

Only the first rate converted rate is shown in the `print` output, but all rates within the object are converted. They can be viewed via `summary`.

```{r}
summary(rep1_ms_rate)
```


## Extracting rate results 

In an intermittent-flow experiment, we would now have a rate (or rates) for each replicate. 

The final rate result depends on the particular research question and many other factors. For example, if interested in basal metabolic rates perhaps the lowest rate across a fixed time period in *any* replicate, or if interested in routine metabolic rates maybe a mean rate across all replicates. What is important is that these criteria are applied consistently and reported alongside the results. 

Rates can be easily extracted from `convert_rate` objects for further analysis or selection in a variety of ways. 

You can using `$rate.output` to get the numeric values.

```{r}
rep1_ms_rate$rate.output
```

The `summary` function also allows the entire summary table to be extracted and saved as a `data.frame` with `export = TRUE`. You can also use `pos` to view or export only a selection of rows.
```{r}
rep1_df <- summary(rep1_ms_rate, export = TRUE, pos = 1:5)
```

## Advanced selection of rate results

For more advanced selection of `auto_rate` results, including selection by r-squared range, rate value or percentile, data region and many more see `subset_rate()` and `vignette("subset_rate")` for full explanation and examples. This would typically be done before passing the object to `convert_rate`.

## Longer intermittent-flow experiments

An example of analysis of a much longer, more regular intermittent-flow experiment can be seen in `vignette("intermittent_long")`. 
