---
title: "Intermittent-flow respirometry 1: Simple example"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Intermittent-flow respirometry 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE, tidy = TRUE, 
  highlight = TRUE, fig.width = 6, fig.height = 6)
```

## Intermittent-flow respirometry

Intermittent-flow respirometry involves obtaining multiple replicates in an experiment by periodically flushing the respirometer with fresh treatment water or air after a set period of time or $O_2$ decrease. Typically, the oxygen sensing system is left recording during flushes so the system returning to normoxia can be seen in the data. If the dataset is not split into the separate replicates, this results in an dataset with oscillating levels of oxygen. This vignette shows a simple example of how an intermittent-flow respirometry experiment can be analysed in `respR`. The next vignette **LINK** shows how a longer, more complex intermittent-flow experiment with differing background rates can be analysed.

### A relatively simple example

The example data, `intermittent.rd`, contains intermittent-flow measurements of oxygen consumption in the sea urchin *Heliocidaris erythrogramma*. Detailed information about the data, including its source and methods, can be obtained with the command `?intermittent.rd`. 

We use `inspect_data()` to scan for errors in the data frame, preview the data and save it to an object to pass to the next stage.

```{r} 
urchin_int <- inspect_data(intermittent.rd)
```

We can see from the top plot that this dataset has three replicates. The flush periods obviously skew the bottom plot which shows rate of change in O2 concentration, however uptake rates (flat regions) appear to be consistent within each experimental replicate. 

### Analysis with `calc_rate`
In the future `respR` will support intermittent-flow data directly with a dedicated function to easily identify and subset replicates, but at present intermittent data can be analysed quite easily via the main `calc_rate` function with a little data organisation.

#### Subsetting by row and time
The `calc_rate` function supports input of multiple subset regions by `row` or `time`. The `from` and `to` operators can be vectors with paired values:

```{r fig.width=7, fig.height=3.5}
urchin_int_rates <- calc_rate(urchin_int, from = c(200, 2300, 4100), to = c(1800, 3200, 4600), by = "time")
print(urchin_int_rates)
```

Calling `summary()` will show the coefficients and locations of each subset:
```{r}
summary(urchin_int_rates)
```

We can also use `plot()` to view the diagnostic plots of any individual replicate. The argument `rep` determines which data subset to plot:
```{r}
plot(urchin_int_rates, rep = 2)
```

We can even query multiple regions of each replicate (which can overlap, although this would obscure the output plot somewhat):
```{r fig.width=7, fig.height=3.5}
calc_rate(urchin_int, from = c(200, 900, 2300, 2900, 4000, 4400), to = c(700, 1700, 2700, 3300, 4300, 4700), by = "time")
```

#### Extracting rate results 

Rates can be easily extracted for further analysis or selection. For instance, we may be interested in the lowest rate, the one most representative of basal metabolic rates. Of course because rates are negative, the minimum *value* in negative terms would actually represent the highest rate, so we should extract the maximum (least negative) value, or transform the rates first: 
```{r}
urchin_int_rates$rate
min(-1*urchin_int_rates$rate)
```
Or we may be more interested in the mean rate:
```{r}
mean(urchin_int_rates$rate)

```

#### Subsetting by oxygen 

If we are interested in looking at rates over particular oxygen ranges within each replicate, `calc_rate` will find the first match to the range entered:
```{r fig.width=7, fig.height=3.5}
calc_rate(urchin_int, from = 7, to = 6.6, by = "o2")
```

To examine the same range in each replicate, we can follow a number of approaches. We can call the data range containing another replicate directly from the original dataset, or `inspect_data` object element:

```{r fig.keep="none"}
calc_rate(intermittent.rd[2100:3500,], from = 7, to = 6.6, by = "o2")
calc_rate(urchin_int$df[2100:3500,], from = 7, to = 6.6, by = "o2")
```
```{r echo=F, message=FALSE, warning=FALSE, results = "hide", fig.width=7, fig.height=3.5}
calc_rate(urchin_int$df[2100:3500,], from = 7, to = 6.6, by = "o2")
```

Or we can subset them manually into new dataframes, and process them separately through the entire workflow. Here we use the dedicated `subset_data` function in `respR`, which allows easy subsetting by time, oxygen or row number. For best results, these should exclude the flush periods. 
```{r fig.keep="none", message=FALSE, warning=FALSE, results="hide"}
# Create separate replicate data frames
u_rep1 <- subset_data(intermittent.rd, from = 1, to = 1900, by = "time")
u_rep2 <- subset_data(intermittent.rd, from = 2100, to = 3500, by = "time")
u_rep3 <- subset_data(intermittent.rd, from = 3700, to = 4831, by = "time")

# Process one in `inspect_data` (output not shown)
u_rep3_c <- inspect_data(u_rep3)
```
```{r fig.width=7, fig.height=3.5}
calc_rate(u_rep3_c, from = 7, to = 6.6, by = "o2")
```


### Analysis with `auto_rate()`

To more rigourously identify minimum, maximum or most linear rates we can use the `auto_rate()` function, however because this does not support (as yet) subsetting internally, each replicate must be subset and analysed separately. See other vignettes for more detail on using `auto_rate()`. 
```{r results="hide"}
## subset first replicate
u_rep1 <- subset_data(intermittent.rd, from = 0, to = 1900, by = "time")
```

```{r results="hide"}
## analyse in auto_rate
auto_rate(u_rep1) #default = most linear rate
auto_rate(u_rep1, method = "min") # minimum rate

```


### Background adjustments

For background adjustments, the `adjust_rate` function can be used on `calc_rate` objects containing multiple rate values:
```{r}
urchin_int_rates_adj <- adjust_rate(urchin_int_rates, by = -0.00005)
urchin_int_rates_adj$input.rate
urchin_int_rates_adj$corrected
```
We can see each individual rate has been adjusted, and the new rates can be extracted from the resulting object using `$corrected`. 

### Converting rates

Once the rates have been determined, they can be converted to volume-corrected or mass-specific rates using `convert_rate`. 

```{r}
u_rates_final <- convert_rate(urchin_int_rates_adj, o2.unit = "mg/L", time.unit = "s", output.unit = "mg/h/g", volume = 2.379, mass = 0.006955)
```

Only the first rate converted rate is shown in the output, but all rates determined are converted and can be extracted.
```{r}
u_rates_final$input
u_rates_final$output
```


### More complex intermittent-flow experiments

See the second vignette **LINK** for a more complex intermittent-flow analysis.
