---
title: "Intermittent-flow respirometry: Short experiment"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Intermittent-flow respirometry: Short experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 4))
```

Intermittent-flow respirometry allows multiple replicates to be conducted in an experiment by periodically flushing the respirometer with new water or air, for example after a specific period of time or oxygen decrease. 
Typically, probes are left recording during flushes, with the resulting trace showing the oxygen returning to ambient levels. 
If the data is not split into the separate replicates this results in a dataset with oscillating levels of oxygen, and generally we want to extract a rate from each descending section. This can prove tricky especially when experiments are lengthy and have many replicates.

[`respR v2.1`](https://januarharianto.github.io/respR/articles/version_notes.html) introduced new functions to analyse intermittent-flow data: 

- **`calc_rate.int()`** which runs `calc_rate()` on every replicate in an intermittent-flow dataset to extract a rate from the same time or row region of each

- **`auto_rate.int()`** which runs `auto_rate()` on every replicate in an intermittent-flow dataset to automatically extract the *most linear*, *lowest*, or *highest* rate in each

- **`select_rate()`** which works on `convert_rate()` objects and allows for advanced selection of a final rate using consistent criteria, including the results of intermittent-flow analyses

This vignette shows how to use `calc_rate.int` to analyse a fairly short intermittent-flow respirometry experiment. In `vignette("intermittent_long")` we show how `auto_rate.int` and `select_rate` can be used to analyse a much longer intermittent-flow experiment with a regular structure and increasing background rate. However, we advise reading this one first as it better covers the basics. 

## Experiment overview

The example data, `intermittent.rd`, contains intermittent-flow measurements of oxygen consumption in the sea urchin *Heliocidaris erythrogramma*. More information about the data can be viewed via `?intermittent.rd`. 

In this experiment the flushes were triggered manually so replicates are not regularly spaced and are of different durations, with the third in particular quite a bit briefer than the others. Therefore, for the following examples we need to tell the function the locations of the start of each replicate. In `vignette("intermittent_long")` you can see an example of analysis of a regular intermittent-flow experiment where we only need to specify the interval between them. 

## Inspect

The first step in any data analysis task should be to visualise and inspect the data. For this we use the `inspect()` function. This step is optional, however it scans the data for common issues, plots it, and saves it to an object to pass to further stages of the analysis. Passing all of these checks is not strictly necessary depending on the data. See `vignette("inspecting")`.

```{r eval = F}
urchin_int <- inspect(intermittent.rd)
```

```{r echo = F, warning=FALSE}
inspect(intermittent.rd)
```

```{r echo=F, fig.keep="none", message=FALSE, results="hide"}
urchin_int <- inspect(intermittent.rd)
```

We can see from the top plot that this dataset has three replicates, separated by two flushes where new water was added. The bottom plot shows rate of change (positive or negative) in oxygen over a rolling window of 10% of the data, and flush periods obviously skew this plot. Note how rates are plotted on a numerically reversed y-axis so that uptake rates are higher than production rates, or oxygen input rate in the case of the flushes (this can be changed by passing `rate.rev = FALSE`). 

This lower plot provides a quick visual inspection of how the rate varies over the course of the experiment.
Regions of stable and consistent rates can be identified on the rolling rate plots as flat
or level areas. The plot here in `inspect` is for exploratory purposes only; later functions
allow rate to be calculated over specific regions. Here, despite the flushes skewing the rate plot, specimen uptake rates (flat regions) appear to be relatively consistent within each experimental replicate with a value of around -0.005. 

For a closer look, we can use `subset_data()` and `inspect()` without saving the result to check regions of the data, which can help us decide how we are going to proceed.

```{r results='hide', message=FALSE}
intermittent.rd |> 
  subset_data(from = 1, to = 1800, by = "row") |>
  inspect()
```

Inspecting the first replicate shows us rate is indeed fairly consistent within the replicate, but fluctuates from around -0.0004 to -0.0008. However, bear in mind these rate values are determined over the default of a 10% window of the inspected data, and so are sensitive to noise. 

`inspect` however lets us pass a different `width` input for the purposes of the rolling rate plot. Let's try a wider window of 50% of the data.

```{r results='hide', message=FALSE}
intermittent.rd |> 
  subset_data(from = 1, to = 1800, by = "row") |>
  inspect(width = 0.5)
```

Now we can see rate is much more consistent, although it decreases slightly over the course of the replicate. But we can tell that we would expect our extracted rates to be around -0.0006. Note the values in the rolling rate plot are plotted against the middle of the range over which they were calculated, which is why they are centred like this. 

## `calc_rate.int` {#calc_rate.int}

Introduced in [`respR v2.1`](https://januarharianto.github.io/respR/articles/version_notes.html#version-2-1-2022-12-06), `calc_rate.int()` is a dedicated function for intermittent-flow respirometry data, in which you specify a replicate structure and extract a rate from the same data region of each replicate. This can be the same rows of each replicate or the same time window.

`calc_rate.int` uses the `starts` locations to subset each replicate from the data in `x`. The function assumes each replicate ends at the row preceding the start of the next. It then extracts a rate using the `wait` and `measure` inputs. The `by` input (`"row"` or `"time"`) determines how these other inputs are applied to the data. 

Here we will run through a `calc_rate.int` analysis of the `intermittent.rd` data. Reading `vignette("calc_rate.int")` first will be helpful in explaining the general functionality, not all of which will be covered here. 

As with `calc_rate` and most functions in `respR`, `calc_rate.int` will accept a data frame of time against oxygen as well as `inspect` objects, however we will use the `urchin_int` object we created above for the remainder of this vignette. 

### Rate from complete replicates {#complete}

The row locations of the replicates in this dataset can be found in the help file (`help("intermittent.rd")`). Here is what happens if we only specify the starting location of each using `starts`:

```{r eval = FALSE, tidy=FALSE}
calc_rate.int(urchin_int,
              starts = c(1, 2101, 3901)) 
```

```{r echo=FALSE, message=FALSE}
calc_rate.int(urchin_int,
              starts = c(1, 2101, 3901), quiet = TRUE) |>
  summary()
```

This is obviously not what we want as the flush periods for the first two replicates are included. Therefore, we need to specify a `measure` phase to exclude the flushes. With irregular replicates like this we can even specify a different `measure` phase in each, and in this case use all data except the flush. 

```{r eval = FALSE, tidy=FALSE}
calc_rate.int(urchin_int,
              starts = c(1, 2101, 3901),
              measure = c(1800, 1400, 900)) |>
  summary()
```

```{r echo=FALSE, message=FALSE}
calc_rate.int(urchin_int,
              starts = c(1, 2101, 3901), measure = c(1800, 1400, 900), quiet = TRUE) |>
  summary()
```


### Rate from specific regions

However, we usually don't want to use the entire replicate. Typically we use a `wait` phase directly after a flush in case the animal was disturbed or the oxygen probe has some lag in detecting the rapid change in oxygen concentration. We can also use the `measure` input to standardise the region from which we extract a rate. 

For example, `wait = 10, measure = 20, by = "time"` will exclude the first 10 units of time used in the original data, and calculate a rate from the following 20 time units. 

Similarly, `wait = 120, measure = 360, by = "row"` will exclude the first 120 rows of each replicate, and calculate a rate across the following 360 rows. 

For this experiment, let's say we want to exclude the first 5 minutes (300 rows) of each replicate to allow for the specimen to settle down after the flush, then calculate rates over a 15 minute period (900 rows). In these data where oxygen is recorded every second this will be `wait = 300, measure = 900`. The default `by` method is `by = "row"` so we do not need to specify this. 

```{r eval = FALSE, tidy=FALSE}
urch_rates <- calc_rate.int(urchin_int,
                            starts = c(1, 2101, 3901),
                            wait = 300,
                            measure = 900,
                            legend = TRUE) 
```

```{r echo=FALSE, message=FALSE}
urch_rates <-  calc_rate.int(urchin_int,
                             starts = c(1, 2101, 3901),
                             wait = 300,
                             measure = 900,
                             quiet = TRUE,
                             legend = TRUE) 
```

In this plot, the rate data is the yellow points, the red shaded region is the `wait` phase, and the green shaded region the `measure` phase (we added `legend = TRUE` so we can see labels for this example).

In these data of oxygen recorded once per second the time and row values are equivalent, so in the first replicate we can see the rate has been determined from timepoints 300 to 900 timepoints later (i.e. 1200). In the other replicates it is from the same equivalent time or row window. Note the third replicate is not long enough to *have* a timepoint of 1200. The default is to use the closest value which in this case is the final row of the dataset, and is acceptable here because there is no flush to worry about.

We can view the full results using `summary`:

```{r echo=TRUE, message=FALSE}
summary(urch_rates)
```

The replicate number is indicated by the `$rep` column. The `$rank` column indicates ranking or ordering of rates *within* each individual replicate. Since `calc_rate.int` only returns one rate per replicate it is not relevant here, but is useful for `auto_rate.int()` results where multiple rates per replicate may be returned. 

### Plot 

The output object can also be plotted, and if we change the `type` input we can see each rate in context of the whole dataset. For larger datasets this may be of limited utility. 

```{r message=FALSE, results='hide'}
plot(urch_rates,
     type = "full")
```

See `vignette("calc_rate.int")` for additional plotting options. 

### Adjust {#adjustcrint}

`calc_rate.int` objects can also be passed to `adjust_rate` for background adjustments. 

There is no specific background data associated with the `intermittent.rd` dataset, but we will use the `background_con.rd` data as an example. We just need to convert it to the same units as our urchin data. 

```{r message=FALSE}
bg_data <- background_con.rd
bg_data[[2]] <- convert_DO(bg_data[[2]],
                           from = "%Air",
                           to = "mg/l",
                           S = 30, t = 15)
```

Now we calculate a background rate. 

```{r message=FALSE, results='hide'}
bg_rate <- calc_rate.bg(bg_data)
```

```{r message=FALSE}
print(bg_rate)
```

And use it to adjust our urchin rates. 

```{r}
urch_rates_adj <- adjust_rate(urch_rates,
                              by = bg_rate) |>
  summary()
```

Note how the `rep` column is carried forward to the results. We can see a small adjustment to the rates in the final column. Here we are adjusting by a single value, but other adjustment methods in `adjust_rate` can be used with `calc_rate.int` objects. See `vignette("adjust_rate")` for examples, and `vignette("intermittent_long")` for how a dynamic background rate can be applied to intermittent-flow data.

### Convert {#convert_short}

Lastly, we convert the rates to units. 

```{r}
urch_rates_conv <- convert_rate(urch_rates_adj,
                                oxy.unit = "mg/l",
                                time.unit = "s",
                                output.unit = "mg/h/g",
                                mass = 0.006955,
                                volume = 2.379) |>
  summary()
```

Again note how the `rep` column is carried forward.

A final rate can be determined in a number of ways and will be specific to each experiment. As an example we can use the mean of all three replicates. 

```{r}
mean(urch_rates_conv)
```

See also `vignette("select_rate")` for more advanced rate selection. 

## `auto_rate.int`

Also introduced in [`respR v2.1`](https://januarharianto.github.io/respR/articles/version_notes.html#version-2-1-2022-12-06), `auto_rate.int()` is a dedicated function for intermittent-flow respirometry data which allows you to run the `auto_rate()` function on every replicate. You can use it to extract from each replicate the most linear (i.e. most consistently sustained) rates, the lowest or highest rates of a specific duration, or any of the other methods supported by `auto_rate`. This vignette - `vignette("auto_rate.int")` - covers the basic functionality. 

Here we'll repeat the above analysis, but this time use `auto_rate.int` to extract the most linear (i.e. most consistently maintained) rate from each replicate, which could be considered a routine metabolic rate (RMR). 

### Rate

`auto_rate.int` has a few additional inputs to consider related to the use of `auto_rate()`. See `help("auto_rate")` and `vignette("auto_rate")` for full details, but briefly these are to specify the `method` and `width`. The `by` input which controls the units of `starts`, `wait` and `measure` also applies to the `width` input. Lastly, `n` controls how many results to return from each replicate, the default being `1`.

A notable difference from `calc_rate.int` is that while you also specify a `measure` phase the `auto_rate.int` analysis occurs *within* this region, and the width of the rates (i.e. duration or rows over which they occur) may be of various lengths depending on the `method` and `width` inputs. 

Here, we again want to exclude the first 5 minutes (300 rows) of each replicate to allow for the specimen to settle down after the flush. We then calculate rates from within a different `measure` period in each replicate to use as much of the data as possible. We specify the `"linear"` method, and use a `width` of `400`, which was arrived at after some testing of different values. See [here](https://januarharianto.github.io/respR/articles/auto_rate.html#width) for discussion of appropriate widths to use with `auto_rate`.

```{r eval = FALSE, tidy=FALSE}
urch_rates <- auto_rate.int(urchin_int,
                            starts = c(1, 2101, 3901),
                            wait = 300,
                            measure = c(1500, 1100, 600),
                            method = "linear",
                            width = 400, 
                            legend = TRUE) |>
  summary()
```

```{r echo=FALSE, message=FALSE}
urch_rates <-  auto_rate.int(urchin_int,
                             starts = c(1, 2101, 3901),
                             wait = 300,
                             measure = c(1500, 1100, 600),
                             method = "linear",
                             width = 400, 
                             quiet = TRUE,
                             legend = TRUE) |>
  summary()
```

That's it. This is our RMR for each replicate. 

### Adjust

In this example, just to show a different adjustment method, let's say we had a concurrent control experiment running. We'll pretend the same background data we used [above](#adjustcrint) is from this concurrent control. Assuming that the data have the same starting timepoint, this adjustment method calculates a background rate from the exact same region in the background data that each rate was calculated from. 

```{r}
urch_rates_adj <- adjust_rate(urch_rates,
                              by = bg_rate,
                              method = "concurrent") |>
  summary()
```
We can see there has been a very small, but different, adjustment made to each rate. See [here](https://januarharianto.github.io/respR/articles/adjust_rate.html#example-2---intermittent-flow-respirometry) for a better example of this adjustment method in the context of intermittent-flow analyses. 

### Convert

Last step is to convert. 

```{r}
urch_rates_conv <- convert_rate(urch_rates_adj,
                                oxy.unit = "mg/l",
                                time.unit = "s",
                                output.unit = "mg/h/g",
                                mass = 0.006955,
                                volume = 2.379) |>
  summary()
```
This method could also be easily adapted to extract the lowest or highest rates of a specific duration from each replicate. 

## Alternative approaches

Previous versions of this vignette demonstrated use of different methods, such as looping and subsetting of data, to extract individual replicates for processing through `calc_rate` or `auto_rate`. These approaches still work and are archived [here](https://januarharianto.github.io/respR/articles/archive/intermittent_old.html) in case they are of use in some cases. However for most analyses use of `calc_rate.int` or `auto_rate.int` is the recommended approach.

