---
title: "Oxygen production respirometry"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Oxygen production respirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 3))
```

## Introduction {#intro}

While the majority of respirometry studies are concerned with oxygen uptake by various organisms, there are many studies of organisms such as plants, algae and corals where oxygen production rates are of primary interest. We have designed `respR` to be able to process and calculate rates from oxygen production data in exactly the same way as for consumption data. In fact, the package has already [been used]() by several studies to determine the production rates of diatoms, algae, and corals. 

Here we will use the included `algae.rd` dataset to briefly show how production data may be processed and used to extract production rates. It contains 20 hours of data recorded once per minute on oxygen production by an algae. More information on the data can be seen by running `?algae.rd`.

## Inspect the data {#inspect}

Like all analyses in `respR` the first step is to inspect the data for common issues.   

```{r}
inspect(algae.rd, rate.rev = FALSE)
```

In `respR` oxygen uptake rates are represented by negative values since they represent a negative slope of oxygen against time. Since the focus of the package is on uptake rates, by default the `inspect()` function plots rolling rates on a reversed axis so that higher oxygen uptake rates appear higher on the plot. To examine production rates, which are positive, we pass `rate.rev = FALSE` to reverse this behaviour 

We can see the data passes most of the checks. The warning about uneven time can be safely disregarded as it stems from the per minute recording interval being converted to decimalised hours, and we can see the minimum and maximum intervals indicate there are no major gaps. 

We can see that at the default rolling rate `width` input of 10% of the data the rate is very variable, from zero to over 0.15. However, there are characteristics of the raw data which could cause this; it has a fairly shallow overall slope and high relative noise level, therefore rates over narrow time windows would be expected to be variable. Instead we can pick a more appropriate width that will give a better idea of the true rate. 

```{r results = "hide"}
algae_insp <- inspect(algae.rd, width = 0.5, rate.rev = FALSE)
```

We can see now that rate is actually highly stable across the dataset, and this tells us when we come to calculate rates we should probably use rates determined over this width or higher, and the value will be around 0.09 to 0.10. 

## Calculate rates 

Since the rate appears to be fairly stable across the dataset we will use `auto_rate()` to identify the most linear region.

```{r}
algae_rate <- auto_rate(algae_insp, rate.rev = FALSE) 
```

This is a good opportunity to explain how the `auto_rate` `"linear"` method works and how to interpret the outputs. Briefly (see `vignette("auto_rate")` for more detail), the linear method uses the input `width` as a starting seed value to calculate a rolling rate across the dataset (seen in panel 3). It then uses these rates to identify linear regions using kernel density estimation (KDE, see panel 6). Peaks in this plot represent linear regions, that is areas of stable rates at that `width` representative of that region as a whole. It then re-samples these regions and runs additional linear analysis at different widths to arrive at a final rate for this region. This is why the final, high ranked rates tend to be over widths greater than the input `width`, as can be seen here with the top ranked result. 

Here, we have a very good top ranked result; it is calculated over a large part of the dataset, is around the value we expected from inspecting it earlier, this value is in the middle of the range of rolling rates at the default input width (dashed line, panel 3), and has a high r-squared in comparison to the other results. 

Generally, the higher and wider the peak in the KDE, the more linear the region. We can see three peaks in the KDE (panel 6). However, the other two are very much lower and narrower. Let's look at the second ranked result using `pos`. 

```{r results='hide'}
plot(algae_rate, pos = 2, rate.rev = FALSE) 
```

Here we can see this rate is over a much narrower region towards the end of the data, the value of 0.05 is very much lower than we are expecting, and at the extremes of the range (panel 3). While this may be a linear region within the data, it is very much not representative of the data as a whole. 

In this analysis, we can easily see that the top ranked result is the one we want, and we can exclude the other two. See `vignette("subset_rate")` for more advanced filtering of `auto_rate` results by many different criteria. 

### Rolling rate widths in `auto_rate`

We saw in the [inspect section](#inspect) above a higher `width` made the rolling rate much more consistent. It is not necessarily always the case increasing the default of `width = 0.2` will improve the results of `auto_rate` however, as it gives the KDE much less variation to work with. 

Here is the summary table of the results when we increase the width to 0.5.

```{r}
auto_rate(algae_insp, width = 0.5, rate.rev = FALSE)$summary
```

This doesn't really improve our analysis very much, and could be more confusing since there are more results. There is still one central peak in the KDE analysis, which uses more of the dataset, but it is largely the same as the top ranked result in the earlier analysis, differing in rate value by only 1-2%. The other results are more consistent in their rate values, but they don't add much to our understanding of what our final rate should be. Therefore, in this case the default `width` provided us with a good result, and there is no real reason to change it. This does however serve as a good example of how users should experiment with different widths to get an idea of how it might affect analyses. Certain datasets will benefit from this more than others. 

## Adjust rate

Oxygen production rates can be adjusted in the exact same way as uptake rates. In most oxygen production respirometry, background is likely similar to that in uptake respirometry, that is microbial organisms are consuming oxygen. However there may be cases where some of the production could be contributed by other organisms, such as another algae growing on the surfaces of the respirometer. What is important is to quantify this by running blank experiments which are identical in every way except for the presence of the specimen you are interested in. See `vignette("adjust_rate")` for details of the numerous ways experiments can be background adjusted, most of which is equally relevant to oxygen production respirometry.

Here, we don't have a control experiment to work with, so we will just invent some values. The most important aspect here is to be careful with the ***signs*** of the rates we are using; production rates are positive, uptake rates negative. 

This is how you would adjust for a oxygen *uptake* by micro-organisms. The uptake background is negative, which means the production rates have been *under*-estimated.

```{r}
algae_rate_adj <- adjust_rate(algae_rate, by = -0.007)  
summary(algae_rate_adj)
```

If however you find the control has a background *input* of oxygen it is entered as a positive value, in which case the specimed production rates have been *over-*estimated.

```{r}
adjust_rate(algae_rate, by = 0.009) |>
  summary()
```

See also [Case 9]() in `vignette("adjust_rate")` where we perform a similar adjustment to production rates. 

## Convert rate

Production rates are converted to units in exactly the same way as uptake rates. 

New in `respR` v2.0 is the ability to output rates as surface-area specific, which is often used to express oxygen production of algae and corals per unit area, so we'll add an `area` input and specify an area-specific `output.unit`. 

The `area` input must be in units of m^2. Here we can also use the new `convert_val()` function to enter the `area` input. This function converts between units of area, mass, etc. It is particularly handy in `convert_rate` because the default output unit is m^2. Let's say our algae is 10 cm^2. We don't have to do this conversion ourselves, but can rely on `convert_val`. We don't even need to tell it this is an area conversion - it detects this automatically from the input units. 

```{r tidy = FALSE}
algae_rate_conv <- convert_rate(algae_rate_adj, 
                                oxy.unit = "%Air", 
                                time.unit = "hr", 
                                output.unit = "mg/h/m2",
                                area = convert_val(10, "cm2"),
                                volume = 0.1,
                                t = 12, S = 30, P =1.01)
```

In this case, we were only interested in the top ranked rate result. We can extract it using `summary` and the `pos` and `export` inputs. We can see the area has been converted to m^2. 

```{r}
summary(algae_rate_conv, pos = 1, export = TRUE)
```
