---
title: "Critical oxygen values (PCrit)"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Critical oxygen values (PCrit)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
zotero: true
bibliography: references_oxycrit.bib
csl: resprcsl.csl
link-citations: true
nocite: |
  @dugglebyRegressionAnalysisNonlinear1984a; @leivaPhysiologicalResponsesGhost2015; @lightonThermolimitRespirometryObjective2004; @mangumResponsesAquaticInvertebrates1973a; @marshallEstimatingPhysiologicalTolerances2013; @monteiroCardiorespiratoryResponsesGraded2013; @muellerRegulationIndexNew2011; @muggeoModelingTemperatureEffects2008; @negreteMethodologicalEvaluationDetermination2019b; @nickersonEstimatingPhysiologicalThresholds1989; @reemeyerStandardizingDeterminationInterpretation2019a; @reganDonThrowFish2019; @reganRatesHypoxiaInduction2017; @rodgersDoubleWhammyNitrate2021a; @rogersNewAnalysisHypoxia2016a; @seibelOxygenSupplyCapacity2020; @seibelOxygenSupplyCapacity2021; @stinchcombeGeneticsEvolutionFunctionvalued2012; @tangRateOxygenConsumption1933; @ultschUtilityDeterminationCrit2019; @vanwinkleOxyconformersOxyregulatorsQuantitative1975; @woodFallacyCritAre2018; @yeagerPhysiologicalRegulationConformation1989; @snyderEffectClosedIntermittentflow2016
---

```{r setup, echo = F, message=FALSE}
library(knitr) # load knitr to enable options
library(respR) # load respR
library(rMR) 

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = FALSE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                                #scipen = 999, 
                                digits = 5))
```

## Introduction

Some species are *oxyconformers*, with their routine oxygen uptake rate directly proportional to the oxygen supply available from the environment. So, for example, at 50% oxygen their uptake rate will be half that at 100% oxygen. Many species however are *oxyregulators*, and are able to regulate uptake and maintain routine rates as oxygen supply declines until a value is reached below which  uptake precipitously declines. 


The oxygen value below which routine uptake becomes unsustainable is termed the critical oxygen value (COV), or more commonly $P_{crit}$. Historically, this was expressed in units of partial pressure of oxygen (e.g. kPa), hence $P_{crit}$ as in the critical *pressure* of oxygen, but it is also commonly expressed in units of concentration. Here we will use the term COV to describe the value regardless of units. 

COV is typically determined in long duration, closed-chamber respirometry experiments in which the specimen is allowed to deplete the oxygen in the chamber, with the resulting oxygen trace used to identify the breakpoint in the relationship of uptake rate to oxygen concentration.

Note that the classification of species into strict oxyconformers and oxyregulators is somewhat simplistic; there can be a continuum of responses between these extremes, and many intermediate cases [@muellerRegulationIndexNew2011].

## `oxy_crit`

`oxy_crit()` is the `respR` function for determining critical oxygen values. It accepts two forms of data input.

The first is the general structure that all other `respR` functions accept, a `data.frame` or `inspect` object containing paired values of time and oxygen pressure or concentration. For data frame inputs, if time and oxygen are not the first and second columns respectively, the columns can be specified with the `time` and `oxygen` inputs. For `inspect` objects, these will have been specified already. These `oxygen~time` data are used to calculate a rolling rate of the specified `width`, the default being 0.1 or 10% of the width of the entire dataset. This rolling rate, similar to the lower panel in the `inspect` plot in the example [below](#inspect), is the data upon which the critical value analysis is conducted.

The second data input option is to enter a `data.frame` containing already calculated rates, paired with relevant oxygen values. Typically these would be the mean or central oxygen value of the range over which the rate was determined. These columns can be specified using the `rate` and `oxygen` inputs. In this case, the function does not calculate a rolling rate internally, and the critical point analysis is performed on the input data directly.

### Methods

The `oxy_crit()` function currently provides two methods to detect the breakpoint in the `rate~oxygen` relationship. With high-resolution, relatively non-noisy data these typically return similar results, but results may vary depending on the characteristics of the data.

#### Broken-stick

`method = "bsr"`

The first method is the "broken-stick" regression (BSR) approach adapted from @yeagerPhysiologicalRegulationConformation1989, in which two segments of the data are iteratively fitted and the intersection with the smallest sum of the residual sum of squares between the two linear models is the estimated critical point. Two slightly different ways of reporting this breakpoint are detailed by Yeager & Ultsch; the *intercept* and *midpoint*. These are usually close in value, and `oxy_crit` returns both.

#### Segmented {#seg}

`method = "segmented"`

The second method is a wrapper for the "segmented regression" approach available as part of the [`segmented`](https://cran.r-project.org/web/packages/segmented/index.html) R package by @muggeoModelingTemperatureEffects2008. This estimates the critical point by iteratively fitting two intersecting models and selecting the point that minimises the "gap" between the fitted lines.

### Additional inputs

The function has the following additional inputs:

#### `width`

For data entered as `time` and `oxygen` values, this controls the width of the internally calculated rolling regression which provides the rates upon which the critical point analysis is conducted. The default is `0.1`, representing 10% of the length of the dataset and seems to perform well with most data we have tested it with. However, the `width` should be chosen carefully after experimenting with different values and examining the results. Too low a value and the rolling rate will be noisy making it more difficult to detect a critical breakpoint, too high and it will be oversmoothed. See [here](https://januarharianto.github.io/respR/articles/auto_rate.html#width) for a discussion of appropriate widths and the implications of overfitting rolling rates, and also [Prinzing et al. 2021](https://januarharianto.github.io/respR/articles/refs.html#references) for an excellent discussion of appropriate widths in rolling regressions. This is an important analysis parameter with major implications for the output values and should be reported alongside the results.

#### `thin`

This affects the `"bsr"` method only, which is computationally intensive, and thins large datasets which take a prohibitively long time to process to more manageable lengths. This is entered as an integer value, with the default being `thin = 5000`, and the data frame will be uniformly subsampled to this number of rows before analysis. The default value of 5000 has in testing provided a good balance between speed and results accuracy and repeatability. However, results may vary with different datasets, so users should experiment with varying the value. To perform no subsampling and use the entire dataset enter `thin = NULL`. As an example of the time difference, using the `squid.rd` data this increases the processing time from \~6s to more than 60s, but outputs an identical result. It has no effect on datasets shorter than the `thin` input, or on the `"segmented"` method. 

#### `parallel`

Default is `FALSE`. This controls the use of parallel processing in running the analysis. If your dataset is particularly large or high resolution and analyses are taking a prohibitively long time to process try changing this to `TRUE`. Note, this option sometimes causes errors depending on the system or platform, so should only be used if it is really needed.

#### `plot`

Default is `TRUE`. Controls if a plot is produced.

## Determine COV from oxygen~time data

Here we will run through an example critical oxygen value analysis of a dataset included with `respR`.

### Data

The example data, `squid.rd`, contains data from an experiment on the market squid, *Doryteuthis opalescens*. More information about the data, including its source and methods, can be obtained with `?squid.rd`.

```{r}
squid.rd
```

### Inspecting data {#inspect}

We can visualise and examine the dataset using the `inspect()` function.

```{r}
squid <- inspect(squid.rd)
```

Note how the date are plotted against both time (bottom blue axis) and row index (top red axis), which in these data, recordings of oxygen once per second, happen to be equivalent. We can see from the bottom plot, a rolling rate across 10% of the total data, that the uptake rate is relatively consistent to around row 12,000, after which it declines steadily. We cannot easily tell from these plots what oxygen concentration this occurs at. For now, we can see the dataset is free of errors or issues, so we can pass the saved `squid` object to the dedicated `oxy_crit()` function.

### Critical oxygen value analyses

We will process the squid data using both methods and compare the results.

### Broken-stick method {#bsr}

This is the default method, so does not need to be explicitly specified with the `method` input, and we will let the default `width = 0.1` be applied.

```{r results='hide', fig.keep='none'}
squid.bsr <- oxy_crit(squid)
```

```{r, echo = FALSE}
print(squid.bsr)
plot(squid.bsr, quiet = TRUE)
```

The output figure shows the two different BSR results, the *intercept* and *midpoint* critical oxygen values, indicated by horizontal lines against the original timeseries data and vertical lines on a `rate~oxygen` plot. In this case, the lines overlay each other, and the `print()` output shows that, in these data, both methods give virtually identical values for the COV: 2.61 mg L^-1^. Note that for some data, depending on various characteristics, such as noise, abruptness of the break, etc., the two different BSR methods may provide different results.

#### Summary results

Full analysis results can be seen using `summary()`.

```{r}
summary(squid.bsr)
```

#### Changing the width

Let's change the `width` input to see how it affects the results. We'll try smaller and larger `width` values. We can also use `panel` to output only the rolling rate plot.

```{r message=FALSE, results='hide', fig.height=3}
oxy_crit(squid, width = 0.05, panel = 2)
oxy_crit(squid, width = 0.2, panel = 2)
```

Note how the rolling rate plot with the higher width is much smoother, with a less pronounced breakpoint. Increasing the width by too much can smooth out critical breakpoints, which is the very thing that the function is trying to detect. We can see the smaller `width` gives a similar result as the default value of `0.1`, while the larger `width` estimates the breakpoints as a slightly higher value. Therefore, in this case the default value of `0.1` seems to be appropriate and give reliable results. 

When running COV analyses, users should experiment with different widths, choose them appropriately and objectively, and report them in the analysis methods alongside results.

### Segmented method {#seg}

Now we'll run the analysis using the `"segmented"` method, again with the default `width = 0.1`.

```{r results='hide', fig.keep='none'}
squid.seg <- oxy_crit(squid, method = "segmented")
```

```{r, echo = FALSE}
print(squid.seg)
plot(squid.seg, quiet = TRUE)
```

For these particular data, we get the exact same result as the BSR method: 2.61 mg L^-1^. This will not be the case with every dataset.

#### Summary results

Full analysis results can be seen using `summary()`.

```{r}
summary(squid.seg)
```

#### Changing the width

Again, let's try different `width` inputs.

```{r message=FALSE, results='hide', fig.height=3}
oxy_crit(squid, width = 0.05, method = "segmented", panel = 2)
oxy_crit(squid, width = 0.2, method = "segmented", panel = 2)
```

We can see again we get the same values with these different widths as in the BSR method above, though this won't necessarily be the case with every dataset. The higher width tends to oversmooth the rolling rate, and gives a slightly higher COV. 

## Determine COV from rate~oxygen data

The example above used raw `oxygen~time` data, and so the function calculated a rolling rate internally, then performed the breakpoint analysis on these data. However, the function can accept already calculated `rate~oxygen` data, and the function will perform the analysis using these data directly. These rates can be either absolute rates (i.e. whole specimen or whole chamber), or mass-specific rates; both will give identical critical value results. The rates can also be unitless rates, that is the rate values output by `respR` functions such as `calc_rate` and `auto_rate` before conversion to proper oxygen rate units, as well as rates already converted to units. As long as all rates are dimensionally equivalent the critical oxygen analysis will output identical results.

Here, we'll use the `"rolling"` method in `auto_rate()` to perform a rolling regression on the squid data to get a rolling rate, and we'll convert these to a mass-specific rate in `mlO2/h/g`. We'll pair these values in a data frame with a rolling mean of the oxygen value from the same window using the [`roll`](https://cran.r-project.org/web/packages/roll/index.html) package.

This new dataset will be shorter by the input `width` setting, because the rolling methods start at row `width/2` and run to `width/2` before the final row (these functions can output results of the same length, but it necessarily involves using partial windows at the start and end of the data, which can result in questionable outputs).

```{r}
## Perform rolling rate analysis 
squid_ar <- auto_rate(squid.rd, method = "rolling", width = 0.1, plot = FALSE)
## Convert rates
squid_conv <- convert_rate(squid_ar,
                           time.unit = "sec",
                           oxy.unit = "mg/L",
                           output.unit = "ml/h/g",
                           volume = 12.3,
                           mass = 0.02141,
                           t = 15,
                           S = 30,
                           P = 1.01)
## Extract rates
rate <- squid_conv$rate.output

## Rolling mean of oxygen value
oxy <- na.omit(roll::roll_mean(squid.rd[[2]], width = 0.1 * nrow(squid.rd)))
  
## Combine to data.frame
squid_oxy_rate <- data.frame(oxy,
                             rate)
```

Now we run the `oxy_crit` analysis. We use the `oxygen` and `rate` inputs to specify the columns.

```{r}
oxy_crit(squid_oxy_rate, oxygen = 1, rate = 2)
```

Notice that we get the same critical values as we got with the analysis on `oxygen~time` data [above](#bsr), despite the very different rate values (y-axis) and the fact these are *mass-specific* rates. Both absolute and mass-specific rates will give identical results.

### Extracting results

Results of `oxy_crit` analyses can be extracted from the results object. The critical oxygen value is in the `$crit` element. If the `"bsr"` method has been used it will contain both results, if the `"segmented"` method it is a single value.

```{r}
## bsr results
squid.bsr$crit

## segmented result
squid.seg$crit
```

Summary results can also be exported to a `data.frame` by using the `summary()` function and `export = TRUE`.

```{r results='hide'}
## bsr
bsr_res <- summary(squid.bsr, export = TRUE)
```

```{r echo=FALSE}
bsr_res
```

```{r results='hide'}
## seg
seg_res <- summary(squid.seg, export = TRUE)
```

```{r echo=FALSE}
seg_res
```

## Which method to use

With high resolution, non-noisy data the BSR and Segmented methods identify very similar, if not identical, critical oxygen values (COVs). This will vary with other data, and one of the methods might work better than the other depending on the characteristics of the data. Which to report should be informed after familiarisation with how they work and testing with the data. Most importantly results should be reported alongside analysis parameters which affect the outputs, such as the width of the rolling regression.

There are additional methods to identify critical breakpoints in oxygen\~time data, such as the non-linear regression method of @marshallEstimatingPhysiologicalTolerances2013, and the α-method of @seibelOxygenSupplyCapacity2021. It is our plans to support these and others in updates to `oxy_crit`, as well as metrics which describe the degree of oxyregulation of a specimen in intermediate cases e.g. [@tangRateOxygenConsumption1933; @muellerRegulationIndexNew2011].

There is a huge literature (see [below](#refs)) and healthy debate about the use and application of $P_{crit}$, much of it very recent. Before running any COV analysis you should familiarise yourself with the literature, particularly critiques of the different methods, and their utility.

We particularly recommend the recent debate by @woodFallacyCritAre2018 and @reganDonThrowFish2019, and also @ultschUtilityDeterminationCrit2019. Also be sure to check out the latest developments by @seibelOxygenSupplyCapacity2021.

An important point to note is that $P_{crit}$ and COVs are most frequently used as a *comparative* metric. Since analytical options chosen by the investigator such as regression width inherently affect the result, it is arguably more important that these are kept the same amongst analyses that will be the basis of comparisons, rather than consideration of the ultimate values *per se*. Therefore, it is important that investigators fully report the parameters under which these analyses have been conducted. This allows editors and reviewers to reproduce and assess analyses, and subsequent investigators to judge if comparisons to their own results are appropriate. `respR` has been designed to make the process of reporting these analyses straightforward - see `vignette("reproducibility")`.

## References {#refs}
