---
author: "Januar Harianto"
title: "respR Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pkgdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE, tidy = TRUE, 
  highlight = TRUE, fig.width = 6, fig.height = 6)
```

# Introduction to respR

When working with aquatic respirometry data, you will often need to:

1. Ensure that the data, or at least a subset of the data, is representative of the research question of interest.
2. Perform an initial analysis of the data to estimate the rate of change in oxygen concentration or amount.
3. Depending on the experimental setup, correct for background usage of oxygen by micro-organisms, or correct for oxygen flux from the air.
4. Scale the resulting usage rate to the volume (of the medium) and mass (of the animal), and convert to appropriate units.

The `respR` package contains several functions to make these actions quick and easy:

- It provides visual feedbacks and diagnostic plots to help you subset and analyse your data.
- It uses computational techniques such as rolling regressions and kernel density estimates to estimate ranked maximum, minimum or best fit sections of the data.
- By separating the workflow into a series of connected functions, you can "mix and match" functions to help you achieve your result.

# Example Data

We have provided example data that can be used immediately once `respR` is loaded. The data are `urchins.rd`, `intermittent.rd`, `sardine.rd` and `squid.rd`. These data were obtained from past experiments and more information can be obtained by invoking the `?` command in the R console, for instance, `?urchins.rd`. 

# The typical respR workflow

`respR` was designed to work best with data frame objects. This includes objects of class `data.frame`, `data.table` and `tibble`. The user should already know how to import data into these formats as they are fundemental data structures in R. Here we describe a typical workflow in the analysis of respirometry data once it has been loaded into R. The example data used is `urchins.rd`, where the first column of the data frame is time data, while the remaining 18 columns are dissolved $O_2$ data. Columns `b1` and `b2` contain background respiration.

```{r}
head(urchins.rd)
```

## Step 1: Check for common errors

We first use `inspect_data()` on the data frame to scan for:

- Missing/non numeric (`NA`/`NaN`) data;
- sequential time data;
- duplicates in time data; and
- evenly-spaced time data.

By default, the function always assumes the first column (`xcol`) is time data, while the second column (`ycol`) is $O_2$ data. In the case of `urchins.rd` where a multi-column dataset is provided, the function also defaults to the processing of just the first 2 columns. You can use the `xcol` and `ycol` arguments to modify that behaviour.

```{r}
urchin <- inspect_data(urchins.rd, xcol =  1, ycol = 15)
```

From the plot, we can see irregularities in these data near the end of the timeseries (in this case the sea urchin had interfered with the oxygen sensor). A linear regression of the entire data series would therefore give an erroneous calculation of the true rate. The function also warns us that time data is not evenly-spaced. We have several options to manage non-linear and irregular timeseries data. For now, the data frame is saved as an object, `urchin`.

**It should be noted that using `inspect_data` is optional** - the main functions in our package will readily accept any data frame as long as data are all numeric and error-free. Running the function is a qualitative step that simply flags potential issues about the data before it is analysed.

## Step 2: Process background respiration

The presence of microorganisms in the respirometry medium may be a potential source of significant experimental bias, and we may want to account for background rates during experiments. Since background rates typically account for a small percentage of experimental rates, these often-called “blank” experiments are routinely conducted alongside, or before and after main experiments, and the rates are averaged across several datasets to obtain a more accurate estimate of the correction.

The function `calc_rate.bg()` can be used to simultaneously process multiple background rate measurements as long as they share the time data. In `urchins.rd`, background respiration was recorded and saved in columns 18 and 19. We analyse the data using `calc_rate.bg()` and save the output as an object.

```{r}
# Note that the data is truncated by time, from 5 to 40 minutes 
#   (instead of 0 to 45).
bg <- calc_rate.bg(urchins.rd, xcol = 1, ycol = 18:19, from = 5, to = 40, 
  by = "time")
print(bg)
```


## Step 3: Estimate rate

Calling the function `calc_rate()` on your data, with no additional arguments, will prompt the function to perform a linear regression on the entire data. The function will automatically consider the first column as time data, and the second column as dissolved $O_2$ data.

```{r}
calc_rate(urchin) # same as: calc_rate(urchin$df)
```

Note how the function can recognise and extract data from the output object of `inspect_data()`. Alternatively, you can also use a `data.frame` object to run other analyses using the same function. 

In many cases, there is a need to truncate or subset the data before rate is determined. For example, we may want to determine rate over an exact period of time, or within a threshold of O~2~ concentrations. Equipment interference or other factors may also cause irregularities or “spikes” in the data. We can work around the error(s) and subset the regions that are not erroneous and still obtain valid results.

Based on the `from` and `to` arguments, a user may use `calc_rate()` to truncate data in any of 4 ways:

1. **Time period** (`by = "time"`) - *"What is the average rate over a 25 minute period?"*
2. **Total oxygen consumed/produced** (`by = "o2"`) - *"At what rate is oxygen consumed between saturation points of 95% and 80%?"*
3. **Proportion based on total oxygen consumed** (`by = "proportion"`) - *"What is the rate from halfway down the data?"*
4. **Precise subsetting by row** (`by = "row"`). - *"I'd like to subset between rows 11 and 273."*

Note that for values of O~2~ and time that do not match any value in the data, the function will identify the closest matching time or O~2~ values to the inputs, rounded down, and uses these for subsequent calculations.

```{r}
rate <- calc_rate(urchin, from = 4, to = 29, by = "time")
rate # print(rate)
```

Calling `summary()` will produce a `data.frame` object that can be saved directly into a file.
```{r}
# write.csv(summary(rate), file = "results.csv")
summary(rate)
```

Plotting the output provides a series of diagnostic plots of the data subset that was analysed.
```{r}
plot(rate)
```

## Step 4: Adjust for background respiration

Since background rate has been calculated in `calc_rate.bg()`, adjustment is straightforward:

```{r}
a.rate <- adjust_rate(rate, bg)
a.rate
```


## Step 5: Convert the results

Once the rate of change of oxygen has been determined, we may either want to calculate:

1. Total change in O~2~ per unit time; or
2. Mass-specific rate of change in O~2~.

The function `convert_rate()` can be used to scale rate values to chamber volume and/or specimen mass. For example, the user may convert the output of `calc_rate()` to O~2~ consumed over an hour:

```{r}
convert_rate(a.rate, o2.unit = "mg/L", time.unit = "sec", output.unit = "mg/s", 
  volume = 1.09)
```

Or, the user may convert a known rate to a volume-corrected, mass-specific rate:

```{r}
convert_rate(a.rate, o2.unit = "mgl-1", time.unit = "s", output.unit = "mg/s/kg",
  volume = 1.09, mass = 0.19)
```

A string matching algorithm is used to automatically recognise variations in base units. For example, `"ml/s"`, `"mL/sec"`, `"milliliter/s"`, and `"millilitre/second"` are all equally identified as `mL/s`. Unit delimiters can be any combination of a space, dot (`.`), forward-slash (`/`), or the "per" unit (`-1`). Thus, `"ml/kg"`, `"mL / kg"`, `"mL /kilogram"`, `"ml kg-1"` or `"ml.kg-1"` are equally recognised as `mL/kg`.

Due to the "fuzzy" approach in the matching algorithm, writing unit strings for `convert_rate()` should come naturally and intuitively for most people. As long as the units are recognised by the function, the conversion will take place. If we need a reminder on what units are available to use, we can call `unit_args()`:

```{r}
unit_args()
```
