---
title: "Thin larger datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thin larger datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F, warning = F, message = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(
                 scipen = 999, 
                 digits = 3))
```

Another useful utility function in `respR` in `subsample()`. This is intended to thin or reduce the resolution of large datasets so analysis processing times are reduced, which may be useful for computationally intensive functions such as `auto_rate()` and `oxy_crit()`.

`respR` has been designed to be extremely fast in running analyses of large datasets, and as computing power increases over time, there is rarely any need to do this, and caution would suggest datasets are kept complete where possible. We have found it useful primarily to reduce data sizes to use with other [packages](https://januarharianto.github.io/respR/articles/archive/auto_rate_comp.html) which take a prohibitively long time to process data, so their results can be compared with those of `respR`.

`subsample` has two ways of selecting the degree of subsampling. The `n` input will cause every n'th element or row to be subsampled. The probably more useful `length.out` input uniformly subsamples the data to this total length or number of rows. 

The function works with both vectors and data frames, and will output an object of the same class. A plot is produced to you can check the subsample is still representative of the original, or it can be suppressed with `plot = FALSE`.
 
## Subsample by every n'th element
 
```{r}
#' # Subsample by every 200th row:
subsample(squid.rd, n = 200, plot = FALSE)
```

## Subsample to a specific length

```{r}
#' # Subsample to 100 rows:
subsample(sardine.rd, length.out = 100)
```

## Subsample a vector

```{r}
subsample(sardine.rd[[2]], length.out = 500)
```
