---
title: "respR and the Tidyverse"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{respR and the Tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE, tidy = TRUE, 
  highlight = TRUE)
```

## Working in the Tidyverse

`respR` integrates nicely with the [tidyverse](https://www.tidyverse.org/), specifically with `dplyr` functions e.g. `select()`, `filter()` and `mutate()`, and `magrittr` pipe operators ("`%>%`") to clearly express workflows in an organised sequence. 

It also works with the new native pipe operator (`|>`) introduced in [R v4.1](https://www.r-bloggers.com/2021/05/new-features-in-r-4-1-0/), however the `dpylr` pipes have some additional functionality. For more information, see the [**Pipes**](http://r4ds.had.co.nz/pipes.html) chapter in the online *R for Data Science* book.

## Examples

Here we show how using `%>%` pipes can make data analysis worklows simpler for the user.

Typical analysis using regular `R` syntax:
```{r, eval=FALSE}
# 1. check data for errors, select cols 1 and 15:
urch <- inspect(urchins.rd, 1, 15) 
# 2. automatically determine linear segment:
rate <- auto_rate(urch)
# 3. convert units
out <- convert_rate(rate, "mg/l", "s", "mg/h/kg", 0.6, 0.4)
```

Alternatively, use `tidyverse` pipes:
```{r, eval=FALSE}
urchins.rd %>%        # using the urchins dataset,
  select(1, 15) %>%   # select columns 1 and 15
  inspect()     %>%   # inspect the data, then
  auto_rate()   %>%   # automatically determine most linear segment
  convert_rate("mg/l", "s", "mg/h/kg", 0.6, 0.4) # convert units

```
