---
title: "respR and the Tidyverse"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{respR and the Tidyverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE, tidy = TRUE, 
  highlight = TRUE, fig.width = 6, fig.height = 6)
```

### Working in the Tidyverse

`respR` integrates nicely in the [Tidyverse](https://www.tidyverse.org/), specifically with `dplyr` functions e.g. `select()`, `filter()` and `mutate()`, and `magrittr` pipe operators ("`%>%`") to clearly express workflows in an organised sequence. For more information about using pipes in particular, see the ["Pipes" chapter](http://r4ds.had.co.nz/pipes.html) in the online R for Data Science book.

We load the data, `sardine.rd`, which contains approximately 2.1 hours (7,513 data points) of a single respirometry experiment.

```{r}
library(respR)
data("sardine.rd")
?sardine.rd # run this to see notes about the dataset
```

