---
title: "calc_rate: Two-point rate calculation"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{calc_rate: Two-point rate calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 6, 
               fig.height = 4,
               fig.align = "center")
```

## Introduction

Best practice in respirometry is usually to use linear regression on timeseries recordings of oxygen concentrations to derive a slope and thus rate of oxygen use. 

Technically, an overall rate between any two timepoints can be calculated using only two oxygen values if they are accurate recordings, and in the past "*closed-bottle*" respirometry was common, where only the initial and final oxygen concentrations were measured. In the modern day, there is little excuse for not using continuous data recordings, when it is relatively cheap and easy to do so. 
Using only two values gives no information about how rates of use may have fluctuated, and are also obviously highly sensitive to error and inaccurate readings. 

For some applications however, there may be valid reasons for using only two datapoints of oxygen concentration separated by a known time period to calculate a rate. 
This could be in cases of periodic and irregular oxygen monitoring or sampling in the field, or alternatively where the investigator is not interested in oxygen uptake or productions rates *per se*, but rather in total oxygen flux over a time period. 
An example of this is in respirometry studies of sediment communities, where oxygen flux is correlated with transport of other chemicals or nutrients (e.g. [Smith Jr. 1987]()). 
In such cases, even if continuous data exist, fitting a linear regression to these data may not be appropriate, and in fact could be grossly incorrect. 
This is particularly the case where oxygen or oxygen use fluctuates significantly over the time period of the experiment, such as with a diurnal rhythm.
In such cases linear regressions may be significantly skewed.  

`respR` includes the ability to calculate two-point rates. 
While we recommend it not be used for analysis of data from any laboratory based studies looking at organismal metabolic rates, we have incorporated the functionality for the above use-cases, and for analysis of historical closed-bottle respirometry data. 

## Example

### Inspecting data
This dataset contain recordings of oxygen from an eutrophic lake over a period of 40 hours starting at midnight. 

```{r echo = F}
lake <- read.csv("lake.csv")
lake[[1]] <- round(lake[[1]], 2)
names(lake) <- c("hours", "oxygen")
```

```{r warning=FALSE, message=FALSE, results='hide'}
lake_insp <- inspect(lake)
```

We can see the oxygen in the lake fluctuates with a diurnal rhythm, increasing over the day and decreasing at night. The rolling rate similarly fluctuates, and because of this there are some strage artefacts in the calculated rates. 

### Calculating rate

The researcher wants to know the overall rate of oxygen flux between 12 noon on the first day (hour 12) and 12 noon on the next (hour 36). We will use `calc_rate` as we normally would to get the rate between these two timepoints. 

```{r tidy = F, warning=FALSE, message=FALSE, results='hide'}
lake_rate <- calc_rate(lake_insp, 
                       from = 12,
                       to = 36,
                       by = "time")
```

It is immediately clear the linear regression based rate completely misrepresents the overall oxygen flux and instead of a small or even negligible net oxygen flux, suggests there is a large net oxygen consumption. 


```{r tidy = F}
summary(lake_rate)
```

However, looking at the summary we can see the `$rate.2pt` as the second from last column. This is simply the oxygen differential between the two timepoints divided by the time difference, and shows us there was a slight net oxygen production. In this case, this rate is much more relevant towards the research question.

### Converting

This two-point rate will not be the one converted if we pass the `calc_rate` object to `convert_rate`, however we can  enter the two-point rate as a value to convert the rate to units.

```{r tidy = F, eval = F}
lake_rate_conv <- convert_rate(0.006235444,
                               o2.unit = "mg/l",
                               time.unit = "hrs",
                               output.unit = "mg/h",
                               volume = 1)
```

Alternatively, extract it directly.

```{r tidy = F, eval = T}
lake_rate_conv <- convert_rate(lake_rate$rate.2pt,
                               o2.unit = "mg/l",
                               time.unit = "hrs",
                               output.unit = "mg/h",
                               volume = 1)
# print
lake_rate_conv
```

Note here the `volume` input. This is not a respirometry experiment, so there is no "volume" as such. In this situation this should refer to the volume of the oxygen concentration unit (in this case litres), so the rate is dimensionally valid for *any* volume or mass of this lake. This will be the case for *any* units of oxygen, because internally they are all converted to mg/L before conversion to the output units. 


