---
title: "Two-point analyses"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Two-point analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, comment = "#>", cache = FALSE, tidy = TRUE, 
  highlight = TRUE, fig.width = 6, fig.height = 6)
```

In some analyses we may want to use only two datapoints of oxygen concentration separated by a known time period to calculate a rate. This could result from **closed-bottle respirometry**, where only the initial and final oxygen concentrations are recorded, in situations of periodic and irregular oxygen monitoring or sampling, or because experimental  problems caused erroneous or suspect data to be recorded during an experiment. This approach of course comes with the assumption that rates are relatively constant over the relevant time period, or alternatively the investigator is not particularly interested in how rate may have fluctuated in the intervening time, but rather the total flux and therefore total rate.

While the linear regression approach of `respR` may still provide valid rate estimates, this may not work in the case of irregular data recordings, and the two-point output of `calc_rate` allows useable rates to be determined from sparse or partially erroneous data via this approach.

### Two-point data
Simple two-point data can be entered into `calc.rate` in a 2x2 data frame created via typical R syntax:
```{r}
two_pt_data <- data.frame(time = c(0, 10), o2 = c(8.5, 6.3))
two_pt_data
```
```{r, fig.keep="none"}
two_pt_result <- calc_rate(two_pt_data)
summary(two_pt_result)

```

The two-point rate can be seen at the end of the summary object. With only two datapoints, it will obviously be equal to the rate in the linear regression method. However, it can be extracted directly:

```{r}
two_pt_result$rate_2pt
```

### Working around erroneous data
More useful is using two-point calculations to work around cases of partially erroneous data. Often we can still extract perfectly valid rates as long as the investigator knows *some* of the data contains the true oxygen values.  The following example uses a dataset where interference has caused regions of erroneous data, however the intitial and end conditions are known to be reliable recordings. 
```{r echo = F}
urchin_mod <- urchins.rd[1:150,c(1,15)]

urchin_mod[30:49,2] <- urchin_mod[30:49,2]-c(seq(0.02,0.46,0.046), rev(seq(0.02,0.46,0.046)))
urchin_mod[70:89,2] <- urchin_mod[70:89,2]-c(seq(0.02,0.36,0.036), rev(seq(0.02,0.36,0.036)))
```

```{r message=FALSE, warning=FALSE, results="hide"}
inspect_data(urchin_mod)
```

Here we can see there are two erroneous regions. We could remove these sections, or use the lower section of the data, but we can also use the subsetting operators of `calc_rate` to select data that starts and ends where we know valid O₂ recordings were made:
```{r message=FALSE, warning=FALSE}
two_pt_result <- calc_rate(urchin_mod, from = 2, to = 23, by = "time")
summary(two_pt_result)
```

We can clearly see the linear regression method substantially underestimates the rate because it is skewed downwards. In this case, the two-point output will clearly be a better estimate of the rate of this specimen. 

### Processing two-point rates
Two-point rates can be passed to `adjust_rate` for background corrections, and `convert_rate` for conversion to the desired output units. However, they must be specifically called using `$rate_2pt` or, by default, these functions will operate using the linear regression derived rate. 
```{r}
adjust_rate(two_pt_result$rate_2pt, by = -0.001)

convert_rate(two_pt_result$rate_2pt, o2.unit = "mg/L", time.unit = "h", output.unit = "mg/h/g", volume = 25, mass = 0.2)
```

They can alternatively be entered manually:
```{r}
convert_rate(-0.0224, o2.unit = "mg/L", time.unit = "h", output.unit = "mg/h/g", volume = 25, mass = 0.2)
```


