---
title: "v1.1: How to install and run older code"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{v1.1: How to install and run older code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = F}
library(knitr) # load knitr to enable options
library(respR) # load respR

opts_chunk$set(collapse = TRUE, 
               comment = "#>", 
               cache = FALSE, 
               tidy = TRUE, 
               highlight = TRUE, 
               fig.width = 10, 
               fig.height = 5,
               fig.align = "center",
               R.options = list(scipen = 999, 
                                digits = 3))
```



`respR` v2.0 contains several, unavoidable code breaking changes that means code written for use with v1.x may produce different outputs or fail to run. 

This page is intended to be a resource for those who have submitted v1.x code as part of a publication, or those who may want to reproduce v.1.x analyses in the future. 

See [here]() for the major code breaking changes in v2.0 and how earlier code may be amended to run in the new version. 

If you are in the process of submitting code as part of a publication and don't want to have to revise it to work with `respR` v2, or if it is too late to do so, please include a link to this page so others will find it easier to run the code in the future. For publications already published this page will remain on the website as a resource for anyone wanting to run the code or reproduce analyses. 

### Installing `respR` v1.1.1 

If you need to install the final v1 release of `respR` to run older code, the easiest way is to use this command:

```{r eval = F}
devtools::install_github("januarharianto/respR@v1.1.1")
```

This will install v1.1.1 of the package and its dependencies (see [below](#deps)) as normal. 

An alternative is to download the package source files and compile the package locally. These can be found [here](https://github.com/januarharianto/respR/releases/tag/v1.1.1) on Github, or [here](https://zenodo.org/record/6112437) on Zenodo. These contain all the files needed to compile `respR` v1.1.1 as a local package via an RStudio project or other R installation. 

As a final option, the functions can be found in the `R` folder within that archive, and these scripts can be run to create the functions in the local environment. This may require some editing of code to refer to these local functions. 

### `respR` v1.1.1 

v1.1.1 contains a function called `installation_help()`. This simply opens a help file which contains further installation instructions. This includes a list of dependencies. These will have been installed if the `devtools::install_github` command above was used, but they will be the most recent versions. We cannot guarantee these will not eventually make code changes that break v1.1.1 functionality. 

Below is a list of the latest versions tested and known to work with v1.1.1. If there are problems running v1.1.1 code that can be linked to a newer version of a particular package, it can be uninstalled and the older known working version installed by adapting this command: 

```
devtools::install_version("lubridate", version = "1.8.0", repos = "http://cran.us.r-project.org")`
```

You can also install older versions from downloaded package zip archives. Visit the [CRAN Archive]( https://cran.r-project.org/src/contrib/Archive/) to view and download the archived versions of CRAN hosted packages.

### Dependencies in v1.1.1 {#deps}

These are the latest versions tested and confirmed to work in `respR` v1.1.1:

 `broom        0.7.12` 
 
 `data.table   1.14.2` 
 
 `dplyr        1.0.8` 
 
 `ggplot2      3.3.5` 
 
 `graphics     4.1.2`
 
 `grDevices    4.1.2` 
 
 `gridExtra    2.3` 
 
 `lubridate    1.8.0` 
 
 `magrittr     2.0.2` 
 
 `marelac      2.1.10` 
 
 `methods      4.1.2` 
 
 `parallel     4.1.2` 
 
 `readxl       1.3.1` 
 
 `roll         1.1.6` 
 
 `segmented    1.4-0` 
 
 `stats        4.1.2` 
 
 `stringr      1.4.0` 
 
 `utils        4.1.2` 
 
 `xml2         1.3.3`
 

### Suggests in v1.1.1 

These are packages which may not strictly be necessary, but are suggested to be installed.

`knitr         1.37` 

`rmarkdown     2.11` 

`rMR           1.1.0` 

`DiagrammeR    1.0.8` 

`FishResp      1.1.0` 

`respirometry  1.3.0`

`testthat      3.1.2`


